### 2025-05-22 paper.tex
Initial draft

\documentclass[acmsmall, screen, nonacm]{acmart}

\usepackage{listings}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{stmaryrd}
\usepackage{fontawesome5}
\usepackage{mathpartir}
\usepackage{sourcecodepro}

\lstset{
  basicstyle=\ttfamily\small,
  columns=fullflexible,
  keepspaces=true
}

\lstdefinestyle{ocaml}{
  keywordstyle=\color{blue},
  emphstyle=\color{violet},
  language=[Objective]Caml,
  morekeywords={effect, exclave},
  escapeinside={(*@}{@*)},
  literate=
    {[]}{$\glob$}{1}
}

\lstdefinestyle{haskell}{
  keywordstyle=\color{blue},
  emphstyle=\color{violet},
  language=Haskell
}

\newcommand{\glob}{\mathop{\Box}}
\newcommand{\cat}{\mathbb{C}}
\newcommand{\catop}{\cat^{\mathtt{op}}}
\newcommand{\id}[1]{\mathrm{id}_{#1}}
\newcommand{\setc}{\mathrm{Set}}
\newcommand{\psh}{\widehat{\cat}}
\newcommand{\terminal}{\top}
\newcommand{\initial}{\bot}
\newcommand{\lawc}{\mathrm{Law}}
\newcommand{\lawcop}{\lawc^{\mathtt{op}}}
\newcommand{\pshlawcop}{\widehat{\lawcop}}
\newcommand{\freef}{Free}
\newcommand{\constfname}{\Omega}
\newcommand{\constf}[1]{\constfname #1}
\newcommand{\yonedaname}{y}
\newcommand{\yoneda}[1]{y(#1)}
\newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
\newcommand{\arrow}[2]{#1 \Rightarrow #2}
\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\reflectname}{\Uparrow}
\newcommand{\reflect}[2]{\reflectname_{#1}(#2)}
\newcommand{\reifyname}{\Downarrow}
\newcommand{\reify}[2]{\reifyname_{#1}(#2)}
\newcommand{\lock}{\text{\faLock}}
\newcommand{\mon}{\mathcal{T}}
\newcommand{\ind}[1]{W_{#1}}

\newcommand{\todo}[1]{{\color{red}TODO: #1}}

\title{Locality and Effect Reflection}
\author{Leo White}
\affiliation{\institution{Jane Street}\country{UK}}
\date{}
\setcopyright{none}

\begin{document}

\begin{abstract}
  Algebraic effects and handlers allow the programmer to extend their
  languages \emph{ambient monad} \textbf{--} the monad describing which
  effects ordinary functions can perform \textbf{--} with additional
  effectful operations.

  Locality captures the general idea of a value depending on some notion
  of ``location''. We can represent it in a type system using a
  comonadic \emph{global modality} $\glob$ that tracks which values do
  not depend on their location. A particularly useful notion of
  ``location'' is the ambient monad.

  We introduce an alternative presentation of algebraic effects and
  handlers, \emph{effect reflection}, that provides, for any algebraic
  theory $\Sigma$, an isomorphism:
  \begin{equation*}
    \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
  \end{equation*}
  where $\ind{\Sigma}$ is an inductive type representing the free
  algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
  to inject terms of $\Sigma$ into the current ambient
  monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
  $\Sigma$. This presentation falls out naturally from a simple
  denotational semantics for algebraic effects in terms of presheaves
  over bounded Lawvere theories.

  Effect reflection lets you work with any algebraic monad using an
  ordinary function type, allowing monadic code to be written in direct
  style. This provides all the benefits of typed algebraic effects and
  handlers without requiring an effect system. We demonstrate effect
  reflection by implementing it as a library in
  OxCaml\cite{lorenzen2024oxidizing,} and working through some classic
  examples of programming with effects.
\end{abstract}

\maketitle

\section{Reflection for algebraic effects}

\paragraph{Algebraic effects} Algebraic effects and handlers are an
increasingly popular approach to programming with effects. Handlers
allow the programmer to add additional effects to the \emph{ambient
  monad} of their language -- the monad describing which effects an
ordinary function can perform -- for some region of their program.

For example, in OCaml, one can write:
\begin{lstlisting}[style=ocaml]
type 'a Effect.t +=
  | Get : int Effect.t
  | Set : int -> unit Effect.t

let handle (f : unit -> 'a) : 'a =
  match f () with
  | x -> fun _ -> x
  | effect Get, k -> fun s -> continue k s s
  | effect Set s', k -> fun _ -> continue k () s'
\end{lstlisting}
and for the duration of the \lstinline[style=ocaml]{f ()} function call
the ambient monad is extended with two additional effectful operations:
\begin{lstlisting}[style=ocaml]
let get () : int = perform Get
let set (i : int) : unit = perform (Set i)
\end{lstlisting}

Note that, in OCaml, nothing prevents attempting to perform an algebraic
effect operation that is not supported by the current ambient monad,
resulting in an \lstinline[style=ocaml]{Unhandled} exception being
raised. Other languages, e.g. Koka\cite{??}, use an effect system to
statically ensure operations are only performed when they are available.

\paragraph{Locality and effect reflection} Locality captures the general idea of a value
depending on some notion of ``location''. We can represent it in a type
system using a comonadic \emph{global modality} $\glob$ that tracks
which values do not depend on their location. A particularly useful
notion of ``location'' is the ambient monad.

Treating the ambient monad as a notion of location allows a language to
provide an isomorphism:
\begin{equation*}
  \mathop{M}(\glob X) \qquad \cong \qquad Y \mathbin{\rightarrow} \glob X
\end{equation*}
where $M$ is any algebraic monad, $Y$ is a specific type related to that
monad, and $\glob$ is a comonadic global modality. We call this
isomorphism \emph{effect reflection}.

The ability to use the ordinary arrow type
$Y \mathbin{\rightarrow} \glob X$ in place of the monad
$\mathop{M}(\glob X)$ allows us to conveniently write effectful
code in direct style, just as algebraic effects do.

\paragraph{Effect reflection in OxCaml}
OxCaml\cite{lorenzen2024oxidizing} provides type system support for
tracking locality and runtime support for algebraic effects. For
backwards compatibility with OCaml, an unadorned arrow type
\lstinline[style=ocaml]{t -> s} is treated as if the parameter and
return types were actually \lstinline[style=ocaml]{[]t} and
\lstinline[style=ocaml]{[]s}. To get parameters or results that are not
implicitly global, we write
e.g. \lstinline[style=ocaml]{t @ local -> s @ local}.

Using this support, with can provide effect reflection as a simple API
that we can use for programming with effects.  For example, we can use
it to write an effect handler equivalent to the OCaml one above:
\begin{lstlisting}[style=ocaml]
module State = struct
  type 'a t =
    | Get : int t
    | Set : int -> unit t
end

module State_eff = Reflection(State)

let handle (f : Handler(State).t @ local -> 'a) : 'a =
  let rec loop s : 'a Term(State).t -> 'a  = function
    | Return x -> x
    | Op(Get, k) -> loop s (k s)
    | Op(Set s', k) -> loop s' (k ())
  in
  loop (State_eff.reify f)
\end{lstlisting}
which calls \lstinline[style=ocaml]{f} and passes it a local value of
type \lstinline[style=ocaml]{Handler(State).t}. For the duration of the
call to \lstinline[style=ocaml]{f} the ambient monad is extended with
two additional effectful operations, which can be used via reflection:
\begin{lstlisting}[style=ocaml]
let get (h : Handler(State).t @ local) : int =
  State_eff.reflect (Op(Get, fun x -> x)) h

let set (h : Handler(State).t @ local) (i : int) : unit =
  State_eff.reflect (Op(Set i, fun x -> x)) h
\end{lstlisting}

Note that the locality of the handler value ensures that these
operations can only be performed when they are available in the ambient
monad. This provides the same guarantees as using an effect system.

Effect reflection is closely related to \emph{lexical effect handlers} in
systems such as Effeckt\cite{??}. Lexical effect handlers have a value that
represents the effect handler, but these values are second class and
require their own special typing rules.

Effect reflection is also closely related to \emph{named effect
  handlers}\cite{??}. Named effect handlers use higher-rank polymorphism
to emulate locality tracking in much the same way as the
\lstinline[style=haskell]{St} monad in Haskell. Using higher-rank
polymorphism prevents global type inference, requiring complex inference
that extends beyond classic Hindly-Milner, and demanding a higher
annotation burden on the programmer. \todo{Mention the explicit monad
that named effect handlers presumably need.}

Effect reflection is a variation on monadic reflection\cite{??}.
\todo{Make this relationship precise.}

\paragraph{Denotational semantics} The design of effect reflection falls
out naturally from a simple denotational semantics of algebraic
effects. The underlying idea is to index the semantics by the ambient
monad. Rather than the semantics of a type being a set it is instead a
function from ambient monads to set: if you tell it what the ambient
monad is it can tell you what its set of values is.

The semantics is defined in terms of presheaves of the opposite of the
category of \emph{bounded Lawvere theories}. Bounded Lawvere theories
are a categorical representation of algebraic theories: we represent the
ambient algebraic monad by the corresponding algebraic theory. The key
constructs of effect reflection are then defined as follows:

\begin{itemize}
\item The global modality $\glob$ on such presheaves is defined such
  that $\glob(P)(L) = P(\initial)$. In other words, for a given presheaf
  $P$ it gives the constant presheaf that always ignores the ambient
  monad and instead returns $P$ at $\initial$, where $\initial$ is the
  algebraic theory with no operations or equations. This basically
  restricts a type to its terms that do not depend on the ambient monad.

\item The ambient monad can be defined as a monad $\mon$ on such
  presheaves, which maps each algebraic theory to its free algebra. We use
  $\mon$ to define the semantics of computations.

\item The semantics of our handler types $Y$ are given by the Yoneda
  embedding, and our effect reflection isomorphism becomes a simple
  consequence of the Yoneda lemma.
\end{itemize}

These presheaf semantics are closely related to presheaf semantics for
higher-order abstract syntax\cite{??}.

\paragraph{Structure of the paper} We proceed as follows:
\begin{itemize}
\item In Section \ref{sec:pracice}, we give an overview of effect reflection in
  practice in OxCaml
\item In Section \ref{sec:semantics}, we describe the constructions for a
  novel denotational semantics of algebraic effects
\item In Section \ref{sec:language}, we define a simply typed calculus featuring
  effect reflection along with its denotational semantics
\item In Section \ref{sec:implementation}, we give an implementation of effect
  reflection in OxCaml in terms of the untracked algebraic effects it
  inherits from OCaml.
\item In Section \ref{sec:related-work}, we discuss related work, and in
  Section \ref{sec:future-work}, we discuss future work.
\end{itemize}

\section{Effect reflection in practice}
\label{sec:pracice}

We can provide effect reflection as a simple library API in OxCaml. In
this section we describe this API and show how to use it to write a
number of typical examples of algebraic effects.

\subsection{Locality in OxCaml}

The effect reflection API relies on OxCaml's support for tracking
locality in types, so let's start with an overview of that support.

Values in OxCaml are associated with both a type -- which describe how
values can be introduced and eliminated -- and a \emph{mode} -- which
track properties related to operations other than introduction and
elimination. Modes are a product of a number of different axes, tracking
a variety of different properties, but here we are only interested in
one of them: \emph{locality}. Ignoring the other axes, a mode is either
\lstinline[style=ocaml]{local} or \lstinline[style=ocaml]{global}.

Locality tracks the ability of a value to leave \emph{regions}.  Regions
describe some period of time during a program's execution. Every
function call defines a corresponding region that, by default, lasts for
the entire extent of the call. This means that, by default,
\lstinline[style=ocaml]{local} values cannot be returned from a
function.
\begin{lstlisting}[style=ocaml]
let leak_local x =
  let (o @ local) = Some x in
  o
(*@ \textcolor{red}{Error: This value is local and would escape its region.} @*)
\end{lstlisting}

To support functions that return local values, the region of a function
can be ended early using the \lstinline[style=ocaml]{exclave} construct.
\begin{lstlisting}[style=ocaml]
let return_local x =
  exclave
    (let (o @ local) = Some x in
     o)
\end{lstlisting}
\lstinline[style=ocaml]{exclave e} can only appear in tail position
within a function.  It ends the function's region early and then
executes the expression
\lstinline[style=ocaml]{e}. \lstinline[style=ocaml]{e} is essentially
executed within the region of the caller of the function and any local
values created within it can be safely returned to that caller.

The place where modes appear within the type algebra is on arrow
types. A function being \lstinline[style=ocaml]{global} doesn't tell you
that the values it accepts or returns should be
\lstinline[style=ocaml]{global} too, so instead the arrow type is
labelled with two modes: one for the argument and one for the return
value.
\begin{lstlisting}[style=ocaml]
val return_local : 'a @ local -> 'a option @ local
\end{lstlisting}
For convenience, and backwards compatibility with OCaml, either mode can
be omitted, in which case it defaults to
\lstinline[style=ocaml]{global}.

By default, modes are deep: a local \lstinline[style=ocaml]{string list}
is a local list of local strings. However, the depth that a mode applies
can be controlled via \emph{modalities}. Record fields can be annotated
with \lstinline[style=ocaml]{global} modality to indicate that the
contents of the field are global even when the surrounding record is
local. For example:
\begin{lstlisting}[style=ocaml]
type 'a gbl = { g : 'a @@ global }
\end{lstlisting}
gives a record whose only field has the \lstinline[style=ocaml]{global}
modality. The contents of the \lstinline[style=ocaml]{g} field is always
global, for example:
\begin{lstlisting}[style=ocaml]
let return_global x =
  let (r @ local) = { g = x } in
  r.g
\end{lstlisting}
is allowed and gives a function of type:
\begin{lstlisting}[style=ocaml]
val return_global : 'a @@ global -> 'a @@ global
\end{lstlisting}
Variant constructor arguments can also be annotated with modalities.

Note that, whilst there is a \lstinline[style=ocaml]{global} modality,
there is no \lstinline[style=ocaml]{local} modality. It does not make
sense for a value which has the ability to leave a region to contain a
value which doesn't, as when the first value leaves a region is will
implicit cause the second one to leave that region as well. This means
the \lstinline[style=ocaml]{global} mode is always deep.

\subsection{Effect reflection interface}
\label{sec:interface}

\begin{lstlisting}[style=ocaml]
module type Op = sig type 'a t end

module Term (O : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
end

module Handler(O : Op) : sig
  type 'a t
end

module Reflection (O : Op) : sig

  val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t

  val reflect : 'a Term(Op).t -> (Handler(O).t @ local -> 'a)

  val reify_local : (Handler(O).t @ local -> 'a) @ local -> 'a Term(O).t @ local

  val reflect_local : 'a Term(O).t @ local -> (Handler(O).t @ local -> 'a) @ local

end
\end{lstlisting}

\begin{lstlisting}[style=ocaml]
let perform (op : 'a O.t) : 'a =
  reflect (Op(op, fun x -> x))
\end{lstlisting}

\subsection{One-shot continuations}

For various practical reasons, OCaml's algebraic effects are restricted
to only allow continuations to be resumed a single time. This rules out
interesting effects like backtracking search, but improves the ability
to reason about the interaction of effects and resources.

There is no fundamental reason that effect reflection should be
restricted to only one-shot continuations, so for the purposes of this
paper we shall ignore the restriction. However, for the same reasons as
OCaml, our actual implementation is restricted one-shot
continuations. Note that OxCaml's \lstinline[style=ocaml]{once}
mode\cite{lorenzen2024oxidizing} allows us to enforce this restriction
statically.

\subsection{Effect sums}

\begin{lstlisting}[style=ocaml]
module Sum (L : Op) (R : Op) : sig
  type 'a t =
    | Left : 'a L.t -> 'a t
    | Right : 'a R.t -> 'a t
end

module Project(L : Op) (R : Op) : sig
  val outl : Handler(Sum(L)(R)).t -> Handler(L).t

  val outr : Handler(Sum(L)(R)).t -> Handler(R).t
end
\end{lstlisting}

As a Lawvere theory, \lstinline[style=ocaml]{Sum(L)(R)} is the sum of
\lstinline[style=ocaml]{L} and \lstinline[style=ocaml]{R}. This means
that the terms of its free algebra $T_{L + R}(\tau)$ are equivalent to
values of the initial algebra $\mu X. T_L(\tau + F_L(X)$. We can expose
this fact via the following functions:
\begin{lstlisting}
module Half_term (L : Op) (R : Op) : sig
  type 'a t =
    | Return : 'a @@ global -> 'a t
    | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t

  let rec reflect_left : 'a Term(Sum(L)(R)).t -> (Handler(L).t @ local -> 'a Half_term(L)(R).t)
   = function
    | Return v -> fun _ -> Return v
    | Op(Left op, k) -> fun l -> L.reflect (Op(op, (fun res -> reflect_left (k res) l)))
    | Op(Right op, k) -> fun _ -> Op(op, (fun res -> reflect_left (k res)))

  let rec reify_left : (Handler(L).t @ local -> 'a Half_term(L)(R).t) -> 'a Term(Sum(L)(R)).t =
    fun h ->
      let rec loop = function
        | Return (Return x) -> Return x
        | Return (Op(r, k)) -> Op(Right r, fun res -> reify_left (k res))
        | Op(l, k) -> Op(Left l, fun res -> loop (k res))
      in
      loop (L.reify h)

end
\end{lstlisting}

\subsection{Examples}

\subsubsection{State}

\subsubsection{Concurrency}

\subsubsection{Generators and concurrency}

\section{Presheaves for algebraic effects}
\label{sec:semantics}

Our language falls out naturally from its categorical semantics, so we
focus on that. We intend to index the meaning of types by an ambient
algebraic monad of effects: rather than have the semantics of each type
be a set of elements, the semantics of each type is a mapping from the
ambient monad to a set of elements.

For a category $\cat$, $\psh = \setc^{\catop}$ is the category of
presheaves over $\cat$. If $\cat$ has a terminal object $\terminal$, we
can define the global modality $\glob \mathrel{:} [\psh, \psh]$ such that:
\begin{equation*}
  \glob(P)(L) = P(\terminal)
\end{equation*}

We use the ambient monad as our notion of location, so we take $\cat$ to
be $\lawcop$, where $\lawc$ is the category of countable Lawvere
theories. Lawvere theories are a categorical representation of algebraic
theories. Morphisms in $\lawc$ are maps from the operations of one
theory to the operations of another that preserve the equations of the
first theory according to the equations of the second.  The initial
theory is the algebraic theory with no operations. The coproduct of two
theories is the theory made by combining the operations and equations of
both theories.

Given a Lawvere theory $L$, the underlying set of its free algebra forms
a monad $T_L$ on $\setc$. $T_{\_}$ is functorial and we can use it to
construct a monad $\mon$ on $\pshlawcop$ in the obvious way, such that:
\begin{equation*}
\mon(P)(L) = T_L(P(L))
\end{equation*}
We will use $\mon$ as our computation monad.

$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding. For each
Lawvere theory $L$ there is a presheaf $\yoneda{L}$ in $\pshlawcop$ such
that:
\begin{equation*}
\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}}
\end{equation*}
We will use $\yoneda{L}$ to represent maps from $L$ into the ambient
monad. Note that $\yonedaname$ preserves finite products:
$\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}$
where $+$ is the coproduct in $\lawc$, which is the product in
$\lawcop$.

Categories of presheaves are cartesian closed, and the exponential
$\arrow{P}{Q}$ in $\pshlawcop$ maps $L$ to
$\homset{\pshlawcop}{\yoneda{L} \times P}{Q}$. So, via the Yoneda lemma,
we get the following:
\begin{align*}
(\arrow{\yoneda{L}}{\mon(\glob A)})(L') \quad \cong \quad &T_{(L' + L)}(A(\terminal))
\end{align*}
which will form the basis of our language's isomorphisms between arrows
$Y \rightarrow \glob A$ and algebraic monads $M(\glob A)$.

\section{A language with effect reflection}
\label{sec:language}

We build a simple language to (mostly) achieve our aim. Rather than
working with arbitrary algebraic monads as promised, we follow the
finest traditions of the algebraic effects literature and restrict our
attention to algebraic theories without equations. Our language is
simply typed, featuring all the usual type-formers. It uses a variation
of fine-grained call-by-value: it has separate syntax and typing
judgements for values and computations.

We define algebraic signatures $\Sigma$ as finite sets of operations
\[
\{ op_1 \mathop{:} \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
\mathop{:} \tau_{p_n} \rightarrowtail \tau_{a_n} \}
\]
An operation type
$\tau_p \rightarrowtail \tau_a$ indicates an operation parameterised by
$\tau_p$ and with arity $\tau_a$. Such signatures correspond to Lawvere
theories without equations. We restrict parameters and arities to base
types to ensure that our theories only have a countable number of
operations of countable arities. We write $\sem{\Sigma}$ for the
corresponding Lawvere theory, and $F_\Sigma$ for the endofunctor on
$\setc$ for which $T_{\sem{\Sigma}}$ is the free monad.

The language provides inductive types $\ind{\Sigma}$ to directly
represent terms of an algebraic signature $\Sigma$. $\ind{\Sigma}(\tau)$
has a constructor $op$ with arguments of type $\tau_p$ and
$\tau_a \rightarrow \ind{\Sigma}(\tau)$ for each operation
$op \mathop{:} \tau_p \rightarrowtail \tau_a$ in $\Sigma$, as well as a
constructor $ret$ with an argument of type $\tau$.

The language provides types $\yoneda{\Sigma}$ to represent morphisms
from $\sem{\Sigma}$ into the ambient monad. Using these, our desired
isomorphism appears in the language as two operations, reflection
$\reflect{\Sigma}{c}$ and reification $\reify{\Sigma}{v}$.
\begin{mathpar}
\inferrule{\Gamma \mathrel{\vdash_C} c : \ind{\Sigma}(\glob \tau)}
{\Gamma \mathrel{\vdash_V} \; \reflect{\Sigma}{c} : \yoneda{\Sigma} \rightarrow \glob \tau}
\and
\inferrule{\Gamma \mathrel{\vdash_V} v : \yoneda{\Sigma} \rightarrow \glob \tau}
{\Gamma \mathrel{\vdash_C} \; \reify{\Sigma}{v} : \ind{\Sigma}(\glob \tau)}
\end{mathpar}

Our denotational semantics in terms of $\pshlawcop$ is
straightforward. For types, the translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
\begin{mathpar}
\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}} \and
\sem{\glob \tau} = \glob \sem{\tau} \and
\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \and
\sem{\ind{\Sigma}(\tau)} = \mu X.(\sem{\tau} + F_\Sigma(\mon(X)))
\end{mathpar}
where $\mu X. F(X).$ represents the initial algebra of a suitable
endofunctor. For any Lawvere theory $L$,
$\mu X.T_L(\sem{\tau} + F_\Sigma(X))$ is isomorphic to to
$T_{(L + \sem{\Sigma})}$ \cite{hyland2006combining}. That means that
$\mon(\sem{\ind{\Sigma}(\glob \tau)})$ is isomorphic to the presheaf that
maps $L$ to $T_{(L + \sem{\Sigma})}(\sem{\tau}(\terminal))$, which in
turn is isomorphic to $\sem{\yoneda{\Sigma} \rightarrow \glob \tau}$ by the
result in the previous section. This isomorphism directly provides the
semantics of reflection and reification.

\begin{align*}
&\sem{\Gamma \vdash_V v \mathop{:} \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
&\sem{\glob \Gamma \vdash_V \mathop{\mathrm{box}} c \mathop{:} \glob \tau} = \glob \sem{\Gamma \vdash_C c \mathop{:} \tau}
\end{align*}
\begin{align*}
&\sem{\Gamma \vdash_C c \mathop{:} \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
&\sem{\Gamma \vdash_C \mathop{\mathrm{return}} v \mathop{:} \tau} = \eta_{\sem{\tau}} \circ \sem{\Gamma \vdash_V v \mathop{:} \tau} \\
&\sem{\Gamma \vdash_C \mathop{\mathrm{let}} x : \tau_1 = c_1 \mathop{\mathrm{in}} c_2\mathop{:} \tau_2} = \mu_{\sem{\tau_2}} \circ \mon(\sem{\Gamma; x \mathop{:} \tau_1 \vdash_C C_2 \mathop{:} \tau_2}) \circ t_{\sem{\Gamma}, \sem{\tau_1}} \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \vdash_C C_1 \mathop{:} \tau} \rangle \\
&\sem{\Gamma \vdash_C \mathop{\mathrm{unbox}} c \mathop{:} \tau} = \mon(\epsilon_{\sem{\tau}}) \circ \sem{\Gamma \vdash_C c \mathop{:} \glob \tau} \\
&\sem{\Gamma \vdash_C \reflect{\Sigma}{c} \mathop{:} \yoneda{\Sigma} \rightarrow \glob \tau} = \reflect{\Sigma\tau}{\sem{\Gamma \vdash_C c \mathop{:} \ind{\Sigma}(\glob \tau)}} \\
&\sem{\Gamma \vdash_C \reify{\Sigma}{c} \mathop{:} \ind{\Sigma}(\glob \tau)} = \reify{\Sigma\tau}{\sem{\Gamma \vdash_C c \mathop{:} \yoneda{\Sigma} \rightarrow \glob \tau}} \\
\end{align*}
where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
respectively of $\mon$; $\epsilon$ is the counit of $\glob$;
$\reflect{\Sigma}{}$ and $\reify{\Sigma}{}$ are the natural
transformation in the isomorphism from the previous section.

\section{Implementation}
\label{sec:implementation}

\subsection{Implementation with untyped effect handlers}

We can implement the interface from Section \ref{sec:interface} directly
in terms of OCaml's untyped effect
handlers\cite{Sivaramakrishnan2021retrofitting}. The trick is to
generate a fresh effect constructor for each reification, pass that
constructor around as part of the \lstinline[style=ocaml]{handler}, and
then use it to perform operations during reflection.

\begin{lstlisting}[style=ocaml]
module Reflection (O : Op) = struct

  type 'a term =
    | Return : 'a @@ global -> 'a term
    | Op : 'r O.t @@ global * ('r -> 'a term) -> 'a term

  type handler = { perform : 'r. 'r O.t -> 'r }

  let reify f =
    let module Handler =
      struct type 'a Effect.t += H : 'a O.t -> 'a Effect.t end
    in
    let handler = { perform = fun o -> Effect.perform (Handler.H o) } in
    match f handler with
    | v -> Return v
    | effect Handler.H o, k -> Op(o, fun x -> Effect.continue k x)

  let reflect t =
    fun {perform} ->
      let rec loop = function
        | Return v -> v
        | Op(o, f) -> loop (f (perform o))
      in
      loop t

  let reify_local f = exclave
    let module Handler =
      struct type 'a Effect.t += H : 'a O.t -> 'a Effect.t end
    in
    let handler = { perform = fun o -> Effect.perform (Handler.H o) } in
    match f handler with
    | v -> Return v
    | effect Handler.H o, k -> Op(o, fun x -> Effect.continue k x)

  let reflect_local t = exclave
    fun {perform} ->
      let rec loop = function
        | Return v -> v
        | Op(o, f) -> loop (f (perform o))
      in
      loop t

end
\end{lstlisting}
Sums are then implemented by just lifting the injections on operations:
\begin{lstlisting}[style=ocaml]
module Sum (L : Op) (R : Op) = struct

  module Op = struct
    type 'a t =
      | Left : 'a L.t -> 'a t
      | Right : 'a R.t -> 'a t
  end

  let outl {perform} =
    {perform = fun op -> perform (Left op)}

  let outr {perform} =
    {perform = fun op -> perform (Right op)}

end
\end{lstlisting}

\subsection{More efficient implementation in terms of fibers}

In the above implementation, whenever an operation is reflected we
perform a linear search up the stack of handlers until we find the
corresponding one. This search is not actually necessary...

\section{Related work}
\label{sec:related-work}

\section{Future work}
\label{sec:future-work}

\bibliography{references}
\bibliographystyle{plain}

\end{document}

### 2025-07-01 paper.tex
Write more

@@
 \usepackage{sourcecodepro}
+\usepackage{amsthm}
+
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
 
@@
 
+\newcommand{\nat}{\mathbb{N}}
 \newcommand{\glob}{\mathop{\Box}}
 \newcommand{\cat}{\mathbb{C}}
-\newcommand{\catop}{\cat^{\mathtt{op}}}
+\newcommand{\op}{\mathtt{op}}
+\newcommand{\catop}{\cat^{\op}}
 \newcommand{\id}[1]{\mathrm{id}_{#1}}
@@
 \newcommand{\reflectname}{\Uparrow}
-\newcommand{\reflect}[2]{\reflectname_{#1}(#2)}
+\newcommand{\reflectraw}[1]{\mathop{\reflectname_{#1}}}
+\newcommand{\reflect}[3]{\reflectraw{#1}(#2, #3)}
+\newcommand{\reflectc}[2]{\mathop{\reflectname_{#1 #2}}}
+\newcommand{\performraw}[2]{\mathop{\reflectname(#1(#2))}}
+\newcommand{\perform}[5]{\performraw{#1}{#2}(#3, #4. #5)}
 \newcommand{\reifyname}{\Downarrow}
-\newcommand{\reify}[2]{\reifyname_{#1}(#2)}
-\newcommand{\lock}{\text{\faLock}}
+\newcommand{\reifyraw}[1]{\mathop{\reifyname_{#1}}}
+\newcommand{\reify}[3]{\reifyraw{#1}(#2.\,#3)}
+\newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
+\newcommand{\outl}[1]{\mathop{\mathrm{out}_L} #1}
+\newcommand{\outr}[1]{\mathop{\mathrm{out}_R} #1}
 \newcommand{\mon}{\mathcal{T}}
 \newcommand{\ind}[1]{W_{#1}}
+\newcommand{\indintro}[4]{#2_{#1}(#3, #4)}
+\newcommand{\retname}{\mathrm{ret}}
+\newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
+\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \{#3\}}
+\newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
+\newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
+\newcommand{\seq}{\,;\,}
+\newcommand{\fst}{\mathop{\pi_1}}
+\newcommand{\snd}{\mathop{\pi_2}}
+\newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
+\newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
+\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
+\newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
+\newcommand{\app}[2]{#1\,#2}
+\newcommand{\types}{\mathrel{:}}
+\newcommand{\gtypes}{\mathrel{:_{\Box}}}
+\newcommand{\cempty}{\varepsilon}
+\newcommand{\ccons}[2]{#1;\,#2}
+\newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
+\newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
+\newcommand{\cdiv}[1]{#1/\Box}
+\newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
+\newcommand{\letv}[3]{\mathop{\mathrm{let}} #1 = #2 \mathop{\mathrm{in}} #3}
+\newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
+\newcommand{\gunbox}[4]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
+\newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
+\newcommand{\sbst}[3]{#1[#2\backslash#3]}
+\newcommand{\turnv}{\mathrel{\vdash_V}}
+\newcommand{\turnc}{\mathrel{\vdash_C}}
+\newcommand{\evmap}{\mathop{ev}}
+\newcommand{\currymap}{\mathop{\lambda}}
+\newcommand{\reducesto}{\rightsquigarrow}
 
@@
   Algebraic effects and handlers allow the programmer to extend their
-  languages \emph{ambient monad} \textbf{--} the monad describing which
+  language's \emph{ambient monad} \textbf{--} the monad describing which
   effects ordinary functions can perform \textbf{--} with additional
@@
   handlers, \emph{effect reflection}, that provides, for any algebraic
-  theory $\Sigma$, an isomorphism:
+  signature $\Sigma$, an isomorphism:
   \begin{equation*}
@@
   Effect reflection lets you work with any algebraic monad using an
-  ordinary function type, allowing monadic code to be written in direct
-  style. This provides all the benefits of typed algebraic effects and
-  handlers without requiring an effect system. We demonstrate effect
-  reflection by implementing it as a library in
+  ordinary function type, allowing effectful code to be written in
+  direct style. This provides all the benefits of typed algebraic
+  effects and handlers without requiring an effect system. We
+  demonstrate effect reflection by implementing it as a library in
   OxCaml\cite{lorenzen2024oxidizing,} and working through some classic
@@
 
-let handle (f : unit -> 'a) : 'a =
+let handle_state (f : unit -> 'a) : 'a =
   match f () with
@@
 
-\paragraph{Locality and effect reflection} Locality captures the general idea of a value
-depending on some notion of ``location''. We can represent it in a type
-system using a comonadic \emph{global modality} $\glob$ that tracks
-which values do not depend on their location. A particularly useful
-notion of ``location'' is the ambient monad.
+\paragraph{Locality and effect reflection}
+Locality captures the general idea of a value depending on some notion
+of ``location''. We can represent it in a type system using a comonadic
+\emph{global modality} $\glob$ that tracks which values do not depend on
+their location. A particularly useful notion of ``location'' is the
+ambient monad.
 
@@
 isomorphism \emph{effect reflection}.
+
+\todo{Restrict the above claim to signatures}
 
@@
 
-let handle (f : Handler(State).t @ local -> 'a) : 'a =
-  let rec loop s : 'a Term(State).t -> 'a  = function
+let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
+  let rec loop s : 'a Term(State).t @ local -> 'a  = function
     | Return x -> x
@@
   in
-  loop (State_eff.reify f)
+  loop i (State_eff.reify_local f)
 \end{lstlisting}
@@
 let get (h : Handler(State).t @ local) : int =
-  State_eff.reflect (Op(Get, fun x -> x)) h
+  State_eff.perform Get h
 
 let set (h : Handler(State).t @ local) (i : int) : unit =
-  State_eff.reflect (Op(Set i, fun x -> x)) h
+  State_eff.perform (Set i) h
 \end{lstlisting}
@@
 \begin{itemize}
+\item The ambient monad can be defined as a monad $\mon$ on such
+  presheaves, which maps each algebraic theory to its free algebra. We use
+  $\mon$ to define the semantics of computations.
+
 \item The global modality $\glob$ on such presheaves is defined such
@@
   monad and instead returns $P$ at $\initial$, where $\initial$ is the
-  algebraic theory with no operations or equations. This basically
-  restricts a type to its terms that do not depend on the ambient monad.
-
-\item The ambient monad can be defined as a monad $\mon$ on such
-  presheaves, which maps each algebraic theory to its free algebra. We use
-  $\mon$ to define the semantics of computations.
+  algebraic theory with no operations or equations. This allows us to
+  restrict a type to its terms that do not depend on the ambient monad.
 
@@
   o
-(*@ \textcolor{red}{Error: This value is local and would escape its region.} @*)
+(*@ \textcolor{red}{Error: o} @*)
+       (*@ \textcolor{red}{\textasciicircum} @*)
+(*@ \textcolor{red}{This value is local and would escape its region.} @*)
 \end{lstlisting}
@@
 \begin{lstlisting}[style=ocaml]
-let return_global x =
-  let (r @ local) = { g = x } in
+let project_global (r @ local) =
   r.g
 \end{lstlisting}
-is allowed and gives a function of type:
-\begin{lstlisting}[style=ocaml]
-val return_global : 'a @@ global -> 'a @@ global
+gives a function of type:
+\begin{lstlisting}[style=ocaml]
+val project_global : 'a gbl @ local -> 'a @ global
 \end{lstlisting}
@@
 sense for a value which has the ability to leave a region to contain a
-value which doesn't, as when the first value leaves a region is will
-implicit cause the second one to leave that region as well. This means
+value which doesn't, as when the first value leaves a region it will
+implicitly cause the second one to leave that region as well. This means
 the \lstinline[style=ocaml]{global} mode is always deep.
@@
 
+The types and operations of our effect reflection API are parameterized
+by type constructors representing the operations of an algebraic
+effect. In OxCaml, such higher-kinded interfaces must be built using
+functors, so we start with a simple module type
+\lstinline[style=ocaml]{Op} to act as the parameter type for these
+functors:
 \begin{lstlisting}[style=ocaml]
 module type Op = sig type 'a t end
-
+\end{lstlisting}
+The fundamental types of the API are the \lstinline[style=ocaml]{Term}
+and \lstinline[style=ocaml]{Handler} types.
+\begin{lstlisting}[style=ocaml]
 module Term (O : Op) : sig
@@
 module Handler(O : Op) : sig
-  type 'a t
-end
-
+  type t
+end
+\end{lstlisting}
+For an effect signature \lstinline[style=ocaml]{O},
+\lstinline[style=ocaml]{'a Term(O).t} is the type of terms in the free
+algebra of the corresponding effect -- otherwise known as the freer
+monad\cite{??}. \lstinline[style=ocaml]{Handler(O).t} is the type of
+mappings from the operations of \lstinline[style=ocaml]{O} into the
+operations of the current ambient monad. Alternatively,
+\lstinline[style=ocaml]{Handler(O).t} can be described as representing
+algebraic effect handlers that are handling the current computation --
+hence the name.
+
+The main piece of the API is a family of isomorphisms:
+\begin{equation*}
+  \text{\lstinline[style=ocaml]{'a Term(O).t}} \qquad
+  \cong \qquad \text{\lstinline[style=ocaml]{Handler(O).t @ local -> 'a}}
+\end{equation*}
+Due to the absence of mode polymorphism in OxCaml, we expose this
+isomorphism as three functions: \lstinline[style=ocaml]{reify} and
+\lstinline[style=ocaml]{reify_local} which implement the left-to-right
+direction at the \lstinline[style=ocaml]{global} and
+\lstinline[style=ocaml]{local} modes respectively; and
+\lstinline[style=ocaml]{perform}, which is equivalent to the
+right-to-left direction but makes for a more convenient primitive in
+practice. \lstinline[style=ocaml]{perform} amounts to reflection of
+individual operations as opposed to full terms.
+\begin{lstlisting}[style=ocaml]
 module Reflection (O : Op) : sig
-
   val reify : (Handler(O).t @ local -> 'a) -> 'a Term(O).t
 
-  val reflect : 'a Term(Op).t -> (Handler(O).t @ local -> 'a)
-
-  val reify_local : (Handler(O).t @ local -> 'a) @ local -> 'a Term(O).t @ local
-
-  val reflect_local : 'a Term(O).t @ local -> (Handler(O).t @ local -> 'a) @ local
-
-end
-\end{lstlisting}
-
-\begin{lstlisting}[style=ocaml]
-let perform (op : 'a O.t) : 'a =
-  reflect (Op(op, fun x -> x))
+  val reify_local :
+    (Handler(O).t @ local -> 'a) @ local -> 'a Term(O).t @ local
+
+  val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
+end
+\end{lstlisting}
+Using \lstinline[style=ocaml]{perform} we can implement reflection of
+terms as:
+\begin{lstlisting}[style=ocaml]
+let rec reflect t h =
+  match t with
+  | Return x -> x
+  | Op(op, k) -> reflect (k (perform op h)) h
+
+let rec reflect_local (t @ local) h =
+  match t with
+  | Return x -> x
+  | Op(op, k) -> reflect_local (k (perform op h)) h
+\end{lstlisting}
+with types:
+\begin{lstlisting}[style=ocaml]
+val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
+
+val reflect_local :
+   'a Term(O).t @ local -> (Handler(O).t @ local -> 'a) @ local
+\end{lstlisting}
+
+\subsection{Example: State}
+
+As shown in the introduction we can implement an effect handler for
+state using this API. We start by defining the operations of the effect:
+\begin{lstlisting}[style=ocaml]
+module State = struct
+  type 'a t =
+    | Get : int t
+    | Set : int -> unit t
+end
+\end{lstlisting}
+Then we instantiate the reflection isomorphism for that effect and use
+it to write a handler:
+\begin{lstlisting}[style=ocaml]
+module State_eff = Reflection(State)
+
+let handle_state f i =
+  let rec loop s = function
+    | Term(State).Return x -> x
+    | Term(State).Op(Get, k) -> loop s (k s)
+    | Term(State).Op(Set s', k) -> loop s' (k ())
+  in
+  loop i (State_eff.reify_local f)
+\end{lstlisting}
+Finally we add an alias for the handler type and some convenient
+wrappers for the generic effects:
+\begin{lstlisting}[style=ocaml]
+type handler = Handler(State).t
+
+let get h =
+  State_eff.perform Get h
+
+let set h i =
+  State_eff.perform (Set i) h
+\end{lstlisting}
+Giving us the following interface:
+\begin{lstlisting}[style=ocaml]
+type handler
+
+val handle_state : (handler @ local -> 'a) @ local -> int -> 'a
+val get : handler @ local -> int
+val set : handler @ local -> int -> unit
+\end{lstlisting}
+Note that we are able to use \lstinline[style=ocaml]{reify_local} to allow
+the function passed to \lstinline[style=ocaml]{handle_state} to be local.
+
+This shows how to implement state using effect reflection, but note that
+the same interface can also be implemented direcly using mutable state
+and locality:
+\begin{lstlisting}[style=ocaml]
+type handler = int ref
+
+let handle_state f i = f (ref i)
+let get r = !r
+let set r i = r := i
+\end{lstlisting}
+This implementation gives the same behaviour as the one built using
+effect reflection. Crucially it is just as easy to reason about, which
+in turn means it is just as easy to reason about as the state monad. In
+some sense, the improved reasoning ability of the state monad over
+ordinary mutable state is precisely captured by the notion of locality.
+
+This transformation from effect relection to local mutable state works
+for any \emph{tail-resumptive} effect handler.
+
+\subsection{Lightweight effect polymorphism}
+
+The following function uses state to calculate the total of a list of
+integers:
+\begin{lstlisting}[style=ocaml]
+let total l =
+  handle_state (fun h ->
+    List.iter (fun x -> set h (get h + x)) l;
+    get h)
+\end{lstlisting}
+This relies on the locality of the function parameter of
+\lstinline[style=ocaml]{List.iter}:
+\begin{lstlisting}[style=ocaml]
+val iter : ('a -> unit) @ local -> 'a list -> unit
+\end{lstlisting}
+The \lstinline[style=ocaml]{local} annotation on the parameter ensures
+that any function passed will not escape the current region. This allows
+us to pass in a closure that closes over the
+\lstinline[style=ocaml]{local} handler \lstinline[style=ocaml]{h} and
+use it to perform some effects.
+
+This ability for higher-order functions like
+\lstinline[style=ocaml]{List.iter} to be reused with different ambient
+monads achieves the ``lightweight effect polymorphism'' described by ??
+et al.\cite{??} without requiring a separate notion of second-class
+function.
+
+\subsection{Example: Generators}
+
+A more realistic use of effect reflection is a \emph{generator}
+effect. This effect has a single operation:
+\begin{lstlisting}[style=ocaml]
+module Gen = struct
+  type 'a t =
+    | Gen : int -> unit t
+end
+\end{lstlisting}
+Along with a handler that turns computations using
+\lstinline[style=ocaml]{Gen} into streams of integers.
+\begin{lstlisting}[style=ocaml]
+module Gen_eff = Reflection(Gen)
+
+type ints =
+  | Finished
+  | More of int * (unit -> ints)
+
+let handle_gen f =
+  let rec loop = function
+    | Term(Gen).Return () -> Finished
+    | Term(Gen).Op(Gen i, k) -> More(i, k)
+  in
+  loop (Gen_eff.reify f)
+\end{lstlisting}
+
+Unlike the state handler, here we must use
+\lstinline[style=ocaml]{reify} rather than
+\lstinline[style=ocaml]{reify_local} because our intention is that the
+resulting stream of integers be a global value that can escape the
+current region. Since it closes over the continuation
+\lstinline[style=ocaml]{k}, we must use \lstinline[style=ocaml]{reify}
+and require that \lstinline[style=ocaml]{f} be global:
+\begin{lstlisting}[style=ocaml]
+val handle_gen : (Handler(Gen).t @ local -> unit) -> ints
+\end{lstlisting}
+
+\subsection{Example: Asynchronous I/O}
+
+Perhaps the most common use case for algebraic effects is for
+concurrency and asynchronous I/O. Implementing asynchronous I/O from
+scratch is beyond the scope of this paper, but we can build a simple
+version on top of an existing monadic concurrency library. Assuming we
+have a monadic future type \lstinline[style=ocaml]{'a Deferred.t} we can
+write concurrent code in direct style by using an effect with a single
+\lstinline[style=ocaml]{Await} operation that waits for a future to be
+completed:
+\begin{lstlisting}[style=ocaml]
+module Await : sig
+  type 'a t =
+    | Await : 'a Deferred.t -> 'a t
+end
+\end{lstlisting}
+
+\begin{lstlisting}[style=ocaml]
+module Await_eff = Reflection(Await)
+
+let handle_await f =
+  let rec loop = function
+    | Term(Await).Return x -> Deferred.return x
+    | Term(Await).Op(Await d, k) ->
+        Deferred.bind d (fun x -> loop (k x))
+  in
+  loop (Await_eff.reify f)
+
+let await h d =
+  Await_eff.perform (Await d) h
+\end{lstlisting}
+
+Using this we can write direct-style code using asynchronous I/O:
+\begin{lstlisting}[style=ocaml]
+let copy_file h src dst =
+  let s = await h (File.read src) in
+  await h (File.write dst s)
 \end{lstlisting}
@@
 
+\subsection{Nested effect handlers}
+
+A key feature of algebraic effects is how easy they are to compose. We
+can use state whilst using asynchronous I/O by simply nesting their
+handlers:
+\begin{lstlisting}[style=ocaml]
+handle_await (fun ch ->
+  handle_state i (fun sh ->
+    List.iter files (fun fl ->
+      let s = await ch (File.read fl) in
+      let i = int_of_string s in
+      set sh (i + get sh));
+    get sh))
+\end{lstlisting}
+
+This relies on the fact that the \lstinline[style=ocaml]{handle_state}
+uses \lstinline[style=ocaml]{reify_local} and so is able to operate on a
+local computation that closes over the local handler value
+\lstinline[style=ocaml]{ch}.
+
+If we tried to do something similar using our
+\lstinline[style=ocaml]{handle_gen} handler, then we get and error:
+\begin{lstlisting}[style=ocaml]
+handle_await (fun ch ->
+  handle_gen i (fun gh ->
+    List.iter files (fun fl ->
+      let s = await ch (File.read fl) in
+      let i = int_of_string s in
+      gen gh i)))
+(*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
+                     (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
+(*@ \textcolor{red}{ This value is local and would escape its region.} @*)
+\end{lstlisting}
+
+We could change the definition of \lstinline[style=ocaml]{handle_gen} to
+use \lstinline[style=ocaml]{reify_local} instead of
+\lstinline[style=ocaml]{reify}, but then we would lose the property that
+the generator we produce can be passed around freely. Instead we need to
+build a handler that can forward the \lstinline[style=ocaml]{Await}
+operation to different effect handlers at different points in the
+program's execution.
+
 \subsection{Effect sums}
 
+Composing handlers, like \lstinline[style=ocaml]{handle_gen}, that use
+\lstinline[style=ocaml]{reify} rather than
+\lstinline[style=ocaml]{reify_local} is a fundamentally more involved
+operation. Consider the case of generators that are able to use
+asynchronous I/O. We are given a computation that performs both
+\lstinline[style=ocaml]{Await} and \lstinline[style=ocaml]{Gen}
+operations and we'd like to turn it into a stream of integers.  We would
+like to be able to freely pass this stream around. In particular, we do
+not wish the stream to be prevented from escaping the scope of the
+current \lstinline[style=ocaml]{Await} handler.  However forcing parts
+of the stream requires performing additional
+\lstinline[style=ocaml]{Await} operations. Whenever we force part of the
+stream we'll need to provide a handler for those
+\lstinline[style=ocaml]{Await} operations.
+
+To acheive this more general form of composition, we require an
+additional capability in our effect reflection API. We need to be able
+to manipulate \lstinline[style=ocaml]{Handler} types for sums of
+effects.
 \begin{lstlisting}[style=ocaml]
@@
 
-As a Lawvere theory, \lstinline[style=ocaml]{Sum(L)(R)} is the sum of
-\lstinline[style=ocaml]{L} and \lstinline[style=ocaml]{R}. This means
-that the terms of its free algebra $T_{L + R}(\tau)$ are equivalent to
-values of the initial algebra $\mu X. T_L(\tau + F_L(X)$. We can expose
-this fact via the following functions:
-\begin{lstlisting}
+\lstinline[style=ocaml]{Sum(L)(R)} is the coproduct of the effects
+\lstinline[style=ocaml]{L} and \lstinline[style=ocaml]{R}.
+
+Terms in the free algebra of the sum of two effects
+(i.e. \lstinline[style=ocaml]{Term(Sum(L)(R)).t}) are isomorphic to a
+recursive datatype that is built from the terms of one of the effects
+and the operations of the other\cite{??}. We can definet this type as:
+\begin{lstlisting}[style=ocaml]
 module Half_term (L : Op) (R : Op) : sig
@@
     | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
-
-  let rec reflect_left : 'a Term(Sum(L)(R)).t -> (Handler(L).t @ local -> 'a Half_term(L)(R).t)
-   = function
-    | Return v -> fun _ -> Return v
-    | Op(Left op, k) -> fun l -> L.reflect (Op(op, (fun res -> reflect_left (k res) l)))
-    | Op(Right op, k) -> fun _ -> Op(op, (fun res -> reflect_left (k res)))
-
-  let rec reify_left : (Handler(L).t @ local -> 'a Half_term(L)(R).t) -> 'a Term(Sum(L)(R)).t =
-    fun h ->
+end
+\end{lstlisting}
+and expose the isomorphism between
+\lstinline[style=ocaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
+and \lstinline[style=ocaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
+\begin{lstlisting}[style=ocaml]
+module Half_reflection (L : Op) (R : Op) : sig
+  module L_eff = Reflection(L)
+  module R_eff = Reflection(R)
+  module Sum_eff = Reflection(Sum(L)(R))
+  module Proj = Project(L)(R)
+
+  let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
+    let rec loop lh = function
+      | Term(Sum(L)(R)).Return ->
+          Half_term(L)(R).Return v
+      | Term(Sum(L)(R)).Op(Left op, k) ->
+          loop lh (k (L_eff.perform op lh))
+      | Term(Sum(L)(R)).Op(Right op, k) ->
+          Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
+    in
+    loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
+
+  let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
+    match f lh with
+    | Return x -> x
+    | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
+end
+\end{lstlisting}
+
+Now we can use this machinery to conveniently build a handler of
+\lstinline[style=ocaml]{Gen} that also forwards operations from
+\lstinline[style=ocaml]{Await}:
+\begin{lstlisting}[style=ocaml]
+module Await_gen_eff = Half_reflection(Await)(Gen)
+
+type aints =
+  | Finished
+  | More of int * (Handler(Await).t @ local -> ints)
+
+let handle_gen_in_await f ah =
     let rec loop = function
-        | Return (Return x) -> Return x
-        | Return (Op(r, k)) -> Op(Right r, fun res -> reify_left (k res))
-        | Op(l, k) -> Op(Left l, fun res -> loop (k res))
+    | Half_term(Await)(Gen).Return () -> Finished
+    | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, k ())
   in
-      loop (L.reify h)
-
-end
-\end{lstlisting}
-
-\subsection{Examples}
-
-\subsubsection{State}
-
-\subsubsection{Concurrency}
-
-\subsubsection{Generators and concurrency}
+  loop (Await_gen_eff.reify f)
+\end{lstlisting}
+where \lstinline[style=ocaml]{handle_gen_in_await} has the type:
+\begin{lstlisting}[style=ocaml]
+val handle_gen_in_await :
+  (Handler(Await).t @ local -> Handler(Gen).t @ local -> 'a)
+  -> Handler(Await).t @ local
+  -> aints
+\end{lstlisting}
+
+The literature on lexical effect handlers often ignores this issue. For
+example, \cite{??} proposed a system which only supported directly
+nesting effect handlers and had no mechanism for writing a composable
+handler for generators that could be moved between different underlying
+asynchronous I/O handlers.
+
+This is because this composition requiring forwarding is much less
+common than nesting and it fundamentally requires each use of the
+\lstinline[style=ocaml]{Await} operation to perform work proportional to
+the number of handlers it is forwarded through -- often referred to as a
+``linear search''. There has been a tendency to treat this linear search
+as some form of deficiency of algebraic effect handlers, but it is
+simply a classic trade-off between the extra expressivity allowed by an
+indirection and the runtime cost of traversing that indirection. With
+our effect reflection API we are able to use nesting for the cases that
+it supports -- avoiding the cost of the indirection -- and fall back to
+forwarding for the more general cases.
+
+\subsection{Parameterized effects and handlers}
+
+The state handlers in this paper so far have all been restricted to
+\lstinline[style=ocaml]{int} state. This is a limitation of the effect
+reflection API that we have been using. We would like to support
+\emph{parameterized effects} -- allowing a single handler to be used
+with a whole family of effects. For example, a state handler
+parameterized by the type of the state.
+
+We can support parameterized effects by adding a module type for them,
+along with parameterized versions of \lstinline[style=ocaml]{Term} and
+\lstinline[style=ocaml]{Handler}:
+\begin{lstlisting}[style=ocaml]
+module type Op1 = sig type ('a, 'e) t end
+
+module Term1 (O : Op1) : sig
+  type ('a, 'e) t =
+    | Return : 'a @@ global -> 'a t
+    | Op : ('r, 'e) O.t @@ global * ('r -> 'a t) -> 'a t
+end
+
+module Handler1(O : Op1) : sig
+  type 'e t
+end
+\end{lstlisting}
+with a corresponding parameterized version of
+\lstinline[style=ocaml]{Reflection}. We can then define a parameterized
+state effect:
+\begin{lstlisting}[style=ocaml]
+module State = struct
+  type ('a, 'b) t =
+    | Get : 'b t
+    | Set : 'b -> unit t
+end
+\end{lstlisting}
+and write a handler that provides a parameterized interface for the
+state effect:
+\begin{lstlisting}[style=ocaml]
+type 's handler
+
+val handle : ('s handler @ local -> 'a) @ local -> 's -> 'a
+val get : 's handler @ local -> 's
+val set : 's handler @ local -> 's -> unit
+\end{lstlisting}
+
+The need for a new copy of the interface for each type constructor arity
+is an artifact of OCaml's approach to higher-kinded types. Languages
+with direct support for higher-kinded types should be able to support
+parameterized effects of arbitrary arity with a single interface.
 
@@
 
-Our language falls out naturally from its categorical semantics, so we
-focus on that. We intend to index the meaning of types by an ambient
-algebraic monad of effects: rather than have the semantics of each type
-be a set of elements, the semantics of each type is a mapping from the
-ambient monad to a set of elements.
+The design of effect reflection falls out naturally from a simple
+denotational semantics built on presheaves. The key idea is to index the
+meaning of types by an ambient algebraic monad of effects: rather than
+have the semantics of each type be a set of elements, the semantics of
+each type is a mapping from the ambient monad to a set of elements.
+
+\subsection{Algebraic monads}
+
+In order to index the meaning of our types by the ambient monad we need
+some way of representing that ambient monad. We restrict the ambient
+monad to be an \emph{algebraic monad}: one that can be expressed as the
+free algebra of an algebraic theory. Such monads have nice properties,
+in particular the sum of two algebraic monads always exists and is
+itself an algebraic monad.
+
+Most monads used in practice for programming are algebraic, with the
+notable exception of the continuation monad. The issue with trying to
+represent non-algebraic monads algebraically is essentially one of size:
+if you try to write down the collection of algebirac operations for the
+continuation monad you get something that is too large to be a set. This
+is closely related to the strict positivity restriction on inductive
+type definitions.
+
+We recall the definitions of algebraic signature, terms of a signature,
+algebraic theory and models of a theory.
+\begin{definition}[Algebraic signature]
+  An \emph{algebraic signature} is a set of operations with associated
+  arities.
+\end{definition}
+
+\begin{example}
+  The algebraic signature for boolean state is:
+  \begin{equation*}
+    \{ (\mathrm{get}, 2); (\mathrm{set\_true}, 1); (\mathrm{set\_false}, 1) \}
+  \end{equation*}
+\end{example}
+
+\begin{definition}[Term]
+  A \emph{term} $t$ of some algebraic signature $\Sigma$ is either a
+  free variable $x$ or $op(t_1, \ldots, t_k)$ where $op$ is an operation
+  of $\Sigma$ with arity $k$ and each $t_i$ is itself a term.
+\end{definition}
+
+\begin{definition}[Algebraic theory]
+  An algebraic theory consists of an algebraic signature $\Sigma$ and a set of
+  equations on terms of $\Sigma$.
+\end{definition}
+
+\begin{example}
+  The algebraic theory for boolean state adds the following equations to
+  the signature of boolean state:
+  \begin{align*}
+    &\mathrm{set\_true}(\mathrm{get}(x, y)) = \mathrm{set\_true}(x)
+    &\mathrm{set\_false}(\mathrm{get}(x, y)) = \mathrm{set\_false}(y) \\
+    &\mathrm{set\_true}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
+    &\mathrm{set\_true}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
+    &\mathrm{set\_false}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
+    &\mathrm{set\_false}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
+    &\mathrm{get}(\mathrm{set\_true}(x), \mathrm{set\_false}(x))) = x
+  \end{align*}
+\end{example}
+
+\begin{definition}[Model]
+
+\end{definition}
+
+Arities in algebraic signatures are often considered as natural numbers,
+but they can equivalently be just sets, with families of terms indexed
+by that set used as the arguments of that operator in a term. It is also
+convenient to allow families of operations parameterized by some set. As
+such, we can present a signature in the form:
+\begin{equation*}
+\{ op_1 \types P_1 \rightarrowtail A_1; \ldots; op_n
+\types P_n \rightarrowtail A_n \}
+\end{equation*}
+where $op \types P \rightarrowtail A$ indicates an operation $op$
+parameterised by $P$ and with arity $A$.
+
+Traditionally, algebraic signatures are restricted to the case of a
+finite set of operations each with a finite arity. Most theoretical work
+in the algebraic effects literature extends this to the case of
+countable sets of operations with countable arities, which allows us to
+construct the signature for natural number state:
+
+\begin{example}
+  The algebraic signature for natural number state is:
+  \begin{equation*}
+    \{ \mathrm{get} \types 1 \rightarrowtail \nat; \mathrm{set} \types \nat \rightarrowtail 1 \}
+  \end{equation*}
+\end{example}
+
+However, it is not sufficient to construct the signature for state
+containing a function $\nat \rightarrow \nat$. For that we must extend
+algebraic signatures further to allow for arbitrary sets of operations
+with arbitrary sets as arities. Note that we still require that these
+collections be sets, so we continue to exclude monads like the
+continuation monad.
+
+\subsection{(Bounded) Lawvere theories}
+
+We choose \emph{Lawvere theories} as our concrete representation of the
+ambient algebraic monad. Lawvere theories are a categorical
+representation of algebraic theories. However, we will mostly describe
+Lawvere theories and constructions on them in terms of their
+presentations as algebraic theories, so an understanding the details of
+Lawvere theories is not required for understanding the rest of this
+paper. We refer the interested reader to \cite{??} for a detailed
+explanation of Lawvere theories.
+
+Traditional Lawvere theories can only represent algebraic signatures
+with a finite set of operations with finite arities. We instead use a
+more general notion: \emph{bounded Lawvere theories}. These allow the
+set of operations and their arities to be any (small) set. This allows
+our effect operations to have arbitrary simple types as parameters and
+results. In Appendix \ref{??}, we prove that all the properties of
+Lawvere theories that we rely on can be extended from the finitary to
+the bounded case.
+
+Bounded Lawvere theories form a category $\lawc$ whose morphisms are
+maps from the operations of one theory to the operations of another that
+preserve the equations of the first theory according to the equations of
+the second.
+
+$\lawc$ has an initial object $\initial$, which is the algebraic theory
+with no operations. It also has coproducts $L + M$ which is the theory
+made by unioning the operations and equations of $L$ and $M$.
+
+\subsection{Presheaves}
 
 For a category $\cat$, $\psh = \setc^{\catop}$ is the category of
-presheaves over $\cat$. If $\cat$ has a terminal object $\terminal$, we
-can define the global modality $\glob \mathrel{:} [\psh, \psh]$ such that:
+presheaves over $\cat$. It's objects are contravariant functors from
+$\cat$ to $\setc$ and it's morphisms are natural transformations between
+such functors.
+
+A common use of presheaves is where $\cat$ has finite products and is
+thought of as some form of context over which everything is
+parameterized. In our case we wish to be parameterized by a context of
+\emph{algebraic effects that can be mapped into the ambient monad}. Note
+that the effects theories in such a context are \emph{contravariant}:
+they describe effect theories that can be handled by the ambient
+monad. As such, we use
 \begin{equation*}
-  \glob(P)(L) = P(\terminal)
+  \pshlawcop = \setc^{\lawc}
 \end{equation*}
-
-We use the ambient monad as our notion of location, so we take $\cat$ to
-be $\lawcop$, where $\lawc$ is the category of countable Lawvere
-theories. Lawvere theories are a categorical representation of algebraic
-theories. Morphisms in $\lawc$ are maps from the operations of one
-theory to the operations of another that preserve the equations of the
-first theory according to the equations of the second.  The initial
-theory is the algebraic theory with no operations. The coproduct of two
-theories is the theory made by combining the operations and equations of
-both theories.
+as the basis of our semantics. The $\op$ in the definition of presheaves
+and the $\op$ due to our dependence being contravariant cancel out, but
+it is still best to think in terms of presheaves as that is a more
+precise description of what is going on here.
+
+Our semantics will use objects from $\pshlawcop$ to represent types and
+contexts, and morphisms in $\pshlawcop$ to represent terms. This allows
+their meanings to be parameterized by the ambient monad. By using
+presheaves rather than ordinary functions we ensure terms and types at a
+particular ambient monad can always be mapped naturally to a larger
+ambient monad.
+
+Categories of presheaves have products and sums, which are defined
+pointwise. We will use these as the semantics of product types and sum
+types in our denotational semantics. Such categories also have
+exponentials, which we will use as the semantics of functions types.
+
+\subsection{The free algebra monad}
 
 Given a Lawvere theory $L$, the underlying set of its free algebra forms
-a monad $T_L$ on $\setc$. $T_{\_}$ is functorial and we can use it to
-construct a monad $\mon$ on $\pshlawcop$ in the obvious way, such that:
+a monad $T_L$ on $\setc$. Given a set $A$, the values $T_L(A)$ are terms
+build using the elements of $A$ and the operations of $L$ quotiented by
+the equations in $L$.
+
+\todo{Add example}
+
+$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
+$\pshlawcop$ such that:
 \begin{equation*}
@@
 \end{equation*}
-We will use $\mon$ as our computation monad.
-
-$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding. For each
+$\mon$ is essentially the ambient monad, and in our denotational
+semantics, we will use it as our computation monad.
+
+\subsection{The Yoneda embedding}
+
+$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding: for each
 Lawvere theory $L$ there is a presheaf $\yoneda{L}$ in $\pshlawcop$ such
 that:
+\begin{align*}
+&\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}} \\
+&\yoneda{L}(l^{\prime}) = \lambda f.\; l^{\prime} \circ f
+\end{align*}
+We will use $\yoneda{L}$ as the semantics for our handler types. Their
+elements are just maps from $L$ into the ambient monad.
+
+The Yoneda embedding preserves finite products, which for our
+case means:
 \begin{equation*}
-\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}}
+\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
 \end{equation*}
-We will use $\yoneda{L}$ to represent maps from $L$ into the ambient
-monad. Note that $\yonedaname$ preserves finite products:
-$\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}$
-where $+$ is the coproduct in $\lawc$, which is the product in
-$\lawcop$.
-
-Categories of presheaves are cartesian closed, and the exponential
-$\arrow{P}{Q}$ in $\pshlawcop$ maps $L$ to
-$\homset{\pshlawcop}{\yoneda{L} \times P}{Q}$. So, via the Yoneda lemma,
-we get the following:
-\begin{align*}
-(\arrow{\yoneda{L}}{\mon(\glob A)})(L') \quad \cong \quad &T_{(L' + L)}(A(\terminal))
-\end{align*}
+
+\subsection{The global modality}
+
+For any category $\cat$, if $\cat$ has a terminal object $\terminal$, we
+can define the global modality $\glob \types [\psh, \psh]$ such that:
+$\glob(P)(L) = P(\terminal)$. In the case of $\pshlawcop$ this becomes:
+\begin{equation*}
+\glob(P)(L) = P(\initial)
+\end{equation*}
+
+Within our semantics, this amounts to restricting a type to only the
+elements which do not depend on the ambient monad. This is very similar
+to the purity comonad \todo{check name} in ?? \cite{??}. In particular,
+\begin{equation*}
+  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob
+  \cong \quad \glob P
+\end{equation*}
+as elements of the free algebra of the initial algebraic theory are just
+elements of the carrier set.
+
+For any non-empty Lawvere theory $L$, $\glob \yoneda{L}$ is the constant
+presheave that always returns the empty set because there are no
+morphisms from $L$ into $\initial$. In our denotational semantics this
+amounts to all non-trivial global handler types being uninhabited.
+
+\subsection{The Yoneda lemma}
+
+Via the Yoneda lemma, we get the following:
+\begin{equation*}
+(\arrow{\yoneda{L}}{\mon(\glob A)})(L') \quad \cong \quad T_{(L' + L)}(A(\terminal))
+\end{equation*}
 which will form the basis of our language's isomorphisms between arrows
@@
 
+Exponentials in categories of presheaves are defined such that:
+\begin{equation*}
+\arrow{P}{Q}(L) = \homset{\pshlawcop}{\yoneda{L} \times P}{Q}
+\end{equation*}
+
+\subsection{W-types and effect sums}
+
+\subsection{The effect reflection isomorphism}
+
 \section{A language with effect reflection}
@@
 
-We build a simple language to (mostly) achieve our aim. Rather than
-working with arbitrary algebraic monads as promised, we follow the
-finest traditions of the algebraic effects literature and restrict our
-attention to algebraic theories without equations. Our language is
-simply typed, featuring all the usual type-formers. It uses a variation
-of fine-grained call-by-value: it has separate syntax and typing
-judgements for values and computations.
-
-We define algebraic signatures $\Sigma$ as finite sets of operations
+We can now build a simple language with effect reflection and give it a
+semantics in terms of presheaves. Our language is simply typed,
+featuring all the usual type-formers. It uses a variation of
+fine-grained call-by-value: it has separate syntax and typing judgements
+for values and computations. The syntax is in Fig.\ref{fig:syntax} and 
+the full typing rules are in Fig.\ref{fig:typing}.
+\begin{figure}
+\begin{align*}
+  \mathrm{types}, \tau ::=\;& \tau \times \tau \mid 1 \mid \tau + \tau
+                              \mid \tau \rightarrow \tau \mid
+                              \ind{\Sigma}(\tau) \mid \yoneda{\Sigma}
+                              \mid \glob \tau \\
+  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \gbind{\Gamma}{x}{\tau} \\
+  \mathrm{values}, v :=\;& x \mid (v, v) \mid \fst v \mid \snd v
+                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \\
+                         &\mid \indintro{\Sigma}{op}{v}{v} \mid \ret{\Sigma}{v}
+                           \mid \gbox{v} \mid \gunbox{x}{v}{v} \mid \grun{c} \\
+                            &\mid \outl{v} \mid \outr{v} \\
+  \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
+                                 \mid \app{v}{v} \mid \\
+                            & \case{v}{x}{c}{x}{c} \mid \\
+                            & \indelim{\Sigma}{v}
+                              {\indretcase{x}{c} \seq
+                              \indcase{op}{x}{x}{c} \seq \ldots
+                              \seq \indcase{op}{x}{x}{c}} \\
+                            & \perform{v}{op}{v}{x}{c} \mid \reify{\Sigma}{x}{c}
+\end{align*}
+\caption{Syntax}
+\label{fig:syntax}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\inferrule{\\}
+{\lbind{\Gamma}{x}{\tau} \turnv x \types \tau}
+\and
+\inferrule{\Gamma \turnv v_1 \types \tau_1 \\ \Gamma \turnv v_1 \types \tau_1}
+{\Gamma \turnv (v_1, v_2) \types \tau_1 \times \tau_2}
+\and
+\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
+{\Gamma \turnv \fst{v} \types \tau_1}
+\and
+\inferrule{\Gamma \turnv v \types \tau_1 \times \tau_2}
+{\Gamma \turnv \snd{v} \types \tau_2}
+\and
+\inferrule{\\}
+{\Gamma \turnv () \types 1}
+\and
+\inferrule{\Gamma \turnv v \types \tau_1}
+{\Gamma \turnv \inl{v} \types \tau_1 + \tau_2}
+\and
+\inferrule{\Gamma \turnv v \types \tau_2}
+{\Gamma \turnv \inr{v} \types \tau_1 + \tau_2}
+\and
+\inferrule{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2}
+{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2}
+\and
+\inferrule{\Gamma \turnv v \types \tau}
+{\Gamma \turnc \return{v} \types \tau}
+\and
+\inferrule{\Gamma \turnc c_1 \types \tau_1 \\
+           \lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}
+{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2}
+\and
+\inferrule{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2 \\
+           \Gamma \turnv v_2 \types \tau_1}
+{\Gamma \turnc v_1 v_2 \types \tau_2}
+\and
+\inferrule{\Gamma \turnv v \types \tau_1 + \tau_2 \\
+           \lbind{\Gamma}{x_1}{\tau_1} \turnc c_1 \types \tau_3 \\
+           \lbind{\Gamma}{x_2}{\tau_2} \turnc c_2 \types \tau_3}
+{\Gamma \turnc \case{v}{x_1}{c_1}{x_2}{c_2} \types \tau_3}
+\end{mathpar}
+\caption{Standard typing rules}
+\label{fig:typing-standard}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\inferrule{\\}
+{\gbind{\Gamma}{x}{\tau} \turnv x \types \tau}
+\and
+\inferrule{\cdiv{\Gamma} \turnv v \types \tau}
+{\Gamma \turnv \gbox{v} \types \glob \tau}
+\and
+\inferrule{\Gamma \turnv v_1 \types \glob \tau_1 \\
+           \gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
+{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2}
+\end{mathpar}
+\caption{Locality typing rules}
+\label{fig:typing-locality}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\inferrule{op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
+           \Gamma \turnv v_1 \types \glob \tau_1 \\
+           \Gamma \turnv v_2 \types \glob \tau_2 \rightarrow \ind{\Sigma}(\tau)}
+{\Gamma \turnv \indintro{\Sigma}{op}{v_1}{v_2} \types \ind{\Sigma}(\tau)}
+\and
+\inferrule{\Gamma \turnv v \types \tau}
+{\Gamma \turnv \ret{\Sigma}{v} \types \ind{\Sigma}(\tau)}
+\and
+\inferrule{\Gamma \turnv v \types \ind{\Sigma}(\tau_1) \\
+           \Sigma = \{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1};
+             \ldots; op_n\types \tau_{p_n} \rightarrowtail \tau_{a_n} \} \\
+           \lbind{\Gamma}{x_r}{\tau_1} \turnc c_r \types \tau_2 \\
+           \lbind{\lbind{\Gamma}{x_{p_1}}{\glob \tau_{p_1}}}{x_{a_1}}{\glob \tau_{a_1} \rightarrow \tau_2} \turnc c_1 \types \tau_2 \\
+           \ldots \\
+           \lbind{\lbind{\Gamma}{x_{p_n}}{\glob \tau_{p_n}}}{x_{a_n}}{\glob \tau_{a_n} \rightarrow \tau_2} \turnc c_n \types \tau_2}
+{\Gamma \turnc \indelim{\Sigma}{v}{\indretcase{x_r}{c_r} \seq
+                                   \indcase{op_1}{x_{p_1}}{x_{a_1}}{c_1} \seq \ldots \seq
+                                   \indcase{op_n}{x_{p_n}}{x_{a_n}}{c_n}} \types \tau_2}
+\end{mathpar}
+\caption{Inductive type typing rules}
+\label{fig:typing-inductive}
+\end{figure}
+
+\begin{figure}
+\begin{mathpar}
+\inferrule{\cdiv{\Gamma} \turnc c \types \tau}
+{\Gamma \turnv \grun{c} \types \tau}
+\and
+\inferrule{\Gamma \turnv v \types \yoneda{\Sigma_1 + \Sigma_2}}
+{\Gamma \turnv \outl{v} \types \yoneda{\Sigma_1}}
+\and
+\inferrule{\Gamma \turnv v \types \yoneda{\Sigma_1 + \Sigma_2}}
+{\Gamma \turnv \outr{v} \types \yoneda{\Sigma_2}}
+\and
+\inferrule{\Gamma \turnv v_1 \types \yoneda{\Sigma} \\
+           op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
+           \Gamma \turnv v_2 \types \glob \tau_1 \\
+           \lbind{\Gamma}{x}{\glob \tau_2} \turnc c \types \tau_3}
+{\Gamma \turnc \; \perform{v_1}{op}{v_2}{x}{c} \types \tau_3}
+\and
+\inferrule{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}
+{\Gamma \turnc \; \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)}
+\end{mathpar}
+\caption{Reflection typing rules}
+\label{fig:typing-reflection}
+\end{figure}
+
+As with other algebraic effects systems in the literature, we only
+support handlers for theories without any equations, even though our
+semantics is defined in terms of arbitrary bounded Lawvere theories. We
+define algebraic signatures $\Sigma$ as finite sets of operations
 \[
-\{ op_1 \mathop{:} \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
-\mathop{:} \tau_{p_n} \rightarrowtail \tau_{a_n} \}
+\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
+\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
 \]
-An operation type
-$\tau_p \rightarrowtail \tau_a$ indicates an operation parameterised by
-$\tau_p$ and with arity $\tau_a$. Such signatures correspond to Lawvere
-theories without equations. We restrict parameters and arities to base
-types to ensure that our theories only have a countable number of
-operations of countable arities. We write $\sem{\Sigma}$ for the
-corresponding Lawvere theory, and $F_\Sigma$ for the endofunctor on
-$\setc$ for which $T_{\sem{\Sigma}}$ is the free monad.
+An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
+parameterised by $\tau_p$ and with arity $\tau_a$. Such signatures
+correspond to bounded Lawvere theories without equations.
 
@@
 $\tau_a \rightarrow \ind{\Sigma}(\tau)$ for each operation
-$op \mathop{:} \tau_p \rightarrowtail \tau_a$ in $\Sigma$, as well as a
-constructor $ret$ with an argument of type $\tau$.
+$op \types \tau_p \rightarrowtail \tau_a$ in $\Sigma$, as well as a
+constructor $ret$ with an argument of type $\tau$. $\ind{\Sigma}$ is
+sometimes called the \emph{freer monad} of $\Sigma$ \cite{??}.
 
@@
 isomorphism appears in the language as two operations, reflection
-$\reflect{\Sigma}{c}$ and reification $\reify{\Sigma}{v}$.
+$\reflectraw{\Sigma}{op}$ and reification $\reifyraw{\Sigma}$.
 \begin{mathpar}
-\inferrule{\Gamma \mathrel{\vdash_C} c : \ind{\Sigma}(\glob \tau)}
-{\Gamma \mathrel{\vdash_V} \; \reflect{\Sigma}{c} : \yoneda{\Sigma} \rightarrow \glob \tau}
-\and
-\inferrule{\Gamma \mathrel{\vdash_V} v : \yoneda{\Sigma} \rightarrow \glob \tau}
-{\Gamma \mathrel{\vdash_C} \; \reify{\Sigma}{v} : \ind{\Sigma}(\glob \tau)}
+\inferrule{\Gamma \turnv v_1 \types \yoneda{\Sigma} \\
+           op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
+           \Gamma \turnv v_2 \types \glob \tau_1 \\
+           \lbind{\Gamma}{x}{\glob \tau_2} \turnc c \types \tau_3}
+{\Gamma \turnc \; \perform{v_1}{op}{v_2}{x}{c} \types \tau_3}
+\and
+\inferrule{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}
+{\Gamma \turnc \; \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)}
 \end{mathpar}
 
+Note that we treat reflection of individual operations
+$\performraw{v}{op}$ as primitive. We can define reflection of
+terms $\reflectraw{\Sigma}$ in terms of $\performraw{v}{op}$:
+\[
+    \reflect{\Sigma}{v}{c} := \todo{write this}
+\]
+which has this admissible typing rule:
+\begin{mathpar}
+\inferrule{\Gamma \turnv v \types \yoneda{\Sigma} \\
+           \Gamma \turnv c \types \ind{\Sigma}(\glob \tau)}
+{\Gamma \turnc \; \reflect{\Sigma}{v}{c} \types \glob \tau}
+\end{mathpar}
+
 Our denotational semantics in terms of $\pshlawcop$ is
-straightforward. For types, the translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
+straightforward. We write $\sem{\Sigma}$ for the Lawvere theory
+corresponding to $\Sigma$, and $F_\Sigma$ for the endofunctor on $\setc$
+for which $T_{\sem{\Sigma}}$ is the free monad. For types, the
+translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
 \begin{mathpar}
@@
 \begin{align*}
-&\sem{\Gamma \vdash_V v \mathop{:} \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
-&\sem{\glob \Gamma \vdash_V \mathop{\mathrm{box}} c \mathop{:} \glob \tau} = \glob \sem{\Gamma \vdash_C c \mathop{:} \tau}
+&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
+&\sem{\glob \Gamma \turnv \gbox{v} \types \glob \tau} =
+  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
+&\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
+  \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
+  \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle
+&\sem{\glob \Gamma \turnv \grun{c} \types \glob \tau} =
+  \epsilon \circ \todo{removing identity monad} \circ
+  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
 \end{align*}
 \begin{align*}
-&\sem{\Gamma \vdash_C c \mathop{:} \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
-&\sem{\Gamma \vdash_C \mathop{\mathrm{return}} v \mathop{:} \tau} = \eta_{\sem{\tau}} \circ \sem{\Gamma \vdash_V v \mathop{:} \tau} \\
-&\sem{\Gamma \vdash_C \mathop{\mathrm{let}} x : \tau_1 = c_1 \mathop{\mathrm{in}} c_2\mathop{:} \tau_2} = \mu_{\sem{\tau_2}} \circ \mon(\sem{\Gamma; x \mathop{:} \tau_1 \vdash_C C_2 \mathop{:} \tau_2}) \circ t_{\sem{\Gamma}, \sem{\tau_1}} \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \vdash_C C_1 \mathop{:} \tau} \rangle \\
-&\sem{\Gamma \vdash_C \mathop{\mathrm{unbox}} c \mathop{:} \tau} = \mon(\epsilon_{\sem{\tau}}) \circ \sem{\Gamma \vdash_C c \mathop{:} \glob \tau} \\
-&\sem{\Gamma \vdash_C \reflect{\Sigma}{c} \mathop{:} \yoneda{\Sigma} \rightarrow \glob \tau} = \reflect{\Sigma\tau}{\sem{\Gamma \vdash_C c \mathop{:} \ind{\Sigma}(\glob \tau)}} \\
-&\sem{\Gamma \vdash_C \reify{\Sigma}{c} \mathop{:} \ind{\Sigma}(\glob \tau)} = \reify{\Sigma\tau}{\sem{\Gamma \vdash_C c \mathop{:} \yoneda{\Sigma} \rightarrow \glob \tau}} \\
+&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
+&\sem{\Gamma \turnc \return{v} \types \tau} = \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
+&\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} = \mu_{\sem{\tau_2}} \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}) \circ t_{\sem{\Gamma}, \sem{\tau_1}} \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau} \rangle \\
+&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \glob \tau} = \todo{fix}
+  \evmap \langle \reflectc{\Sigma}{\tau} \circ \eta \circ
+                   \sem{\Gamma \turnv v_1 \types \ind{\Sigma}(\glob \tau)},\,
+                 \sem{\Gamma \turnv v_2 \types \yoneda{\Sigma}} \rangle \\
+&\sem{\Gamma \turnc \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)} =
+  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}) \\
 \end{align*}
 where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
-respectively of $\mon$; $\epsilon$ is the counit of $\glob$;
-$\reflect{\Sigma}{}$ and $\reify{\Sigma}{}$ are the natural
-transformation in the isomorphism from the previous section.
+respectively of $\mon$; $\epsilon$ is the counit of $\glob$; $\evmap$
+and $\currymap$ are the evaluation map and currying map of exponentials;
+$\reflectc{\Sigma}{}$ and $\reifyc{\Sigma}{}$ are the natural
+transformations in the isomorphism from the previous section.
+
+\section{Operational semantics}
+\label{sec:operational}
+
+\begin{figure}
+\begin{align*}
+  \mathrm{neutral\ values}, \eta :=\;& x \mid \fst \eta \mid \snd \eta
+                                     \mid \gunbox{x}{\eta}{\omega} \mid \grun{\mu} \\
+  \mathrm{neutral\ computations}, \mu :=\;& \letv{x}{\mu}{\gamma}
+                                           \mid \app{\eta}{\omega} \mid \\
+                                   & \case{\eta}{x}{\gamma}{x}{\gamma} \mid \\
+                            & \indelim{\Sigma}{\eta}
+                              {\indretcase{x}{\gamma} \seq
+                               \indcase{op}{x}{x}{\gamma} \seq \ldots \seq
+                               \indcase{op}{x}{x}{\gamma}} \\
+                                   & \reify{\Sigma}{x}{\mu} \\
+  \mathrm{normal\ values}, \omega :=\;& \eta \mid (\omega, \omega)
+                           \mid () \mid \inl{\omega} \mid \inr{\omega} \mid \abs{x}{\tau}{\gamma} \mid \\
+                           &\indintro{\Sigma}{op}{\omega}{\omega} \mid \ret{\Sigma}{\omega} \mid \\
+                           &\gbox{\omega} \mid \outl{\omega} \mid \outr{\omega} \\
+  \mathrm{normal\ computations}, \gamma :=\;& \mu \mid \return{\omega}
+                                               \mid \perform{\eta}{op}{\omega}{x}{\gamma}
+\end{align*}
+\caption{Normal forms}
+\label{fig:normal}
+\end{figure}
+
+\begin{figure}
+\begin{align*}
+\fst{(v_1, v_2)} &\reducesto v_1 \\
+\snd{(v_1, v_2)} &\reducesto v_2 \\
+\gunbox{x}{\gbox{v_1}}{v_2} &\reducesto \sbst{v_2}{x}{v_1} \\
+\grun{\return{v}} &\reducesto v \\
+\letv{x}{\return{v}}{c} &\reducesto \sbst{c}{x}{v} \\
+\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducesto \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
+\app{(\abs{x}{\tau}{c})}{v} &\reducesto \sbst{c}{x}{v} \\
+\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducesto \sbst{c_1}{x_1}{v} \\
+\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducesto \sbst{c_2}{x_2}{v} \\
+\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} &\reducesto \sbst{c}{x}{v} \\
+\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
+  \reducesto \sbst{\sbst{c}{x_1}{v_2}}{x_2}{\abs{x_a}{\tau_a}{&\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}} \\
+\reify{\Sigma}{x}{\return{v}} &\reducesto \ret{\Sigma}{v} \\
+\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} &\reducesto \indintro{\Sigma}{op}{v}{\abs{x_2}{\reify{\Sigma}{x_1}{c}}} \\
+\reify{\Sigma}{x_1}{\perform{x_2}{op}{v}{x_3}{c}} &\reducesto \perform{x_2}{op}{v}{x_3}{\reify{\Sigma}{x_1}{c}} \qquad (x_1 \neq x_2) \\
+\perform{(\outl{v_1})}{op}{v_2}{x}{c} &\reducesto \perform{v_1}{\mathrm{left}(op)}{v_2}{x}{c} \\
+\perform{(\outr{v_1})}{op}{v_2}{x}{c} &\reducesto \perform{v_1}{\mathrm{right}(op)}{v_2}{x}{c}
+\end{align*}
+\caption{Reduction relation}
+\label{fig:reduction}
+\end{figure}
 
@@
 \begin{lstlisting}[style=ocaml]
+module Handler (O : Op) = struct
+  type t = { perform : 'r. 'r O.t -> 'r }
+end
+
 module Reflection (O : Op) = struct
-
-  type 'a term =
-    | Return : 'a @@ global -> 'a term
-    | Op : 'r O.t @@ global * ('r -> 'a term) -> 'a term
-
-  type handler = { perform : 'r. 'r O.t -> 'r }
 
@@
     in
-    let handler = { perform = fun o -> Effect.perform (Handler.H o) } in
+    let handler : Handler(O).t =
+      { perform = fun o -> Effect.perform (Handler.H o) }
+    in
     match f handler with
@@
 
+\section{Appendix}
+
+Previous theoretical work on algebraic effects, such as \cite{??},
+has tended to use \emph{countable Lawvere theories}. These correspond to
+monads who rank is \todo{how should I write the countable
+  cardinality}. This still restricts the parameters and arity operations
+to be countable. We instead use \emph{bounded Lawvere
+  theories}, corresponding to monads that have some cardinal as their
+rank. This allows parameters and arities of operations to be arbitrary
+simple types.
+
+Bounded Lawvere theories are not given a consistent name in the
+literature. A number of results that are proved for countable Lawvere
+theories are asserted to be true for the bounded case. As we could not
+find actual proofs for these claims we have included such proofs in
+Appendix \ref{??}. We also include a proof of an interesting property
+of bounded Lawvere theories:
+
+\todo{How to format and label this?}  {\b Bounded Lawvere theories
+  correspond precisely to those (infinitary) Lawvere theories whose
+  term monad can be expressed as a W-type}
+
 \end{document}

### 2025-07-06 paper.tex
Continue to write

@@
 
-\lstdefinestyle{ocaml}{
+\lstdefinestyle{oxcaml}{
   keywordstyle=\color{blue},
@@
 \newcommand{\cat}{\mathbb{C}}
+\newcommand{\catobj}{C}
 \newcommand{\op}{\mathtt{op}}
@@
 \newcommand{\ind}[1]{W_{#1}}
-\newcommand{\indintro}[4]{#2_{#1}(#3, #4)}
+\newcommand{\indintro}[4]{#2_{#1}(#3,\, #4)}
 \newcommand{\retname}{\mathrm{ret}}
 \newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
-\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \{#3\}}
+\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \,\{#3\}}
 \newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
@@
 \newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
-\newcommand{\gunbox}[4]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
+\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
 \newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
@@
 \newcommand{\currymap}{\mathop{\lambda}}
-\newcommand{\reducesto}{\rightsquigarrow}
+\newcommand{\reducestov}{\rightsquigarrow_V}
+\newcommand{\reducestoc}{\rightsquigarrow_C}
+\newcommand{\mif}{\mathop{\mathrm{if}}}
+\newcommand{\mthen}{\mathop{\mathrm{then}}}
+\newcommand{\melse}{\mathop{\mathrm{else}}}
+\newcommand{\embed}[1]{\lbag #1 \rbag}
 
@@
 For example, in OCaml, one can write:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 type 'a Effect.t +=
@@
 \end{lstlisting}
-and for the duration of the \lstinline[style=ocaml]{f ()} function call
+and for the duration of the \lstinline[style=oxcaml]{f ()} function call
 the ambient monad is extended with two additional effectful operations:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 let get () : int = perform Get
@@
 effect operation that is not supported by the current ambient monad,
-resulting in an \lstinline[style=ocaml]{Unhandled} exception being
+resulting in an \lstinline[style=oxcaml]{Unhandled} exception being
 raised. Other languages, e.g. Koka\cite{??}, use an effect system to
@@
 backwards compatibility with OCaml, an unadorned arrow type
-\lstinline[style=ocaml]{t -> s} is treated as if the parameter and
-return types were actually \lstinline[style=ocaml]{[]t} and
-\lstinline[style=ocaml]{[]s}. To get parameters or results that are not
+\lstinline[style=oxcaml]{t -> s} is treated as if the parameter and
+return types were actually \lstinline[style=oxcaml]{[]t} and
+\lstinline[style=oxcaml]{[]s}. To get parameters or results that are not
 implicitly global, we write
-e.g. \lstinline[style=ocaml]{t @ local -> s @ local}.
+e.g. \lstinline[style=oxcaml]{t @ local -> s @ local}.
 
@@
 it to write an effect handler equivalent to the OCaml one above:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module State = struct
@@
 \end{lstlisting}
-which calls \lstinline[style=ocaml]{f} and passes it a local value of
-type \lstinline[style=ocaml]{Handler(State).t}. For the duration of the
-call to \lstinline[style=ocaml]{f} the ambient monad is extended with
+which calls \lstinline[style=oxcaml]{f} and passes it a local value of
+type \lstinline[style=oxcaml]{Handler(State).t}. For the duration of the
+call to \lstinline[style=oxcaml]{f} the ambient monad is extended with
 two additional effectful operations, which can be used via reflection:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 let get (h : Handler(State).t @ local) : int =
@@
 one of them: \emph{locality}. Ignoring the other axes, a mode is either
-\lstinline[style=ocaml]{local} or \lstinline[style=ocaml]{global}.
+\lstinline[style=oxcaml]{local} or \lstinline[style=oxcaml]{global}.
 
@@
 the entire extent of the call. This means that, by default,
-\lstinline[style=ocaml]{local} values cannot be returned from a
+\lstinline[style=oxcaml]{local} values cannot be returned from a
 function.
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 let leak_local x =
@@
 To support functions that return local values, the region of a function
-can be ended early using the \lstinline[style=ocaml]{exclave} construct.
-\begin{lstlisting}[style=ocaml]
+can be ended early using the \lstinline[style=oxcaml]{exclave} construct.
+\begin{lstlisting}[style=oxcaml]
 let return_local x =
@@
 \end{lstlisting}
-\lstinline[style=ocaml]{exclave e} can only appear in tail position
+\lstinline[style=oxcaml]{exclave e} can only appear in tail position
 within a function.  It ends the function's region early and then
 executes the expression
-\lstinline[style=ocaml]{e}. \lstinline[style=ocaml]{e} is essentially
+\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
 executed within the region of the caller of the function and any local
@@
 The place where modes appear within the type algebra is on arrow
-types. A function being \lstinline[style=ocaml]{global} doesn't tell you
+types. A function being \lstinline[style=oxcaml]{global} doesn't tell you
 that the values it accepts or returns should be
-\lstinline[style=ocaml]{global} too, so instead the arrow type is
+\lstinline[style=oxcaml]{global} too, so instead the arrow type is
 labelled with two modes: one for the argument and one for the return
 value.
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 val return_local : 'a @ local -> 'a option @ local
@@
 be omitted, in which case it defaults to
-\lstinline[style=ocaml]{global}.
-
-By default, modes are deep: a local \lstinline[style=ocaml]{string list}
+\lstinline[style=oxcaml]{global}.
+
+By default, modes are deep: a local \lstinline[style=oxcaml]{string list}
 is a local list of local strings. However, the depth that a mode applies
 can be controlled via \emph{modalities}. Record fields can be annotated
-with \lstinline[style=ocaml]{global} modality to indicate that the
+with \lstinline[style=oxcaml]{global} modality to indicate that the
 contents of the field are global even when the surrounding record is
 local. For example:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 type 'a gbl = { g : 'a @@ global }
 \end{lstlisting}
-gives a record whose only field has the \lstinline[style=ocaml]{global}
-modality. The contents of the \lstinline[style=ocaml]{g} field is always
+gives a record whose only field has the \lstinline[style=oxcaml]{global}
+modality. The contents of the \lstinline[style=oxcaml]{g} field is always
 global, for example:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 let project_global (r @ local) =
@@
 gives a function of type:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 val project_global : 'a gbl @ local -> 'a @ global
@@
 
-Note that, whilst there is a \lstinline[style=ocaml]{global} modality,
-there is no \lstinline[style=ocaml]{local} modality. It does not make
+Note that, whilst there is a \lstinline[style=oxcaml]{global} modality,
+there is no \lstinline[style=oxcaml]{local} modality. It does not make
 sense for a value which has the ability to leave a region to contain a
@@
 implicitly cause the second one to leave that region as well. This means
-the \lstinline[style=ocaml]{global} mode is always deep.
+the \lstinline[style=oxcaml]{global} mode is always deep.
 
@@
 functors, so we start with a simple module type
-\lstinline[style=ocaml]{Op} to act as the parameter type for these
+\lstinline[style=oxcaml]{Op} to act as the parameter type for these
 functors:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module type Op = sig type 'a t end
 \end{lstlisting}
-The fundamental types of the API are the \lstinline[style=ocaml]{Term}
-and \lstinline[style=ocaml]{Handler} types.
-\begin{lstlisting}[style=ocaml]
+The fundamental types of the API are the \lstinline[style=oxcaml]{Term}
+and \lstinline[style=oxcaml]{Handler} types.
+\begin{lstlisting}[style=oxcaml]
 module Term (O : Op) : sig
@@
 \end{lstlisting}
-For an effect signature \lstinline[style=ocaml]{O},
-\lstinline[style=ocaml]{'a Term(O).t} is the type of terms in the free
+For an effect signature \lstinline[style=oxcaml]{O},
+\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
 algebra of the corresponding effect -- otherwise known as the freer
-monad\cite{??}. \lstinline[style=ocaml]{Handler(O).t} is the type of
-mappings from the operations of \lstinline[style=ocaml]{O} into the
+monad\cite{??}. \lstinline[style=oxcaml]{Handler(O).t} is the type of
+mappings from the operations of \lstinline[style=oxcaml]{O} into the
 operations of the current ambient monad. Alternatively,
-\lstinline[style=ocaml]{Handler(O).t} can be described as representing
+\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
 algebraic effect handlers that are handling the current computation --
@@
 \begin{equation*}
-  \text{\lstinline[style=ocaml]{'a Term(O).t}} \qquad
-  \cong \qquad \text{\lstinline[style=ocaml]{Handler(O).t @ local -> 'a}}
+  \text{\lstinline[style=oxcaml]{'a Term(O).t}} \qquad
+  \cong \qquad \text{\lstinline[style=oxcaml]{Handler(O).t @ local -> 'a}}
 \end{equation*}
 Due to the absence of mode polymorphism in OxCaml, we expose this
-isomorphism as three functions: \lstinline[style=ocaml]{reify} and
-\lstinline[style=ocaml]{reify_local} which implement the left-to-right
-direction at the \lstinline[style=ocaml]{global} and
-\lstinline[style=ocaml]{local} modes respectively; and
-\lstinline[style=ocaml]{perform}, which is equivalent to the
+isomorphism as three functions: \lstinline[style=oxcaml]{reify} and
+\lstinline[style=oxcaml]{reify_local} which implement the left-to-right
+direction at the \lstinline[style=oxcaml]{global} and
+\lstinline[style=oxcaml]{local} modes respectively; and
+\lstinline[style=oxcaml]{perform}, which is equivalent to the
 right-to-left direction but makes for a more convenient primitive in
-practice. \lstinline[style=ocaml]{perform} amounts to reflection of
+practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
 individual operations as opposed to full terms.
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Reflection (O : Op) : sig
@@
 \end{lstlisting}
-Using \lstinline[style=ocaml]{perform} we can implement reflection of
+Using \lstinline[style=oxcaml]{perform} we can implement reflection of
 terms as:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 let rec reflect t h =
@@
 with types:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
@@
 state using this API. We start by defining the operations of the effect:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module State = struct
@@
 it to write a handler:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module State_eff = Reflection(State)
@@
 wrappers for the generic effects:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 type handler = Handler(State).t
@@
 Giving us the following interface:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 type handler
@@
 \end{lstlisting}
-Note that we are able to use \lstinline[style=ocaml]{reify_local} to allow
-the function passed to \lstinline[style=ocaml]{handle_state} to be local.
+Note that we are able to use \lstinline[style=oxcaml]{reify_local} to allow
+the function passed to \lstinline[style=oxcaml]{handle_state} to be local.
 
@@
 and locality:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 type handler = int ref
@@
 integers:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 let total l =
@@
 This relies on the locality of the function parameter of
-\lstinline[style=ocaml]{List.iter}:
-\begin{lstlisting}[style=ocaml]
+\lstinline[style=oxcaml]{List.iter}:
+\begin{lstlisting}[style=oxcaml]
 val iter : ('a -> unit) @ local -> 'a list -> unit
 \end{lstlisting}
-The \lstinline[style=ocaml]{local} annotation on the parameter ensures
+The \lstinline[style=oxcaml]{local} annotation on the parameter ensures
 that any function passed will not escape the current region. This allows
 us to pass in a closure that closes over the
-\lstinline[style=ocaml]{local} handler \lstinline[style=ocaml]{h} and
+\lstinline[style=oxcaml]{local} handler \lstinline[style=oxcaml]{h} and
 use it to perform some effects.
@@
 This ability for higher-order functions like
-\lstinline[style=ocaml]{List.iter} to be reused with different ambient
+\lstinline[style=oxcaml]{List.iter} to be reused with different ambient
 monads achieves the ``lightweight effect polymorphism'' described by ??
@@
 effect. This effect has a single operation:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Gen = struct
@@
 Along with a handler that turns computations using
-\lstinline[style=ocaml]{Gen} into streams of integers.
-\begin{lstlisting}[style=ocaml]
+\lstinline[style=oxcaml]{Gen} into streams of integers.
+\begin{lstlisting}[style=oxcaml]
 module Gen_eff = Reflection(Gen)
@@
 Unlike the state handler, here we must use
-\lstinline[style=ocaml]{reify} rather than
-\lstinline[style=ocaml]{reify_local} because our intention is that the
+\lstinline[style=oxcaml]{reify} rather than
+\lstinline[style=oxcaml]{reify_local} because our intention is that the
 resulting stream of integers be a global value that can escape the
 current region. Since it closes over the continuation
-\lstinline[style=ocaml]{k}, we must use \lstinline[style=ocaml]{reify}
-and require that \lstinline[style=ocaml]{f} be global:
-\begin{lstlisting}[style=ocaml]
+\lstinline[style=oxcaml]{k}, we must use \lstinline[style=oxcaml]{reify}
+and require that \lstinline[style=oxcaml]{f} be global:
+\begin{lstlisting}[style=oxcaml]
 val handle_gen : (Handler(Gen).t @ local -> unit) -> ints
@@
 version on top of an existing monadic concurrency library. Assuming we
-have a monadic future type \lstinline[style=ocaml]{'a Deferred.t} we can
+have a monadic future type \lstinline[style=oxcaml]{'a Deferred.t} we can
 write concurrent code in direct style by using an effect with a single
-\lstinline[style=ocaml]{Await} operation that waits for a future to be
+\lstinline[style=oxcaml]{Await} operation that waits for a future to be
 completed:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Await : sig
@@
 
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Await_eff = Reflection(Await)
@@
 Using this we can write direct-style code using asynchronous I/O:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 let copy_file h src dst =
@@
 OCaml, our actual implementation is restricted one-shot
-continuations. Note that OxCaml's \lstinline[style=ocaml]{once}
+continuations. Note that OxCaml's \lstinline[style=oxcaml]{once}
 mode\cite{lorenzen2024oxidizing} allows us to enforce this restriction
@@
 handlers:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 handle_await (fun ch ->
@@
 
-This relies on the fact that the \lstinline[style=ocaml]{handle_state}
-uses \lstinline[style=ocaml]{reify_local} and so is able to operate on a
+This relies on the fact that the \lstinline[style=oxcaml]{handle_state}
+uses \lstinline[style=oxcaml]{reify_local} and so is able to operate on a
 local computation that closes over the local handler value
-\lstinline[style=ocaml]{ch}.
+\lstinline[style=oxcaml]{ch}.
 
 If we tried to do something similar using our
-\lstinline[style=ocaml]{handle_gen} handler, then we get and error:
-\begin{lstlisting}[style=ocaml]
+\lstinline[style=oxcaml]{handle_gen} handler, then we get and error:
+\begin{lstlisting}[style=oxcaml]
 handle_await (fun ch ->
@@
 
-We could change the definition of \lstinline[style=ocaml]{handle_gen} to
-use \lstinline[style=ocaml]{reify_local} instead of
-\lstinline[style=ocaml]{reify}, but then we would lose the property that
+We could change the definition of \lstinline[style=oxcaml]{handle_gen} to
+use \lstinline[style=oxcaml]{reify_local} instead of
+\lstinline[style=oxcaml]{reify}, but then we would lose the property that
 the generator we produce can be passed around freely. Instead we need to
-build a handler that can forward the \lstinline[style=ocaml]{Await}
+build a handler that can forward the \lstinline[style=oxcaml]{Await}
 operation to different effect handlers at different points in the
@@
 
-Composing handlers, like \lstinline[style=ocaml]{handle_gen}, that use
-\lstinline[style=ocaml]{reify} rather than
-\lstinline[style=ocaml]{reify_local} is a fundamentally more involved
+Composing handlers, like \lstinline[style=oxcaml]{handle_gen}, that use
+\lstinline[style=oxcaml]{reify} rather than
+\lstinline[style=oxcaml]{reify_local} is a fundamentally more involved
 operation. Consider the case of generators that are able to use
 asynchronous I/O. We are given a computation that performs both
-\lstinline[style=ocaml]{Await} and \lstinline[style=ocaml]{Gen}
+\lstinline[style=oxcaml]{Await} and \lstinline[style=oxcaml]{Gen}
 operations and we'd like to turn it into a stream of integers.  We would
@@
 not wish the stream to be prevented from escaping the scope of the
-current \lstinline[style=ocaml]{Await} handler.  However forcing parts
+current \lstinline[style=oxcaml]{Await} handler.  However forcing parts
 of the stream requires performing additional
-\lstinline[style=ocaml]{Await} operations. Whenever we force part of the
+\lstinline[style=oxcaml]{Await} operations. Whenever we force part of the
 stream we'll need to provide a handler for those
-\lstinline[style=ocaml]{Await} operations.
+\lstinline[style=oxcaml]{Await} operations.
 
@@
 additional capability in our effect reflection API. We need to be able
-to manipulate \lstinline[style=ocaml]{Handler} types for sums of
+to manipulate \lstinline[style=oxcaml]{Handler} types for sums of
 effects.
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Sum (L : Op) (R : Op) : sig
@@
 
-\lstinline[style=ocaml]{Sum(L)(R)} is the coproduct of the effects
-\lstinline[style=ocaml]{L} and \lstinline[style=ocaml]{R}.
+\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
+\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}.
 
 Terms in the free algebra of the sum of two effects
-(i.e. \lstinline[style=ocaml]{Term(Sum(L)(R)).t}) are isomorphic to a
+(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
 recursive datatype that is built from the terms of one of the effects
 and the operations of the other\cite{??}. We can definet this type as:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Half_term (L : Op) (R : Op) : sig
@@
 and expose the isomorphism between
-\lstinline[style=ocaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
-and \lstinline[style=ocaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
-\begin{lstlisting}[style=ocaml]
+\lstinline[style=oxcaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
+and \lstinline[style=oxcaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
+\begin{lstlisting}[style=oxcaml]
 module Half_reflection (L : Op) (R : Op) : sig
@@
 Now we can use this machinery to conveniently build a handler of
-\lstinline[style=ocaml]{Gen} that also forwards operations from
-\lstinline[style=ocaml]{Await}:
-\begin{lstlisting}[style=ocaml]
+\lstinline[style=oxcaml]{Gen} that also forwards operations from
+\lstinline[style=oxcaml]{Await}:
+\begin{lstlisting}[style=oxcaml]
 module Await_gen_eff = Half_reflection(Await)(Gen)
@@
 \end{lstlisting}
-where \lstinline[style=ocaml]{handle_gen_in_await} has the type:
-\begin{lstlisting}[style=ocaml]
+where \lstinline[style=oxcaml]{handle_gen_in_await} has the type:
+\begin{lstlisting}[style=oxcaml]
 val handle_gen_in_await :
@@
 common than nesting and it fundamentally requires each use of the
-\lstinline[style=ocaml]{Await} operation to perform work proportional to
+\lstinline[style=oxcaml]{Await} operation to perform work proportional to
 the number of handlers it is forwarded through -- often referred to as a
@@
 The state handlers in this paper so far have all been restricted to
-\lstinline[style=ocaml]{int} state. This is a limitation of the effect
+\lstinline[style=oxcaml]{int} state. This is a limitation of the effect
 reflection API that we have been using. We would like to support
@@
 We can support parameterized effects by adding a module type for them,
-along with parameterized versions of \lstinline[style=ocaml]{Term} and
-\lstinline[style=ocaml]{Handler}:
-\begin{lstlisting}[style=ocaml]
+along with parameterized versions of \lstinline[style=oxcaml]{Term} and
+\lstinline[style=oxcaml]{Handler}:
+\begin{lstlisting}[style=oxcaml]
 module type Op1 = sig type ('a, 'e) t end
@@
 with a corresponding parameterized version of
-\lstinline[style=ocaml]{Reflection}. We can then define a parameterized
+\lstinline[style=oxcaml]{Reflection}. We can then define a parameterized
 state effect:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module State = struct
@@
 state effect:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 type 's handler
@@
 
-\section{Presheaves for algebraic effects}
-\label{sec:semantics}
-
-The design of effect reflection falls out naturally from a simple
-denotational semantics built on presheaves. The key idea is to index the
-meaning of types by an ambient algebraic monad of effects: rather than
-have the semantics of each type be a set of elements, the semantics of
-each type is a mapping from the ambient monad to a set of elements.
-
-\subsection{Algebraic monads}
-
-In order to index the meaning of our types by the ambient monad we need
-some way of representing that ambient monad. We restrict the ambient
-monad to be an \emph{algebraic monad}: one that can be expressed as the
-free algebra of an algebraic theory. Such monads have nice properties,
-in particular the sum of two algebraic monads always exists and is
-itself an algebraic monad.
-
-Most monads used in practice for programming are algebraic, with the
-notable exception of the continuation monad. The issue with trying to
-represent non-algebraic monads algebraically is essentially one of size:
-if you try to write down the collection of algebirac operations for the
-continuation monad you get something that is too large to be a set. This
-is closely related to the strict positivity restriction on inductive
-type definitions.
-
-We recall the definitions of algebraic signature, terms of a signature,
-algebraic theory and models of a theory.
-\begin{definition}[Algebraic signature]
-  An \emph{algebraic signature} is a set of operations with associated
-  arities.
-\end{definition}
-
-\begin{example}
-  The algebraic signature for boolean state is:
-  \begin{equation*}
-    \{ (\mathrm{get}, 2); (\mathrm{set\_true}, 1); (\mathrm{set\_false}, 1) \}
-  \end{equation*}
-\end{example}
-
-\begin{definition}[Term]
-  A \emph{term} $t$ of some algebraic signature $\Sigma$ is either a
-  free variable $x$ or $op(t_1, \ldots, t_k)$ where $op$ is an operation
-  of $\Sigma$ with arity $k$ and each $t_i$ is itself a term.
-\end{definition}
-
-\begin{definition}[Algebraic theory]
-  An algebraic theory consists of an algebraic signature $\Sigma$ and a set of
-  equations on terms of $\Sigma$.
-\end{definition}
-
-\begin{example}
-  The algebraic theory for boolean state adds the following equations to
-  the signature of boolean state:
-  \begin{align*}
-    &\mathrm{set\_true}(\mathrm{get}(x, y)) = \mathrm{set\_true}(x)
-    &\mathrm{set\_false}(\mathrm{get}(x, y)) = \mathrm{set\_false}(y) \\
-    &\mathrm{set\_true}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
-    &\mathrm{set\_true}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
-    &\mathrm{set\_false}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
-    &\mathrm{set\_false}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
-    &\mathrm{get}(\mathrm{set\_true}(x), \mathrm{set\_false}(x))) = x
-  \end{align*}
-\end{example}
-
-\begin{definition}[Model]
-
-\end{definition}
-
-Arities in algebraic signatures are often considered as natural numbers,
-but they can equivalently be just sets, with families of terms indexed
-by that set used as the arguments of that operator in a term. It is also
-convenient to allow families of operations parameterized by some set. As
-such, we can present a signature in the form:
-\begin{equation*}
-\{ op_1 \types P_1 \rightarrowtail A_1; \ldots; op_n
-\types P_n \rightarrowtail A_n \}
-\end{equation*}
-where $op \types P \rightarrowtail A$ indicates an operation $op$
-parameterised by $P$ and with arity $A$.
-
-Traditionally, algebraic signatures are restricted to the case of a
-finite set of operations each with a finite arity. Most theoretical work
-in the algebraic effects literature extends this to the case of
-countable sets of operations with countable arities, which allows us to
-construct the signature for natural number state:
-
-\begin{example}
-  The algebraic signature for natural number state is:
-  \begin{equation*}
-    \{ \mathrm{get} \types 1 \rightarrowtail \nat; \mathrm{set} \types \nat \rightarrowtail 1 \}
-  \end{equation*}
-\end{example}
-
-However, it is not sufficient to construct the signature for state
-containing a function $\nat \rightarrow \nat$. For that we must extend
-algebraic signatures further to allow for arbitrary sets of operations
-with arbitrary sets as arities. Note that we still require that these
-collections be sets, so we continue to exclude monads like the
-continuation monad.
-
-\subsection{(Bounded) Lawvere theories}
-
-We choose \emph{Lawvere theories} as our concrete representation of the
-ambient algebraic monad. Lawvere theories are a categorical
-representation of algebraic theories. However, we will mostly describe
-Lawvere theories and constructions on them in terms of their
-presentations as algebraic theories, so an understanding the details of
-Lawvere theories is not required for understanding the rest of this
-paper. We refer the interested reader to \cite{??} for a detailed
-explanation of Lawvere theories.
-
-Traditional Lawvere theories can only represent algebraic signatures
-with a finite set of operations with finite arities. We instead use a
-more general notion: \emph{bounded Lawvere theories}. These allow the
-set of operations and their arities to be any (small) set. This allows
-our effect operations to have arbitrary simple types as parameters and
-results. In Appendix \ref{??}, we prove that all the properties of
-Lawvere theories that we rely on can be extended from the finitary to
-the bounded case.
-
-Bounded Lawvere theories form a category $\lawc$ whose morphisms are
-maps from the operations of one theory to the operations of another that
-preserve the equations of the first theory according to the equations of
-the second.
-
-$\lawc$ has an initial object $\initial$, which is the algebraic theory
-with no operations. It also has coproducts $L + M$ which is the theory
-made by unioning the operations and equations of $L$ and $M$.
-
-\subsection{Presheaves}
-
-For a category $\cat$, $\psh = \setc^{\catop}$ is the category of
-presheaves over $\cat$. It's objects are contravariant functors from
-$\cat$ to $\setc$ and it's morphisms are natural transformations between
-such functors.
-
-A common use of presheaves is where $\cat$ has finite products and is
-thought of as some form of context over which everything is
-parameterized. In our case we wish to be parameterized by a context of
-\emph{algebraic effects that can be mapped into the ambient monad}. Note
-that the effects theories in such a context are \emph{contravariant}:
-they describe effect theories that can be handled by the ambient
-monad. As such, we use
-\begin{equation*}
-  \pshlawcop = \setc^{\lawc}
-\end{equation*}
-as the basis of our semantics. The $\op$ in the definition of presheaves
-and the $\op$ due to our dependence being contravariant cancel out, but
-it is still best to think in terms of presheaves as that is a more
-precise description of what is going on here.
-
-Our semantics will use objects from $\pshlawcop$ to represent types and
-contexts, and morphisms in $\pshlawcop$ to represent terms. This allows
-their meanings to be parameterized by the ambient monad. By using
-presheaves rather than ordinary functions we ensure terms and types at a
-particular ambient monad can always be mapped naturally to a larger
-ambient monad.
-
-Categories of presheaves have products and sums, which are defined
-pointwise. We will use these as the semantics of product types and sum
-types in our denotational semantics. Such categories also have
-exponentials, which we will use as the semantics of functions types.
-
-\subsection{The free algebra monad}
-
-Given a Lawvere theory $L$, the underlying set of its free algebra forms
-a monad $T_L$ on $\setc$. Given a set $A$, the values $T_L(A)$ are terms
-build using the elements of $A$ and the operations of $L$ quotiented by
-the equations in $L$.
-
-\todo{Add example}
-
-$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
-$\pshlawcop$ such that:
-\begin{equation*}
-\mon(P)(L) = T_L(P(L))
-\end{equation*}
-$\mon$ is essentially the ambient monad, and in our denotational
-semantics, we will use it as our computation monad.
-
-\subsection{The Yoneda embedding}
-
-$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding: for each
-Lawvere theory $L$ there is a presheaf $\yoneda{L}$ in $\pshlawcop$ such
-that:
-\begin{align*}
-&\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}} \\
-&\yoneda{L}(l^{\prime}) = \lambda f.\; l^{\prime} \circ f
-\end{align*}
-We will use $\yoneda{L}$ as the semantics for our handler types. Their
-elements are just maps from $L$ into the ambient monad.
-
-The Yoneda embedding preserves finite products, which for our
-case means:
-\begin{equation*}
-\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
-\end{equation*}
-
-\subsection{The global modality}
-
-For any category $\cat$, if $\cat$ has a terminal object $\terminal$, we
-can define the global modality $\glob \types [\psh, \psh]$ such that:
-$\glob(P)(L) = P(\terminal)$. In the case of $\pshlawcop$ this becomes:
-\begin{equation*}
-\glob(P)(L) = P(\initial)
-\end{equation*}
-
-Within our semantics, this amounts to restricting a type to only the
-elements which do not depend on the ambient monad. This is very similar
-to the purity comonad \todo{check name} in ?? \cite{??}. In particular,
-\begin{equation*}
-  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob
-  \cong \quad \glob P
-\end{equation*}
-as elements of the free algebra of the initial algebraic theory are just
-elements of the carrier set.
-
-For any non-empty Lawvere theory $L$, $\glob \yoneda{L}$ is the constant
-presheave that always returns the empty set because there are no
-morphisms from $L$ into $\initial$. In our denotational semantics this
-amounts to all non-trivial global handler types being uninhabited.
-
-\subsection{The Yoneda lemma}
-
-Via the Yoneda lemma, we get the following:
-\begin{equation*}
-(\arrow{\yoneda{L}}{\mon(\glob A)})(L') \quad \cong \quad T_{(L' + L)}(A(\terminal))
-\end{equation*}
-which will form the basis of our language's isomorphisms between arrows
-$Y \rightarrow \glob A$ and algebraic monads $M(\glob A)$.
-
-Exponentials in categories of presheaves are defined such that:
-\begin{equation*}
-\arrow{P}{Q}(L) = \homset{\pshlawcop}{\yoneda{L} \times P}{Q}
-\end{equation*}
-
-\subsection{W-types and effect sums}
-
-\subsection{The effect reflection isomorphism}
-
 \section{A language with effect reflection}
@@
 
-We can now build a simple language with effect reflection and give it a
-semantics in terms of presheaves. Our language is simply typed,
-featuring all the usual type-formers. It uses a variation of
-fine-grained call-by-value: it has separate syntax and typing judgements
-for values and computations. The syntax is in Fig.\ref{fig:syntax} and 
-the full typing rules are in Fig.\ref{fig:typing}.
+The design of effect reflection falls out naturally from a semantics in
+terms of presheaves. To illustrate this semantics we first build a
+simple language with effect reflection.
+
+\subsection{The standard parts}
+
+Our language is simply typed, featuring all the usual type-formers. It
+uses a variation of fine-grained call-by-value: it has separate syntax
+and typing judgements for values and computations. The syntax and typing
+rules for these standard parts of the language are shown in
+Fig.\ref{fig:typing-standard}, and there isn't much interesting to say
+about them.
 \begin{figure}
@@
   \mathrm{types}, \tau ::=\;& \tau \times \tau \mid 1 \mid \tau + \tau
-                              \mid \tau \rightarrow \tau \mid
-                              \ind{\Sigma}(\tau) \mid \yoneda{\Sigma}
-                              \mid \glob \tau \\
-  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \gbind{\Gamma}{x}{\tau} \\
+                              \mid \tau \rightarrow \tau \mid \ldots \\
+  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \ldots \\
   \mathrm{values}, v :=\;& x \mid (v, v) \mid \fst v \mid \snd v
-                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \\
-                         &\mid \indintro{\Sigma}{op}{v}{v} \mid \ret{\Sigma}{v}
-                           \mid \gbox{v} \mid \gunbox{x}{v}{v} \mid \grun{c} \\
-                            &\mid \outl{v} \mid \outr{v} \\
+                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \mid \ldots \\
   \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
                                  \mid \app{v}{v} \mid \\
-                            & \case{v}{x}{c}{x}{c} \mid \\
-                            & \indelim{\Sigma}{v}
-                              {\indretcase{x}{c} \seq
-                              \indcase{op}{x}{x}{c} \seq \ldots
-                              \seq \indcase{op}{x}{x}{c}} \\
-                            & \perform{v}{op}{v}{x}{c} \mid \reify{\Sigma}{x}{c}
-\end{align*}
-\caption{Syntax}
-\label{fig:syntax}
-\end{figure}
-
-\begin{figure}
+                            & \case{v}{x}{c}{x}{c} \mid \ldots
+\end{align*}
 \begin{mathpar}
@@
 \end{mathpar}
-\caption{Standard typing rules}
+\caption{Standard syntax and typing rules}
 \label{fig:typing-standard}
@@
 
+\subsection{Locality}
+
+We track locality in the language via a global comonadic modality
+$\glob$. In our setting being global means not depending on the ambient
+monad. For example, the function type $1 \rightarrow 1$ represents
+computations in the ambient monad -- e.g. if the ambient monad includes
+state then such a function may mutate that state -- whereas the global
+function type $\glob (1 \rightarrow 1)$ has only a single value: the
+function that does nothing and returns $()$.
+
+When using $\gbox{v}$ to build a value with a global type, we can only
+use existing values that also have a global type. We achieve this using
+a fairly standard approach for modal types: we extend contexts with a
+binding form that requires a global type and then we restrict the
+context in which we type $v$ to only include such bindings. The new
+global binding form is written $\gbind{\Gamma}{x}{\tau}$ and represents
+a value in the context of type $\glob \tau$. We create global bindings
+in the context using the elimination form: $\gunbox{x}{v_1}{v_2}$.
+
+Restricting the context is implemented via an operation on contexts
+$\cdiv{\Gamma}$ that removes all local bindings, leaving only the global
+ones:
+\begin{align*}
+  &\cdiv{\cempty} = \cempty \\
+  &\cdiv{(\lbind{\Gamma}{x}{\tau})} = \cdiv{\Gamma} \\
+  &\cdiv{(\gbind{\Gamma}{x}{\tau})} = \gbind{\cdiv{\Gamma}}{x}{\tau}
+\end{align*}
+
+Global computations do not depend on the ambient monad, so we also
+include a value form $\grun{c}$, which allows treating a global
+computation $c$ as a value.
+
+The syntax and typing rules related to locality are in
+Fig.\ref{fig:typing-locality}
 \begin{figure}
+\begin{align*}
+  \mathrm{types}, \tau ::=\;& \ldots \mid \glob \tau \mid \ldots \\
+  \mathrm{contexts}, \Gamma ::=\;& \ldots \mid \gbind{\Gamma}{x}{\tau} \\
+  \mathrm{values}, v :=\;& \ldots \mid \gbox{v} \mid \gunbox{x}{v}{v}
+                           \mid \grun{c} \mid \ldots
+\end{align*}
 \begin{mathpar}
@@
 {\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2}
+\and
+\inferrule{\cdiv{\Gamma} \turnc c \types \tau}
+{\Gamma \turnv \grun{c} \types \tau}
 \end{mathpar}
-\caption{Locality typing rules}
+\caption{Locality syntax and typing rules}
 \label{fig:typing-locality}
@@
 
+\subsection{Signatures and inductive types}
+
+Effect reflection requires, for any signature $\Sigma$ that is being
+reflected, an inductive type $\ind{\Sigma}$ to represent the free
+algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
+  monad} of $\Sigma$ \cite{??}.
+
+As with other algebraic effects systems in the literature, we only
+support handlers for algebraic theories without any equations, even
+though our semantics is defined in terms of arbitrary theories. We
+define algebraic signatures $\Sigma$ as finite sets of operations:
+\begin{equation*}
+\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
+\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
+\end{equation*}
+An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
+parameterised by $\tau_p$ and with arity $\tau_a$.
+
+\begin{example}
+  The signature for boolean state, which we'll call $\Sigma_{bs}$
+  is:
+  \begin{equation*}
+    \{ \mathrm{get} \types 1 \rightarrowtail (1 + 1);
+       \mathrm{set} \types (1 + 1) \rightarrowtail 1 \}
+  \end{equation*}
+\end{example}
+
+We also allow the sum of two signatures $\Sigma_1 + \Sigma_2$ to be used
+as a signature. It has operations:
+\begin{align*}
+\{ \; &\mathrm{left}(op_{1,1}) \types \tau_{p_{1,1}} \rightarrowtail \tau_{a_{1,1}} \seq
+        \ldots \seq
+        \mathrm{left}(op_{1, n}) \types \tau_{p_{1, n}} \rightarrowtail \tau_{a_{1, n}} \\
+      &\mathrm{right}(op_{2, 1}) \types \tau_{p_{2, 1}} \rightarrowtail \tau_{a_{2,1}} \seq
+        \ldots \seq
+        \mathrm{right}(op_{2, m}) \types \tau_{p_{2, m}} \rightarrowtail \tau_{a_{2, m}}
+        \; \}
+\end{align*}
+where $op_{1,1} \seq \ldots \seq op_{1,n}$ are the operations of
+$\Sigma_1$ and $op_{2,1} \seq \ldots \seq op_{2,m}$ are the operations
+of $\Sigma_2$.
+
+For each operation $op \types \tau_p \rightarrowtail \tau_a$,
+$\ind{\Sigma}(\tau)$ has a constructor $\indintro{\Sigma}{op}{v_p}{v_k}$
+where $v_p \types \tau_p$ and
+$v_k \types \tau_a \rightarrow \ind{\Sigma}(\tau)$. $\ind{\Sigma}(\tau)$
+also has a constructor $\ret{\Sigma}{v}$ where $v \types \tau$.
+
+The eliminator for $\ind{\Sigma}$ is written
+\begin{align*}
+\indelim{\Sigma}{v}
+  {&\indretcase{x}{c_r} \seq \\
+   &\indcase{op_1}{x_{p_1}}{x_{k_1}}{c_1} \seq \\
+   &\ldots \seq \\
+   &\indcase{op_n}{x_{p_n}}{x_{k_n}}{c_n}}
+\end{align*}
+The syntax and typing rules for inductive types are shown in
+Fig.\ref{fig:typing-inductive}.
+
 \begin{figure}
+\begin{align*}
+  \mathrm{types}, \tau ::=\;& \ldots \mid \ind{\Sigma}(\tau) \mid \ldots \\
+  \mathrm{signatures}, \Sigma ::=\;& \{ op \types \tau \rightarrowtail \tau \seq \ldots
+                                       \seq op \types \tau \rightarrowtail \tau \}
+                                     \mid \Sigma + \Sigma \\
+  \mathrm{values}, v :=\;& \ldots \mid \indintro{\Sigma}{op}{v}{v}
+                           \mid \ret{\Sigma}{v} \mid \ldots \\
+  \mathrm{computations}, c :=\;& \ldots \mid \indelim{\Sigma}{v}
+                                 {\indretcase{x}{c} \seq
+                                 \indcase{op}{x}{x}{c} \seq \ldots
+                                 \seq \indcase{op}{x}{x}{c}} \mid \ldots
+\end{align*}
 \begin{mathpar}
@@
            \lbind{\Gamma}{x_r}{\tau_1} \turnc c_r \types \tau_2 \\
-           \lbind{\lbind{\Gamma}{x_{p_1}}{\glob \tau_{p_1}}}{x_{a_1}}{\glob \tau_{a_1} \rightarrow \tau_2} \turnc c_1 \types \tau_2 \\
+           \lbind{\lbind{\Gamma}{x_{p_1}}{\glob \tau_{p_1}}}{x_{k_1}}{\glob \tau_{a_1} \rightarrow \tau_2} \turnc c_1 \types \tau_2 \\
            \ldots \\
-           \lbind{\lbind{\Gamma}{x_{p_n}}{\glob \tau_{p_n}}}{x_{a_n}}{\glob \tau_{a_n} \rightarrow \tau_2} \turnc c_n \types \tau_2}
+           \lbind{\lbind{\Gamma}{x_{p_n}}{\glob \tau_{p_n}}}{x_{k_n}}{\glob \tau_{a_n} \rightarrow \tau_2} \turnc c_n \types \tau_2}
 {\Gamma \turnc \indelim{\Sigma}{v}{\indretcase{x_r}{c_r} \seq
-                                   \indcase{op_1}{x_{p_1}}{x_{a_1}}{c_1} \seq \ldots \seq
-                                   \indcase{op_n}{x_{p_n}}{x_{a_n}}{c_n}} \types \tau_2}
+                                   \indcase{op_1}{x_{p_1}}{x_{k_1}}{c_1} \seq \ldots \seq
+                                   \indcase{op_n}{x_{p_n}}{x_{k_n}}{c_n}} \types \tau_2}
 \end{mathpar}
-\caption{Inductive type typing rules}
+\caption{Inductive type syntax and typing rules}
 \label{fig:typing-inductive}
@@
 
+\begin{example}
+  \label{example:negate}
+  The following value of type $\ind{\Sigma_{bs}}$ represents a
+  reified computation that negates the value of the state:
+  \begin{align*}
+    &\indintro{\Sigma_{bs}}{get}{()}{\abs{x_a}{1 + 1}{\\
+        &\qquad \letv{x_n}{\app{\mathit{negate}}{x_a}}{\\
+          &\qquad \indintro{\Sigma_{bs}}{set}{x_n}{\abs{x_u}{1}{\ret{\Sigma_{bs}}{x_u}}}}}}
+  \end{align*}
+  where $\mathit{negate}$ is the function that negates a boolean.
+\end{example}
+
+\subsection{Effect reflection}
+
+The language provides a type $\yoneda{\Sigma}$ for each $\Sigma$ to
+represent mappings from the operations of $\Sigma$ into the current
+ambient monad. Effect reflection appears in the language as two
+operations: reflection and reification. Reflection
+$\reflectraw{\Sigma}{op}$ uses a value of type $\yoneda{\Sigma}$ to map
+the operation $op$ from $\Sigma$ into the ambient monad.  Reification
+$\reifyraw{\Sigma}$ extends the ambient monad with the operations from
+$\Sigma$, providing a value of type $\yoneda{\Sigma}$ for constructing
+those operations via reflection. Their syntax and typing rules are shown
+in Fig.\ref{fig:typing-reflection}.
+
 \begin{figure}
+\begin{align*}
+  \mathrm{types}, \tau ::=\;& \ldots \mid \yoneda{\Sigma} \mid \ldots \\
+  \mathrm{values}, v :=\;& \ldots \mid \outl{v} \mid \outr{v} \\
+  \mathrm{computations}, c :=\;& \ldots \perform{v}{op}{v}{x}{c} \mid \reify{\Sigma}{x}{c}
+\end{align*}
 \begin{mathpar}
-\inferrule{\cdiv{\Gamma} \turnc c \types \tau}
-{\Gamma \turnv \grun{c} \types \tau}
-\and
 \inferrule{\Gamma \turnv v \types \yoneda{\Sigma_1 + \Sigma_2}}
@@
 \end{mathpar}
-\caption{Reflection typing rules}
+\caption{Effect reflection syntax and typing rules}
 \label{fig:typing-reflection}
 \end{figure}
-
-As with other algebraic effects systems in the literature, we only
-support handlers for theories without any equations, even though our
-semantics is defined in terms of arbitrary bounded Lawvere theories. We
-define algebraic signatures $\Sigma$ as finite sets of operations
-\[
-\{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
-\types \tau_{p_n} \rightarrowtail \tau_{a_n} \}
-\]
-An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
-parameterised by $\tau_p$ and with arity $\tau_a$. Such signatures
-correspond to bounded Lawvere theories without equations.
-
-The language provides inductive types $\ind{\Sigma}$ to directly
-represent terms of an algebraic signature $\Sigma$. $\ind{\Sigma}(\tau)$
-has a constructor $op$ with arguments of type $\tau_p$ and
-$\tau_a \rightarrow \ind{\Sigma}(\tau)$ for each operation
-$op \types \tau_p \rightarrowtail \tau_a$ in $\Sigma$, as well as a
-constructor $ret$ with an argument of type $\tau$. $\ind{\Sigma}$ is
-sometimes called the \emph{freer monad} of $\Sigma$ \cite{??}.
-
-The language provides types $\yoneda{\Sigma}$ to represent morphisms
-from $\sem{\Sigma}$ into the ambient monad. Using these, our desired
-isomorphism appears in the language as two operations, reflection
-$\reflectraw{\Sigma}{op}$ and reification $\reifyraw{\Sigma}$.
-\begin{mathpar}
-\inferrule{\Gamma \turnv v_1 \types \yoneda{\Sigma} \\
-           op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
-           \Gamma \turnv v_2 \types \glob \tau_1 \\
-           \lbind{\Gamma}{x}{\glob \tau_2} \turnc c \types \tau_3}
-{\Gamma \turnc \; \perform{v_1}{op}{v_2}{x}{c} \types \tau_3}
-\and
-\inferrule{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}
-{\Gamma \turnc \; \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)}
-\end{mathpar}
 
@@
 terms $\reflectraw{\Sigma}$ in terms of $\performraw{v}{op}$:
-\[
-    \reflect{\Sigma}{v}{c} := \todo{write this}
-\]
+\begin{align*}
+    \reflect{\Sigma}{v}{c} := \quad &\letv{x_w}{c}{\\
+        &\indelim{\Sigma}{x_w}{ \\
+        &\qquad \indretcase{x_r}{\return{x_r}} \seq \\
+        &\qquad \indcase{op_1}{x_{p_1}}{x_{k_1}}
+          {\perform{v}{op_1}{x_{p_1}}{x_{a_1}}{(\app{x_{k_1}}{x_{a_1}})}} \seq \\
+        &\qquad \ldots \seq \\
+        &\qquad \indcase{op_n}{x_{p_n}}{x_{k_n}}
+          {\perform{v}{op_n}{x_{p_n}}{x_{a_n}}{(\app{x_{k_n}}{x_{a_n}})}}}}
+\end{align*}
 which has this admissible typing rule:
@@
 
-Our denotational semantics in terms of $\pshlawcop$ is
-straightforward. We write $\sem{\Sigma}$ for the Lawvere theory
-corresponding to $\Sigma$, and $F_\Sigma$ for the endofunctor on $\setc$
-for which $T_{\sem{\Sigma}}$ is the free monad. For types, the
-translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
-\begin{mathpar}
-\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}} \and
-\sem{\glob \tau} = \glob \sem{\tau} \and
-\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \and
-\sem{\ind{\Sigma}(\tau)} = \mu X.(\sem{\tau} + F_\Sigma(\mon(X)))
-\end{mathpar}
-where $\mu X. F(X).$ represents the initial algebra of a suitable
-endofunctor. For any Lawvere theory $L$,
-$\mu X.T_L(\sem{\tau} + F_\Sigma(X))$ is isomorphic to to
-$T_{(L + \sem{\Sigma})}$ \cite{hyland2006combining}. That means that
-$\mon(\sem{\ind{\Sigma}(\glob \tau)})$ is isomorphic to the presheaf that
-maps $L$ to $T_{(L + \sem{\Sigma})}(\sem{\tau}(\terminal))$, which in
-turn is isomorphic to $\sem{\yoneda{\Sigma} \rightarrow \glob \tau}$ by the
-result in the previous section. This isomorphism directly provides the
-semantics of reflection and reification.
-
-\begin{align*}
-&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
-&\sem{\glob \Gamma \turnv \gbox{v} \types \glob \tau} =
-  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
-&\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
-  \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
-  \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle
-&\sem{\glob \Gamma \turnv \grun{c} \types \glob \tau} =
-  \epsilon \circ \todo{removing identity monad} \circ
-  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
-\end{align*}
-\begin{align*}
-&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
-&\sem{\Gamma \turnc \return{v} \types \tau} = \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
-&\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} = \mu_{\sem{\tau_2}} \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}) \circ t_{\sem{\Gamma}, \sem{\tau_1}} \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau} \rangle \\
-&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \glob \tau} = \todo{fix}
-  \evmap \langle \reflectc{\Sigma}{\tau} \circ \eta \circ
-                   \sem{\Gamma \turnv v_1 \types \ind{\Sigma}(\glob \tau)},\,
-                 \sem{\Gamma \turnv v_2 \types \yoneda{\Sigma}} \rangle \\
-&\sem{\Gamma \turnc \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)} =
-  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}) \\
-\end{align*}
-where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
-respectively of $\mon$; $\epsilon$ is the counit of $\glob$; $\evmap$
-and $\currymap$ are the evaluation map and currying map of exponentials;
-$\reflectc{\Sigma}{}$ and $\reifyc{\Sigma}{}$ are the natural
-transformations in the isomorphism from the previous section.
-
-\section{Operational semantics}
+We also provide projections $\outl$ and $\outr$ that decompose a mapping
+for $\Sigma_1 + \Sigma_2$ into mappings for $\Sigma_1$ and $\Sigma_2$
+respectively.
+
+\begin{example}
+  The value from Example~\ref{example:negate} can be constructed via reflection as:
+  \begin{align*}
+    &\reify{\Sigma_{bs}}{x_h}{\\
+      &\qquad \perform{x_h}{get}{()}{x_a}{\\
+        &\qquad \qquad \letv{x_n}{\app{\mathit{negate}}{x_a}}{\\
+          &\qquad \qquad \perform{x_h}{set}{x_n}{x_u}{\return{x_u}}}}}
+  \end{align*}
+  where $\mathit{negate}$ is the function that negates a boolean.
+\end{example}
+
+\subsection{Operational semantics}
 \label{sec:operational}
 
+We define the operational semantics as a pair of mutually defined
+reduction relations $\reducestov$ and $\reducestoc$ for values and
+computations respectively shown in Fig.\ref{fig:reduction} The
+definitions of normal forms for values and computations are given in
+Fig.\ref{fig:normal}. Subject reduction and strong normalization of
+these rules can be proved using standard techniques.
+
+\begin{figure}
+\begin{align*}
+\fst{(v_1, v_2)} &\reducestov v_1 \\
+\snd{(v_1, v_2)} &\reducestov v_2 \\
+\gunbox{x}{\gbox{v_1}}{v_2} &\reducestov \sbst{v_2}{x}{v_1} \\
+\grun{\return{v}} &\reducestov v \\
+\letv{x}{\return{v}}{c} &\reducestov \sbst{c}{x}{v} \\
+\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducestov \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
+\app{(\abs{x}{\tau}{c})}{v} &\reducestoc \sbst{c}{x}{v} \\
+\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_1}{x_1}{v} \\
+\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_2}{x_2}{v} \\
+\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} &\reducestoc \sbst{c}{x}{v} \\
+\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
+  \reducestoc \sbst{\sbst{c}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{&\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}} \\
+\reify{\Sigma}{x}{\return{v}} &\reducestoc \return{\ret{\Sigma}{v}} \\
+\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} &\reducestoc \return{\indintro{\Sigma}{op}{v}{\abs{x_2}{\reify{\Sigma}{x_1}{c}}}} \\
+\reify{\Sigma}{x_1}{\perform{x_2}{op}{v}{x_3}{c}} &\reducestoc \perform{x_2}{op}{v}{x_3}{\reify{\Sigma}{x_1}{c}} \qquad (x_1 \neq x_2) \\
+\perform{(\outl{v_1})}{op}{v_2}{x}{c} &\reducestoc \perform{v_1}{\mathrm{left}(op)}{v_2}{x}{c} \\
+\perform{(\outr{v_1})}{op}{v_2}{x}{c} &\reducestoc \perform{v_1}{\mathrm{right}(op)}{v_2}{x}{c}
+\end{align*}
+\caption{Reduction relation}
+\label{fig:reduction}
+\end{figure}
 \begin{figure}
@@
 
-\begin{figure}
-\begin{align*}
-\fst{(v_1, v_2)} &\reducesto v_1 \\
-\snd{(v_1, v_2)} &\reducesto v_2 \\
-\gunbox{x}{\gbox{v_1}}{v_2} &\reducesto \sbst{v_2}{x}{v_1} \\
-\grun{\return{v}} &\reducesto v \\
-\letv{x}{\return{v}}{c} &\reducesto \sbst{c}{x}{v} \\
-\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducesto \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
-\app{(\abs{x}{\tau}{c})}{v} &\reducesto \sbst{c}{x}{v} \\
-\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducesto \sbst{c_1}{x_1}{v} \\
-\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducesto \sbst{c_2}{x_2}{v} \\
-\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} &\reducesto \sbst{c}{x}{v} \\
-\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
-  \reducesto \sbst{\sbst{c}{x_1}{v_2}}{x_2}{\abs{x_a}{\tau_a}{&\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}} \\
-\reify{\Sigma}{x}{\return{v}} &\reducesto \ret{\Sigma}{v} \\
-\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} &\reducesto \indintro{\Sigma}{op}{v}{\abs{x_2}{\reify{\Sigma}{x_1}{c}}} \\
-\reify{\Sigma}{x_1}{\perform{x_2}{op}{v}{x_3}{c}} &\reducesto \perform{x_2}{op}{v}{x_3}{\reify{\Sigma}{x_1}{c}} \qquad (x_1 \neq x_2) \\
-\perform{(\outl{v_1})}{op}{v_2}{x}{c} &\reducesto \perform{v_1}{\mathrm{left}(op)}{v_2}{x}{c} \\
-\perform{(\outr{v_1})}{op}{v_2}{x}{c} &\reducesto \perform{v_1}{\mathrm{right}(op)}{v_2}{x}{c}
-\end{align*}
-\caption{Reduction relation}
-\label{fig:reduction}
-\end{figure}
+Note that the reduction relations do not include $\eta$ expansion. They
+also do not include some conversions you might expect for computations:
+\begin{itemize}
+\item There is no rule to convert
+  $\letv{x_2}{\letv{x_1}{c_1}}{c_2}{c_3}$ to
+  $\letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}}$ or vice versa.
+\item There is no rule to convert $\letv{x}{c}{\return x}$ to $c$.
+\end{itemize}
+These rules are excluded because they are not $\beta$ rules. Instead
+they are better thought of as $\nu$ rules \cite{??}: additional
+equalities between neutral terms that are immediate consequences of
+structural induction. Rather than be included as part of ordinary
+reduction they can be done as an additional standardization pass on
+neutral terms after reduction and $\eta$-expansion have completed. This
+staged approach to computing equality of terms works because $\nu$ rules
+cannot produce additional $\beta$ redexes.
+
+\subsection{Embedding the Simply-Typed Lambda Calculus}
+
+Using the global modality and $\grun{c}$ construct we can embed the
+Simply Typed Lambda Calculus (STLC) into this language. We define an
+embedding $\embed{\_}$ on types, contexts and terms of STLC.
+
+\section{Presheaves for algebraic effects}
+\label{sec:semantics}
+
+The design of effect reflection falls out naturally from a simple
+denotational semantics built on presheaves. The key idea is to index the
+meaning of types by an ambient algebraic monad of effects: rather than
+have the semantics of each type be a set of elements, the semantics of
+each type is a mapping from the ambient monad to a set of elements.
+
+\subsection{Algebraic monads}
+
+In order to index the meaning of our types by the ambient monad we need
+some way of representing that ambient monad. We restrict the ambient
+monad to be an \emph{algebraic monad}: one that can be expressed as the
+free algebra of an algebraic theory. Such monads have nice properties,
+in particular the sum of two algebraic monads always exists and is
+itself an algebraic monad.
+
+\todo{Check ``algebraic monad'' in the literature and informally. Maybe
+  add a footnote about the term.}
+
+Most monads used in practice for programming are algebraic, with the
+notable exception of the continuation monad. The issue with trying to
+represent non-algebraic monads algebraically is essentially one of size:
+if you try to write down the collection of algebirac operations for the
+continuation monad you get something that is too large to be a set. This
+is closely related to the strict positivity restriction on inductive
+type definitions.
+
+We recall the definitions of algebraic signature, terms of a signature,
+algebraic theory and models of a theory.
+\begin{definition}[Algebraic signature]
+  An \emph{algebraic signature} is a set of operations with associated
+  arities.
+\end{definition}
+
+\begin{example}
+  The algebraic signature for boolean state is:
+  \begin{equation*}
+    \{ (\mathrm{get}, 2); (\mathrm{set\_true}, 1); (\mathrm{set\_false}, 1) \}
+  \end{equation*}
+\end{example}
+
+\begin{definition}[Term]
+  A \emph{term} $t$ of some algebraic signature $\Sigma$ is either a
+  free variable $x$ or $op(t_1, \ldots, t_k)$ where $op$ is an operation
+  of $\Sigma$ with arity $k$ and each $t_i$ is itself a term.
+\end{definition}
+
+\begin{definition}[Algebraic theory]
+  An \emph{algebraic theory} consists of an algebraic signature $\Sigma$ and a set of
+  equations on terms of $\Sigma$.
+\end{definition}
+
+\begin{example}
+  The algebraic theory for boolean state adds the following equations to
+  the signature of boolean state:
+  \begin{align*}
+    &\mathrm{set\_true}(\mathrm{get}(x, y)) = \mathrm{set\_true}(x)
+    &\mathrm{set\_false}(\mathrm{get}(x, y)) = \mathrm{set\_false}(y) \\
+    &\mathrm{set\_true}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
+    &\mathrm{set\_true}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
+    &\mathrm{set\_false}(\mathrm{set\_true}(x)) = \mathrm{set\_true}(x)
+    &\mathrm{set\_false}(\mathrm{set\_false}(x)) = \mathrm{set\_false}(x) \\
+    &\mathrm{get}(\mathrm{set\_true}(x), \mathrm{set\_false}(x))) = x
+  \end{align*}
+\end{example}
+
+\begin{definition}[Model]
+  A \emph{model} $M$ of some algebraic theory $L$ consists of a carrier
+  set $S$ and for each operation $op$ in the signature of $L$ a function
+  $op_M : (\Pi_k S) \rightarrow S$ where $k$ is the arity of $op$, such
+  that the equations of $L$ are respected when interpreting the terms
+  using these functions for the operations. Models of $L$ are also
+  called algebras of $L$.
+\end{definition}
+
+\begin{example}
+  We can model the theory of boolean state using the carrier set
+  $2 \rightarrow 2$ and interpreting the operations with:
+  \begin{align*}
+    &\mathrm{get}(x, y) = \lambda s.\, \mif s \mthen x(s) \melse y(s) \\
+    &\mathrm{set\_true}(x) = \lambda s.\, x(\mathrm{true}) \\
+    &\mathrm{set\_false}(x) = \lambda s.\, x(\mathrm{false})
+  \end{align*}
+\end{example}
+
+Arities in algebraic signatures are often considered as natural numbers,
+but they can equivalently be just sets, with families of terms indexed
+by that set used as the arguments of that operator in a term. It is also
+convenient to allow families of operations parameterized by some set. As
+such, we can present a signature in the form:
+\begin{equation*}
+\{ op_1 \types P_1 \rightarrowtail A_1; \ldots; op_n
+\types P_n \rightarrowtail A_n \}
+\end{equation*}
+where $op \types P \rightarrowtail A$ indicates an operation $op$
+parameterised by $P$ and with arity $A$.
+
+\begin{example}
+  The algebraic signature for boolean state can be written as:
+  \begin{equation*}
+    \{ \mathrm{get} \types 1 \rightarrowtail 2; \mathrm{set} \types 2 \rightarrowtail 1 \}
+  \end{equation*}
+\end{example}
+
+Given a Lawvere theory $L$ and a set $A$ we can construct the free
+model, or \emph{free algebra}, $T_L(A)$. The values $T_L(A)$ are terms
+built using the elements of $A$ and the operations of $L$ quotiented by
+the equations in $L$. $T_L$ is a monad and $T$ is one half of an
+equivalence between algebraic theories and algebraic monads.
+
+\begin{example}
+  The free algebra of the theory of boolean state is equivalent
+  to the type:
+  \begin{lstlisting}[style=oxcaml]
+  type 'a t =
+    | Return of 'a
+    | Get of unit * (bool -> 'a t)
+    | Set of bool * (unit -> 'a t)
+  \end{lstlisting}
+  and it forms a monad:
+  \begin{lstlisting}[style=oxcaml]
+  let return x = Return x
+  let rec bind x f =
+    match x with
+    | Return x -> f x
+    | Get((), k) -> Get((), (fun s -> bind (k s) f))
+    | Set(s, k) -> Set(s, (fun () -> bind (k ()) f))
+  \end{lstlisting}  
+\end{example}
+
+Traditionally, algebraic signatures are restricted to the case of a
+finite set of operations each with a finite arity. Most theoretical work
+in the algebraic effects literature extends this to the case of
+countable sets of operations with countable arities, which allows us to
+construct the signature for natural number state:
+
+\begin{example}
+  The algebraic signature for natural number state is:
+  \begin{equation*}
+    \{ \mathrm{get} \types 1 \rightarrowtail \nat;
+       \mathrm{set} \types \nat \rightarrowtail 1 \}
+  \end{equation*}
+\end{example}
+
+However, this extension is not sufficient to construct the signature for
+state containing a function $\nat \rightarrow \nat$. For that we must
+extend algebraic signatures further to allow for arbitrary sets of
+operations with arbitrary sets as arities. Such algebraic theories are
+called \emph{bounded infinitary algebraic theories} because for each
+such theory there is some regular cardinal $\kappa$ that is a bound on
+the arities of the theory. An algebraic theory with some bound $\kappa$
+corresponds to a monad with rank $\kappa$. Thus we take monads with a
+rank as our precise definition of algebraic monad.
+
+Note that we still require that our collections of operations and their
+arities to be sets, so our definition of algebraic monad continues to
+exclude monads like the continuation monad. Most importantly, this
+definition preserves the nice properties of finitary monads required to
+support effect handlers.
+
+\subsection{(Bounded) Lawvere theories}
+
+We use \emph{Lawvere theories} as our concrete representation of the
+ambient algebraic monad. Lawvere theories are a categorical
+representation of algebraic theories that doesn't depend on the
+particular choice of operations and equations.
+
+\begin{definition}
+  A \emph{Lawvere theory} is a category $L$ with finite products in
+  which every object is isomorphic to some finite cartesion product
+  $x^n = x \times \ldots \times x$ of a distinguished object $x$.
+\end{definition}
+The objects of a Lawvere theory correspond to the different possible
+arities and the morphisms $x^n \rightarrow x$ correspond to terms
+of $L$ with $n$ free variables quotiented by the equations in $L$. A
+model of a Lawvere theory $L$ is a product-preserving functor
+$L \rightarrow \setc$.
+
+Lawvere theories correspond to the traditional for of algebraic
+theories: those with a finite set of operations with finite arities, but
+we can extend the definition to get something that corresponds to
+\emph{bounded infinitary algebraic theories}.
+\begin{definition}
+  A \emph{Bounded Lawvere theory} is a small category $L$ with small
+  products in which every object is isomorphic to some small cartesian
+  product $\Pi_S x$ of a distinguished object $x$.
+\end{definition}
+Note that the ``bounded'' aspect of the definition comes from the
+restriction that the category be small.
+
+We will mostly describe Lawvere theories and constructions on them in
+terms of their presentations as algebraic theories, so an understanding
+the details of Lawvere theories is not required for understanding the
+rest of this paper.
+
+Bounded Lawvere theories form a category $\lawc$ whose morphisms are
+equivalent to maps from the operations of one theory to the operations
+of another that preserve the equations of the first theory according to
+the equations of the second.
+
+$\lawc$ has an initial object $\initial$, which is the algebraic theory
+with no operations. It also has coproducts $L + M$ which is the theory
+made by unioning the operations and equations of $L$ and $M$.
+
+\subsection{Presheaves}
+
+For a category $\cat$, $\psh = \setc^{\catop}$ is the category of
+presheaves over $\cat$. It's objects are contravariant functors from
+$\cat$ to $\setc$ and it's morphisms are natural transformations between
+such functors.
+
+A common use of presheaves is where $\cat$ has finite products and is
+thought of as some form of context over which everything is
+parameterized. In our case we wish to be parameterized by a context of
+\emph{algebraic theories that can be mapped into the ambient
+  monad}. Note that the algebraic theories in such a context are
+\emph{contravariant}: they describe theories that can be handled by the
+ambient monad. As such, we use
+\begin{equation*}
+  \pshlawcop = \setc^{\lawc}
+\end{equation*}
+as the basis of our semantics. The $\op$ in the definition of presheaves
+and the $\op$ due to our dependence being contravariant cancel out, but
+it is still best to think in terms of presheaves as that is a more
+precise description of what we are doing.
+
+Our semantics will use objects from $\pshlawcop$ to represent types and
+contexts, and morphisms in $\pshlawcop$ to represent terms. This allows
+their meanings to be parameterized by the ambient monad. By using
+presheaves rather than ordinary functions we ensure terms and types at a
+particular ambient monad can always be mapped naturally to a larger
+ambient monad.
+
+Categories of presheaves have products and sums, which are defined
+pointwise. We will use these as the semantics of product types and sum
+types in our denotational semantics. Such categories also have
+exponentials, which we will use as the semantics of functions types.
+
+\todo{fix this..}
+
+Our denotational semantics in terms of $\pshlawcop$ is
+straightforward. For types, the translation
+$\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
+\begin{mathpar}
+
+\end{mathpar}
+
+\begin{align*}
+&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
+\end{align*}
+\begin{align*}
+&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}\end{align*}
+
+\subsection{Representing computations: The free algebra monad}
+
+Our aim is to have a semantics where the meaning of types is indexed by
+the ambient monad -- or more precisely by the algebraic theory whose
+free algebra is the ambient monad. In particular, we would like the
+interpretation of types of computations to be applications of the
+ambient monad.
+
+Given a Lawvere theory $L$, the underlying set of its free algebra forms
+a monad $T_L$ on $\setc$.  $T_{\_}$ is functorial and we can use it to
+construct a monad $\mon$ on $\pshlawcop$ such that:
+\begin{equation*}
+\mon(P)(L) = T_L(P(L))
+\end{equation*}
+We call $\mon$ the \emph{free algebra monad}, and we will use it as the
+computation monad in our denotational semantics.
+
+\todo{fix this..}
+
+Our denotational semantics in terms of $\pshlawcop$ is
+straightforward. For types, the
+translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
+\begin{mathpar}
+\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \and
+\end{mathpar}
+
+\begin{align*}
+&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
+\end{align*}
+\begin{align*}
+&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
+&\sem{\Gamma \turnc \return{v} \types \tau} = \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
+&\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} = \mu_{\sem{\tau_2}} \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}) \circ t_{\sem{\Gamma}, \sem{\tau_1}} \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau} \rangle \\
+\end{align*}
+where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
+respectively of $\mon$ $\evmap$
+and $\currymap$ are the evaluation map and currying map of exponentials.
+
+\subsection{Reflecting operations: The Yoneda embedding}
+
+We want to be able to reflect operations into the ambient monad. We can
+do that using the \emph{Yoneda embedding}. The Yoneda embedding
+$\yoneda{\_}$ is a full and faithful embedding from a category $\cat$
+into its category of presheaves $\psh$. When treating $\cat$ as a form
+of context, the elements of $y(\catobj)$ can be used as variables of
+type $\catobj$.
+
+$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding: for each
+Lawvere theory $L$ there is a presheaf $\yoneda{L}$ in $\pshlawcop$ such
+that:
+\begin{align*}
+&\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}} \\
+&\yoneda{L}(l^{\prime}) = \lambda f.\; l^{\prime} \circ f
+\end{align*}
+We will use $\yoneda{L}$ as the semantics for our handler types. Their
+elements are just maps from $L$ into the ambient monad.
+
+For an operation $op \types P \rightarrowtail A$ of a Lawvere theory
+$L$, we can define the following morphism of presheaves:
+
+\todo{Write this}
+
+that reflects the operation into the ambient monad. We use this as our
+semantics of reflection.
+
+Note that the Yoneda embedding preserves finite products, which for our
+case means:
+\begin{equation*}
+\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
+\end{equation*}
+which we use to support \todo{something}
+
+\todo{fix this..}
+
+Our denotational semantics in terms of $\pshlawcop$ is
+straightforward. We write $\sem{\Sigma}$ for the Lawvere theory
+corresponding to $\Sigma$, and $F_\Sigma$ for the endofunctor on $\setc$
+for which $T_{\sem{\Sigma}}$ is the free monad. For types, the
+translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
+\begin{mathpar}
+\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}} \and
+\end{mathpar}
+
+\begin{align*}
+&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
+\end{align*}
+\begin{align*}
+&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
+&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \glob \tau} = \todo{fix}
+  \evmap \langle \reflectc{\Sigma}{\tau} \circ \eta \circ
+                   \sem{\Gamma \turnv v_1 \types \ind{\Sigma}(\glob \tau)},\,
+                 \sem{\Gamma \turnv v_2 \types \yoneda{\Sigma}} \rangle \\
+\end{align*}
+
+\subsection{The global modality}
+
+For any category $\cat$, if $\cat$ has a terminal object $\terminal$, we
+can define the global modality $\glob \types [\psh, \psh]$ such that:
+$\glob(P)(L) = P(\terminal)$. In the case of $\pshlawcop$ this becomes:
+\begin{equation*}
+\glob(P)(L) = P(\initial)
+\end{equation*}
+
+Within our semantics, this amounts to restricting a type to only the
+elements which do not depend on the ambient monad. This is very similar
+to the purity comonad \todo{check name} in ?? \cite{??}. In particular,
+\begin{equation*}
+  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob
+  \cong \quad \glob P
+\end{equation*}
+as elements of the free algebra of the initial algebraic theory are just
+elements of the carrier set.
+
+For any non-empty Lawvere theory $L$, $\glob \yoneda{L}$ is the constant
+presheave that always returns the empty set because there are no
+morphisms from $L$ into $\initial$. In our denotational semantics this
+amounts to all non-trivial global handler types being uninhabited.
+
+\todo{fix this..}
+
+Our denotational semantics in terms of $\pshlawcop$ is
+straightforward. For types, the
+translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
+\begin{mathpar}
+\sem{\glob \tau} = \glob \sem{\tau} \and
+\end{mathpar}
+\begin{align*}
+&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
+&\sem{\glob \Gamma \turnv \gbox{v} \types \glob \tau} =
+  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
+&\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
+  \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
+  \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle
+&\sem{\glob \Gamma \turnv \grun{c} \types \glob \tau} =
+  \epsilon \circ \todo{removing identity monad} \circ
+  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
+\end{align*}
+where $\epsilon$ is the counit of $\glob$.
+
+\subsection{Representing effectful computations: The Yoneda lemma}
+
+Via the Yoneda lemma, we get the following:
+\begin{equation*}
+(\arrow{\yoneda{L}}{\mon(\glob A)})(L') \quad \cong \quad T_{(L' + L)}(A(\terminal))
+\end{equation*}
+which will form the basis of our language's isomorphisms between arrows
+$Y \rightarrow \glob A$ and algebraic monads $M(\glob A)$.
+
+Exponentials in categories of presheaves are defined such that:
+\begin{equation*}
+\arrow{P}{Q}(L) = \homset{\pshlawcop}{\yoneda{L} \times P}{Q}
+\end{equation*}
+
+\subsection{Inductive types and effect sums}
+
+For any Lawvere theory $L$,
+$\mu X.T_L(\sem{\tau} + F_\Sigma(X))$ is isomorphic to to
+$T_{(L + \sem{\Sigma})}$ \cite{hyland2006combining}. That means that
+$\mon(\sem{\ind{\Sigma}(\glob \tau)})$ is isomorphic to the presheaf
+that maps $L$ to $T_{(L + \sem{\Sigma})}(\sem{\tau}(\terminal))$, which
+in turn is isomorphic to $\sem{\yoneda{\Sigma} \rightarrow \glob \tau}$
+by the result above. This isomorphism directly provides the semantics of
+reification.
+
+\todo{fix this..}
+
+Our denotational semantics in terms of $\pshlawcop$ is
+straightforward. We write $F_\Sigma$ for the endofunctor on $\setc$
+for which $T_{\sem{\Sigma}}$ is the free monad. For types, the
+translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
+\begin{mathpar}
+\sem{\ind{\Sigma}(\tau)} = \mu X.(\sem{\tau} + F_\Sigma(\mon(X)))
+\end{mathpar}
+where $\mu X. F(X).$ represents the initial algebra of a suitable
+endofunctor.
+
+\begin{align*}
+&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
+\end{align*}
+\begin{align*}
+&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
+\end{align*}
+
+\subsection{Effect reflection}
+
+\todo{fix this..}
+
+Our denotational semantics in terms of $\pshlawcop$ is
+straightforward.
+\begin{align*}
+&\sem{\Gamma \turnc \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)} =
+  \reifyc{\Sigma}{\tau} \circ \currymap(\sem{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}) \\
+\end{align*}
+$\reifyc{\Sigma}{}$ is the natural transformations in the isomorphism
+from the previous section.
 
@@
 generate a fresh effect constructor for each reification, pass that
-constructor around as part of the \lstinline[style=ocaml]{handler}, and
+constructor around as part of the \lstinline[style=oxcaml]{handler}, and
 then use it to perform operations during reflection.
 
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Handler (O : Op) = struct
@@
 Sums are then implemented by just lifting the injections on operations:
-\begin{lstlisting}[style=ocaml]
+\begin{lstlisting}[style=oxcaml]
 module Sum (L : Op) (R : Op) = struct
@@
 \label{sec:related-work}
+
+\subsection{Lexical and named effect handlers}
+
+\subsection{Monadic reflection}
+
+\subsection{Semantics of algebraic effects}
+
+\subsection{Comonadic purity}
+
+\subsection{Locality and presheaves}
 

### 2025-07-08 paper.tex
More writing

@@
 \newcommand{\homset}[3]{\mathrm{Hom}_{#1}(#2, #3)}
-\newcommand{\arrow}[2]{#1 \Rightarrow #2}
+\newcommand{\arrow}[2]{{#2}^{#1}}
 \newcommand{\sem}[1]{\llbracket #1 \rrbracket}
@@
 \newcommand{\snd}{\mathop{\pi_2}}
+\newcommand{\il}{\mathop{\iota_1}}
+\newcommand{\ir}{\mathop{\iota_2}}
 \newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
@@
 \newcommand{\melse}{\mathop{\mathrm{else}}}
-\newcommand{\embed}[1]{\lbag #1 \rbag}
+\newcommand{\embed}[1]{\mathopen{\lbag}#1\mathclose{\rbag}}
 
@@
 \end{align*}
-\caption{Reduction relation}
+\caption{Reduction relations}
 \label{fig:reduction}
@@
 \item There is no rule to convert
-  $\letv{x_2}{\letv{x_1}{c_1}}{c_2}{c_3}$ to
-  $\letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}}$ or vice versa.
-\item There is no rule to convert $\letv{x}{c}{\return x}$ to $c$.
+  $\; \letv{x_2}{\letv{x_1}{c_1}}{c_2}{c_3} \;$ to
+  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \;$ or vice versa.
+\item There is no rule to convert $\; \letv{x}{c}{\return x} \;$ to $c$.
 \end{itemize}
@@
 Using the global modality and $\grun{c}$ construct we can embed the
-Simply Typed Lambda Calculus (STLC) into this language. We define an
-embedding $\embed{\_}$ on types, contexts and terms of STLC.
+Simply Typed Lambda Calculus (STLC) into this language as values. We
+define an embedding $\embed{\_}$ on types, contexts and terms of STLC:
+\begin{equation*}
+\begin{aligned}[c]
+  &\embed{1} = 1 \\
+  &\embed{A \times B} = \embed{A} \times \embed{B} \\
+  &\embed{A \rightarrow B} = (\glob \embed{A}) \rightarrow \embed{B} \\ \\
+  &\embed{\cempty} = \cempty \\
+  &\embed{\lbind{\Gamma}{x}{A}} = \gbind{\embed{\Gamma}}{x}{\embed{A}}
+\end{aligned}
+\qquad
+\begin{aligned}[c]
+  &\embed{x} = x \\
+  &\embed{(e, f)} = (\embed{e}, \embed{f}) \\
+  &\embed{\fst{e}} = \fst{\embed{e}} \\
+  &\embed{\snd{e}} = \snd{\embed{e}} \\
+  &\embed{\abs{x}{A}{e}} =
+    \abs{x^{\prime}}{\glob \embed{A}}{\return{(\gunbox{x}{x^{\prime}}{\embed{e}})}} \\
+  &\embed{\app{f}{e}} = \grun{\app{\embed{f}}{(\gbox{\embed{e}})}}
+\end{aligned}
+\end{equation*}
+
+Note that $\reducestov$ matches the behaviour of $\beta$ reduction of
+STLC too, with each STLC reduction step corresponding to up to 3 steps
+of $\reducestov$.
 
@@
 In order to index the meaning of our types by the ambient monad we need
-some way of representing that ambient monad. We restrict the ambient
-monad to be an \emph{algebraic monad}: one that can be expressed as the
-free algebra of an algebraic theory. Such monads have nice properties,
-in particular the sum of two algebraic monads always exists and is
-itself an algebraic monad.
-
-\todo{Check ``algebraic monad'' in the literature and informally. Maybe
-  add a footnote about the term.}
-
-Most monads used in practice for programming are algebraic, with the
-notable exception of the continuation monad. The issue with trying to
-represent non-algebraic monads algebraically is essentially one of size:
-if you try to write down the collection of algebirac operations for the
-continuation monad you get something that is too large to be a set. This
-is closely related to the strict positivity restriction on inductive
-type definitions.
+some way of representing that ambient monad. As with all approaches to
+algebraic effects, we must restrict ourselves to monads that have some
+nice properties, in particular we need the sum of two of these monads to
+always exists.
+
+To this end, we will restrict the ambient monad to be an \emph{algebraic
+  monad}: one that can be expressed as the free algebra of an algebraic
+theory. Note that this defintion of algebraic monad is not precise:
+different choices for the definition of ``algebraic theory'' give you
+different meanings for ``algebraic monad''. We'll make our particular
+choice over the course of this section.
 
@@
 
-Note that we still require that our collections of operations and their
-arities to be sets, so our definition of algebraic monad continues to
-exclude monads like the continuation monad. Most importantly, this
-definition preserves the nice properties of finitary monads required to
-support effect handlers.
+Most monads used in practice for programming are algebraic by this
+definition, with the notable exception of the continuation monad. The
+issue with trying to represent non-algebraic monads algebraically is
+essentially one of size: if you try to write down the collection of
+algebirac operations for the continuation monad you get something that
+is too large to be a set. This is closely related to the strict
+positivity restriction on inductive type definitions.
 
@@
 Our semantics will use objects from $\pshlawcop$ to represent types and
-contexts, and morphisms in $\pshlawcop$ to represent terms. This allows
+contexts, and morphisms in $\pshlawcop$ to represent values. This allows
 their meanings to be parameterized by the ambient monad. By using
@@
 
-Categories of presheaves have products and sums, which are defined
+\begin{equation*}
+\sem{\tau} \mathrel{:} \pshlawcop \qquad
+\sem{\Gamma} \mathrel{:} \pshlawcop \qquad
+\sem{\Gamma \turnv v \types \tau} \mathrel{:} \homset{\pshlawcop}{\sem{\Gamma}}{\sem{\tau}}
+\end{equation*}
+
+Categories of presheaves have products and coproducts, which are defined
 pointwise. We will use these as the semantics of product types and sum
-types in our denotational semantics. Such categories also have
-exponentials, which we will use as the semantics of functions types.
-
-\todo{fix this..}
-
-Our denotational semantics in terms of $\pshlawcop$ is
-straightforward. For types, the translation
-$\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
-\begin{mathpar}
-
-\end{mathpar}
-
-\begin{align*}
-&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
-\end{align*}
-\begin{align*}
-&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}\end{align*}
+types.
+\begin{align*}
+&\sem{1} = 1 \\
+&\sem{\tau_1 \times \tau_2} = \sem{\tau_1} \times \sem{\tau_2} \\
+&\sem{\tau_1 + \tau_2} = \sem{\tau_1} + \sem{\tau_2} \\
+\end{align*}
+The semantics of the corresponding introduction and elimination forms
+follows the usual categorical semantics of such types.
+
+We also use products for the semantics of contexts:
+\begin{align*}
+&\sem{\cempty} = 1 \\
+&\sem{\lbind{\Gamma}{x}{\tau}} = \sem{\Gamma} \times \sem{\tau} \\
+\end{align*}
 
@@
 Given a Lawvere theory $L$, the underlying set of its free algebra forms
-a monad $T_L$ on $\setc$.  $T_{\_}$ is functorial and we can use it to
-construct a monad $\mon$ on $\pshlawcop$ such that:
+a monad $T_L$ on $\setc$. In addition to the unit of the monad, we can
+construct elements of $T_L$ using the operations of $L$. For an
+operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$ and
+some set $S$, there is a function
+$op_{T_L} \types (P \times (A \rightarrow T_L(S))) \rightarrow T_L(S)$
+that constructs terms in the free algebra using $op$.
+
+$T_{\_}$ is functorial and we can use it to construct a monad $\mon$ on
+$\pshlawcop$ such that:
 \begin{equation*}
@@
 computation monad in our denotational semantics.
-
-\todo{fix this..}
-
-Our denotational semantics in terms of $\pshlawcop$ is
-straightforward. For types, the
-translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
-\begin{mathpar}
-\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \and
-\end{mathpar}
-
-\begin{align*}
-&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
-\end{align*}
-\begin{align*}
-&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
-&\sem{\Gamma \turnc \return{v} \types \tau} = \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
-&\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} = \mu_{\sem{\tau_2}} \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2}) \circ t_{\sem{\Gamma}, \sem{\tau_1}} \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau} \rangle \\
+\begin{align*}
+  &\sem{\Gamma \turnc v \types \tau} \mathrel{:}
+    \homset{\pshlawcop}{\sem{\Gamma}}{\mon(\sem{\tau})} \\
+  &\sem{\Gamma \turnc \return{v} \types \tau} =
+    \eta_{\sem{\tau}} \circ \sem{\Gamma \turnv v \types \tau} \\
+  &\sem{\Gamma \turnc \letv{x}{c_1}{c_2} \types \tau_2} =
+    \mu_{\sem{\tau_2}}
+    \circ \mon(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c_2 \types \tau_2})
+    \circ t_{\sem{\Gamma}, \sem{\tau_1}}
+    \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnc c_1 \types \tau_1} \rangle
 \end{align*}
 where $\eta$, $\mu$ and $t$ are the unit, multiplication and strength
-respectively of $\mon$ $\evmap$
-and $\currymap$ are the evaluation map and currying map of exponentials.
-
-\subsection{Reflecting operations: The Yoneda embedding}
-
-We want to be able to reflect operations into the ambient monad. We can
-do that using the \emph{Yoneda embedding}. The Yoneda embedding
-$\yoneda{\_}$ is a full and faithful embedding from a category $\cat$
-into its category of presheaves $\psh$. When treating $\cat$ as a form
-of context, the elements of $y(\catobj)$ can be used as variables of
-type $\catobj$.
-
-$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding: for each
-Lawvere theory $L$ there is a presheaf $\yoneda{L}$ in $\pshlawcop$ such
-that:
-\begin{align*}
-&\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}} \\
-&\yoneda{L}(l^{\prime}) = \lambda f.\; l^{\prime} \circ f
-\end{align*}
-We will use $\yoneda{L}$ as the semantics for our handler types. Their
-elements are just maps from $L$ into the ambient monad.
-
-For an operation $op \types P \rightarrowtail A$ of a Lawvere theory
-$L$, we can define the following morphism of presheaves:
-
-\todo{Write this}
-
-that reflects the operation into the ambient monad. We use this as our
-semantics of reflection.
-
-Note that the Yoneda embedding preserves finite products, which for our
-case means:
-\begin{equation*}
-\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
-\end{equation*}
-which we use to support \todo{something}
-
-\todo{fix this..}
-
-Our denotational semantics in terms of $\pshlawcop$ is
-straightforward. We write $\sem{\Sigma}$ for the Lawvere theory
-corresponding to $\Sigma$, and $F_\Sigma$ for the endofunctor on $\setc$
-for which $T_{\sem{\Sigma}}$ is the free monad. For types, the
-translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
-\begin{mathpar}
-\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}} \and
-\end{mathpar}
-
-\begin{align*}
-&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
-\end{align*}
-\begin{align*}
-&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
-&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \glob \tau} = \todo{fix}
-  \evmap \langle \reflectc{\Sigma}{\tau} \circ \eta \circ
-                   \sem{\Gamma \turnv v_1 \types \ind{\Sigma}(\glob \tau)},\,
-                 \sem{\Gamma \turnv v_2 \types \yoneda{\Sigma}} \rangle \\
-\end{align*}
+respectively of $\mon$.
+
+Categories of presheaves also have exponentials, which combine with the
+free algebra monad to give us our semantics for function types.
+\begin{align*}
+  &\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \\
+  &\sem{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2} =
+    \currymap(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2})
+     \\
+  &\sem{\Gamma \turnc \app{v_1}{v_2} \types \tau_2} =
+    \evmap
+    \circ \langle \sem{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2},
+                  \sem{\Gamma \turnv v_2 \types \tau_1} \rangle
+\end{align*}
+where $\evmap$ is the evaluation map and $\currymap$ the currying of
+exponentials.
 
@@
 where $\epsilon$ is the counit of $\glob$.
+
+\subsection{Reflecting operations: The Yoneda embedding}
+
+We want to be able to reflect operations into the ambient monad. We can
+use the $op_{T_L}$ functions to build morphisms into $\mon$, but we
+require some way to represent operations of the ambient monad. We can do
+that by using the \emph{Yoneda embedding}. The Yoneda embedding
+$\yoneda{\_}$ is a full and faithful embedding from a category $\cat$
+into its category of presheaves $\psh$. When treating $\cat$ as a form
+of context, the elements of $y(\catobj)$ can be used to represent
+variables of type $\catobj$.
+
+$\lawcop$ embeds into $\pshlawcop$ via the Yoneda embedding: for each
+Lawvere theory $L$ there is a presheaf $\yoneda{L}$ in $\pshlawcop$ such
+that:
+\begin{align*}
+&\yoneda{L}(L^{\prime}) = \homset{\lawc}{L}{L^{\prime}} \\
+&\yoneda{L}(l^{\prime}) = \lambda f.\; l^{\prime} \circ f
+\end{align*}
+
+For an operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$
+and some set $S$, we can build a morphism:
+\begin{equation*}
+op_s \types \homset{\pshlawcop}{\yoneda{L}}{\mon{S}}
+\end{equation*}
+
+We use $\yoneda{L}$ as the semantics for our handler types.
+\begin{align*}
+&\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}}\\
+\end{align*}
+where $\sem{\Sigma}$ is the Lawvere theory corresponding to the
+signature $\Sigma$. If $\Sigma$ is:
+\begin{equation*}
+\{ \, op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
+\types \tau_{p_n} \rightarrowtail \tau_{a_n} \, \}
+\end{equation*}
+then $\sem{\Sigma}$ is the Lawvere theory for:
+\begin{equation*}
+\{ \, op_1 \types \sem{\tau_{p_1}}(\initial) \rightarrowtail \sem{\tau_{a_1}}(\initial); \ldots; op_n
+\types \sem{\tau_{p_n}}(\initial) \rightarrowtail \sem{\tau_{a_n}}(\initial) \, \}
+\end{equation*}
+Note that by applying the types to the initial theory $\initial$ we
+restrict them to their global elements: those that do not depend on the
+current ambient monad.
+
+For an operation $op \types P \rightarrowtail A$ of a Lawvere theory
+$L$, we can define the following morphism of presheaves:
+
+\todo{Write this}
+
+that reflects the operation into the ambient monad. We use this as our
+semantics of reflection.
+
+\todo{fix this..}
+
+\begin{align*}
+&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \glob \tau} = \todo{fix}
+  \evmap \langle \reflectc{\Sigma}{\tau} \circ \eta \circ
+                   \sem{\Gamma \turnv v_1 \types \ind{\Sigma}(\glob \tau)},\,
+                 \sem{\Gamma \turnv v_2 \types \yoneda{\Sigma}} \rangle \\
+\end{align*}
+
+Note that the Yoneda embedding preserves finite products, which for our
+case means:
+\begin{equation*}
+\yoneda{L + L^{\prime}} \cong \yoneda{L} \times \yoneda{L^{\prime}}
+\end{equation*}
+which we use for the semantics of $\outl(v)$ and $\outr{v}$.
 

### 2025-07-08 paper.tex
The writing continues

@@
 \newcommand{\reifyc}[2]{\mathop{\reifyname_{#1 #2}}}
-\newcommand{\outl}[1]{\mathop{\mathrm{out}_L} #1}
-\newcommand{\outr}[1]{\mathop{\mathrm{out}_R} #1}
+\newcommand{\outlname}{\mathrm{out}_L}
+\newcommand{\outl}[1]{\mathop{\outlname} #1}
+\newcommand{\outrname}{\mathrm{out}_R}
+\newcommand{\outr}[1]{\mathop{\outrname} #1}
 \newcommand{\mon}{\mathcal{T}}
@@
 \newcommand{\snd}{\mathop{\pi_2}}
-\newcommand{\il}{\mathop{\iota_1}}
-\newcommand{\ir}{\mathop{\iota_2}}
 \newcommand{\inl}[1]{\mathop{\mathrm{in}_L} #1}
@@
 \item The ambient monad can be defined as a monad $\mon$ on such
-  presheaves, which maps each algebraic theory to its free algebra. We use
+  presheaves, which maps each algebraic theory to its free algebras. We use
   $\mon$ to define the semantics of computations.
@@
 
-Terms in the free algebra of the sum of two effects
+Terms in the free algebras of the sum of two effects
 (i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
@@
 To this end, we will restrict the ambient monad to be an \emph{algebraic
-  monad}: one that can be expressed as the free algebra of an algebraic
+  monad}: one that can be expressed as the free algebras of an algebraic
 theory. Note that this defintion of algebraic monad is not precise:
@@
 \begin{example}
-  The free algebra of the theory of boolean state is equivalent
+  The free algebras of the theory of boolean state are equivalent
   to the type:
@@
   \end{lstlisting}
-  and it forms a monad:
+  and they form a monad:
   \begin{lstlisting}[style=oxcaml]
@@
 
-\subsection{Representing computations: The free algebra monad}
+\subsection{Representing computations: The free algebras monad}
 
@@
 the ambient monad -- or more precisely by the algebraic theory whose
-free algebra is the ambient monad. In particular, we would like the
+free algebras are the ambient monad. In particular, we would like the
 interpretation of types of computations to be applications of the
@@
 
-Given a Lawvere theory $L$, the underlying set of its free algebra forms
+Given a Lawvere theory $L$, the underlying set of its free algebras form
 a monad $T_L$ on $\setc$. In addition to the unit of the monad, we can
@@
 \end{equation*}
-We call $\mon$ the \emph{free algebra monad}, and we will use it as the
+We call $\mon$ the \emph{free algebras monad}, and we will use it as the
 computation monad in our denotational semantics.
@@
 Categories of presheaves also have exponentials, which combine with the
-free algebra monad to give us our semantics for function types.
+free algebras monad to give us our semantics for function types.
 \begin{align*}
@@
 \end{equation*}
+$\glob$ forms a comonad on $\pshlawcop$. $\glob$ also distributes over
+products: $\glob (P \times Q) = (\glob P) \times (\glob Q)$
 
@@
 elements which do not depend on the ambient monad. This is very similar
-to the purity comonad \todo{check name} in ?? \cite{??}. In particular,
-\begin{equation*}
-  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob
-  \cong \quad \glob P
-\end{equation*}
-as elements of the free algebra of the initial algebraic theory are just
-elements of the carrier set.
-
-For any non-empty Lawvere theory $L$, $\glob \yoneda{L}$ is the constant
-presheave that always returns the empty set because there are no
-morphisms from $L$ into $\initial$. In our denotational semantics this
-amounts to all non-trivial global handler types being uninhabited.
-
-\todo{fix this..}
-
-Our denotational semantics in terms of $\pshlawcop$ is
-straightforward. For types, the
-translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
-\begin{mathpar}
-\sem{\glob \tau} = \glob \sem{\tau} \and
-\end{mathpar}
-\begin{align*}
-&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
-&\sem{\glob \Gamma \turnv \gbox{v} \types \glob \tau} =
-  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
+to the purity comonad in ??\cite{??}. In particular,
+\begin{equation*}
+  \glob \mon(P) \quad = \quad T_{\initial} \circ \glob P
+  \quad \cong \quad \glob P
+\end{equation*}
+as elements of the free algebras of the initial algebraic theory are
+just elements of their carrier sets.
+
+We use the $\glob$ comonad as our semantics for $\glob$ types and
+$\glob$ bindings.
+\begin{align*}
+&\sem{\glob \tau} = \glob \sem{\tau} \\
+&\sem{\gbind{\Gamma}{x}{\tau}} = \sem{\Gamma} \times \glob \sem{\tau} \\
+&\sem{\gbind{\Gamma}{x}{\tau} \turnv x \types \tau} = \epsilon \circ \snd \\
+&\sem{\Gamma \turnv \gbox{v} \types \glob \tau} =
+  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau}
+  \circ \eta \circ \cdiv{\Gamma} \\
 &\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
   \sem{\gbind{\Gamma}{x}{\tau_1} \turnv v_2 \types \tau_2}
-  \circ \langle id_{\sem{\Gamma}}, \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle
-&\sem{\glob \Gamma \turnv \grun{c} \types \glob \tau} =
-  \epsilon \circ \todo{removing identity monad} \circ
-  \glob \sem{\cdiv{\Gamma} \turnv v \types \tau} \circ \todo{weakening and globalizing}\\
-\end{align*}
-where $\epsilon$ is the counit of $\glob$.
+  \circ \langle id_{\sem{\Gamma}},
+                \sem{\Gamma \turnv v_1 \types \glob \tau_1} \rangle \\
+&\sem{\Gamma \turnv \grun{c} \types \glob \tau} =
+  \rho_{\sem{\tau}}
+  \circ \glob \sem{\cdiv{\Gamma} \turnc c \types \tau}
+  \circ \eta \circ \cdiv{\Gamma}\\
+\end{align*}
+where $\epsilon$ and $\eta$ are the counit and comultiplication of
+$\glob$, $\cdiv{\Gamma} \types \Gamma \rightarrow \cdiv{\Gamma}$ is the
+morphism that projects the global bindings from $\Gamma$, and $\rho_P$
+is the isomorphism from $\glob \mon(P)$ to $\glob P$.
 
@@
 and some set $S$, we can build a morphism:
-\begin{equation*}
-op_s \types \homset{\pshlawcop}{\yoneda{L}}{\mon{S}}
-\end{equation*}
-
-We use $\yoneda{L}$ as the semantics for our handler types.
-\begin{align*}
-&\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}}\\
+\begin{align*}
+&op_{\mon{S}} \types
+  \homset{\pshlawcop}
+         {\yoneda{L} \times \glob P \times \arrow{\glob A}{\mon{S}}}
+         {\mon{S}} \\
+&{(op_{\mon{S}})}_{L^\prime}
+  = \lambda(y, p, k). (y(op))_{T_{L^\prime}}(p, \lambda a. k_{L^\prime}(id_{L^\prime}, a))
+\end{align*}
+which uses an element of $\yoneda{L}$ to map $op$ into the ambient monad
+and then uses that to construct a term in the free algebras monad.
+
+We use $\yoneda{L}$ as the semantics for our handler types, and
+$op_{\mon{S}}$ as the semantics for reflection.
+\begin{align*}
+&\sem{\yoneda{\Sigma}} = \yoneda{\sem{\Sigma}} \\
+&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \tau} = \\
+&\quad op_{\mon{\sem{\tau}}}
+       \circ \langle \sem{\Gamma \turnv v_1 \types \yoneda{\Sigma}},
+                      \sem{\Gamma \turnv v_2 \types \tau_p},
+                      \currymap(\sem{\lbind{\Gamma}{x}{\tau_a} \turnc c \types \tau})
+             \rangle
 \end{align*}
@@
 \begin{equation*}
-\{ \, op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1}; \ldots; op_n
-\types \tau_{p_n} \rightarrowtail \tau_{a_n} \, \}
+  \{ \, op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1} \seq
+  \ldots \seq
+  op_n \types \tau_{p_n} \rightarrowtail \tau_{a_n} \, \}
 \end{equation*}
@@
 \begin{equation*}
-\{ \, op_1 \types \sem{\tau_{p_1}}(\initial) \rightarrowtail \sem{\tau_{a_1}}(\initial); \ldots; op_n
-\types \sem{\tau_{p_n}}(\initial) \rightarrowtail \sem{\tau_{a_n}}(\initial) \, \}
+  \{ \, op_1 \types \sem{\tau_{p_1}}(\initial)
+                    \rightarrowtail \sem{\tau_{a_1}}(\initial) \seq
+  \ldots \seq
+  op_n \types \sem{\tau_{p_n}}(\initial)
+              \rightarrowtail \sem{\tau_{a_n}}(\initial) \, \}
 \end{equation*}
@@
 
-For an operation $op \types P \rightarrowtail A$ of a Lawvere theory
-$L$, we can define the following morphism of presheaves:
-
-\todo{Write this}
-
-that reflects the operation into the ambient monad. We use this as our
-semantics of reflection.
-
-\todo{fix this..}
-
-\begin{align*}
-&\sem{\Gamma \turnc \perform{v_1}{op}{v_2}{x}{c} \types \glob \tau} = \todo{fix}
-  \evmap \langle \reflectc{\Sigma}{\tau} \circ \eta \circ
-                   \sem{\Gamma \turnv v_1 \types \ind{\Sigma}(\glob \tau)},\,
-                 \sem{\Gamma \turnv v_2 \types \yoneda{\Sigma}} \rangle \\
-\end{align*}
-
-Note that the Yoneda embedding preserves finite products, which for our
-case means:
+The Yoneda embedding preserves finite products, which for our case
+means:
 \begin{equation*}
@@
 \end{equation*}
-which we use for the semantics of $\outl(v)$ and $\outr{v}$.
+which we use for the semantics of $\outlname$ and $\outrname$.
 
@@
 
-Via the Yoneda lemma, we get the following:
-\begin{equation*}
-(\arrow{\yoneda{L}}{\mon(\glob A)})(L') \quad \cong \quad T_{(L' + L)}(A(\terminal))
-\end{equation*}
-which will form the basis of our language's isomorphisms between arrows
-$Y \rightarrow \glob A$ and algebraic monads $M(\glob A)$.
+We can represent computations in a specific algebraic monad using the
+Yoneda lemma. The Yoneda lemma says that, for any $P \types \psh$ and
+$\catobj \types \cat$, there is an isomorphism:
+\begin{equation*}
+  \homset{\psh}{\yoneda{\catobj}}{P} \cong P(\catobj)
+\end{equation*}
+
+Applying that to the composition of the free algebras monad and the
+global modality we get:
+\begin{equation*}
+  \homset{\pshlawcop}{\yoneda{L}}{\mon{\glob \tau}} \cong \mon{\glob \tau}(L) = T_L(\tau(\initial))
+\end{equation*}
+
+In terms of our semantics, this means that computations of type
+$\glob \tau$ with a single variable of type $\yoneda{\Sigma}$ are
+isomorphic to values of the free algebra of $\sem{\Sigma}$ at
+$\sem{\tau}(\initial)$.
+
+\subsection{Effect sums and inductive types}
+
+The essence of algebraic effect handlers is taking a computation in
+$T_{L + \Sigma}(S)$ the free algebras of the sum of an algebraic theory
+$L$ and an algebraic signature $\Sigma$, spliting out and handling
+$\Sigma$, leaving a computation in only the algebraic theory
+$T_{L}(S^\prime)$.
+
+The key to this process is the following lemma
+\cite{hyland2006combining}.
+\begin{lemma}
+\label{lemma:sums}
+  For any Lawvere theory $L$ and algebraic signature $\Sigma$
+  \begin{equation*}
+  T_{L + \Sigma}(S) \cong \mu X.T_L(S + F_\Sigma(X))
+  \end{equation*}
+\end{lemma}
+where $\mu X.F(X)$ is the intial algebra of a suitable endofunctor $F$,
+and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$ is
+the free monad.
+
+Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
+user via an inductive type. In particular, given $\Sigma$ and a
+presheave $P$, we can define the following initial algebra:
+\begin{align*}
+&\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
+&\ind{\Sigma}(P) = \mu X. (P + F_\Sigma \circ \mon{X})
+\end{align*}
+and observe the following variation on Lemma \ref{lemma:sums}.
+\begin{equation*}
+T_{L + \Sigma}(P(L)) \cong \mon{(\ind{\Sigma}(P))}(L)
+\end{equation*}
+
+We use these initial algebras as the denotational semantics of
+$\ind{\Sigma}$:
+\begin{mathpar}
+\sem{\ind{\Sigma}(\tau)} = \ind{\sem{\Sigma}}(\sem{\tau})
+\end{mathpar}
+
+\subsection{Effect reification}
 
@@
 \begin{equation*}
-\arrow{P}{Q}(L) = \homset{\pshlawcop}{\yoneda{L} \times P}{Q}
-\end{equation*}
-
-\subsection{Inductive types and effect sums}
-
-For any Lawvere theory $L$,
-$\mu X.T_L(\sem{\tau} + F_\Sigma(X))$ is isomorphic to to
-$T_{(L + \sem{\Sigma})}$ \cite{hyland2006combining}. That means that
-$\mon(\sem{\ind{\Sigma}(\glob \tau)})$ is isomorphic to the presheaf
-that maps $L$ to $T_{(L + \sem{\Sigma})}(\sem{\tau}(\terminal))$, which
-in turn is isomorphic to $\sem{\yoneda{\Sigma} \rightarrow \glob \tau}$
-by the result above. This isomorphism directly provides the semantics of
-reification.
-
-\todo{fix this..}
-
-Our denotational semantics in terms of $\pshlawcop$ is
-straightforward. We write $F_\Sigma$ for the endofunctor on $\setc$
-for which $T_{\sem{\Sigma}}$ is the free monad. For types, the
-translation $\sem{\tau} \mathrel{:} \pshlawcop$ is as follows:
-\begin{mathpar}
-\sem{\ind{\Sigma}(\tau)} = \mu X.(\sem{\tau} + F_\Sigma(\mon(X)))
-\end{mathpar}
-where $\mu X. F(X).$ represents the initial algebra of a suitable
-endofunctor.
-
-\begin{align*}
-&\sem{\Gamma \turnv v \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \sem{\tau} \\
-\end{align*}
-\begin{align*}
-&\sem{\Gamma \turnc c \types \tau} \mathrel{:} \sem{\Gamma} \rightarrow \mon(\sem{\tau}) \\
-\end{align*}
-
-\subsection{Effect reflection}
-
-\todo{fix this..}
-
-Our denotational semantics in terms of $\pshlawcop$ is
-straightforward.
+\arrow{P}{Q}(\catobj) \cong \homset{\psh}{\yoneda{\catobj} \times P}{Q}
+\end{equation*}
+Applying this to $\arrow{\yoneda{L}}{\mon{(\glob P)}}$ and then applying
+the Yoneda Lemma we get:
+\begin{align*}
+\arrow{\yoneda{L}}{\mon{(\glob P)}}(L')
+  &\cong \homset{\pshlawcop}{\yoneda{L'} \times \yoneda{L}}{\mon{(\glob P)}} \\
+  &\cong \homset{\pshlawcop}{\yoneda{L' + L}}{\mon{(\glob P)}} \\
+  &\cong \mon{(\glob P)}(L' + L) \\
+  &\cong T_{(L' + L)}(P(L'))
+\end{align*}
+
+In the case where $L$ is a signature $\Sigma$, we can further apply our
+variation of Lemma \ref{lemma:sums}, and observing that all the steps in
+this proof were functorial, produce this isomorphism:
+\begin{equation*}
+  \arrow{\yoneda{\Sigma}}{\mon{(\glob P)}} \cong \mon{(\ind{\Sigma}(\glob P))}
+\end{equation*}
+which we call $\reifyc{\Sigma}{}$.
+
+$\reifyc{\Sigma}{}$ directly gives us our semantics for reification.
 \begin{align*}
@@
 \end{align*}
-$\reifyc{\Sigma}{}$ is the natural transformations in the isomorphism
-from the previous section.
 
@@
 
-\section{Appendix}
-
-Previous theoretical work on algebraic effects, such as \cite{??},
-has tended to use \emph{countable Lawvere theories}. These correspond to
-monads who rank is \todo{how should I write the countable
-  cardinality}. This still restricts the parameters and arity operations
-to be countable. We instead use \emph{bounded Lawvere
-  theories}, corresponding to monads that have some cardinal as their
-rank. This allows parameters and arities of operations to be arbitrary
-simple types.
-
-Bounded Lawvere theories are not given a consistent name in the
-literature. A number of results that are proved for countable Lawvere
-theories are asserted to be true for the bounded case. As we could not
-find actual proofs for these claims we have included such proofs in
-Appendix \ref{??}. We also include a proof of an interesting property
-of bounded Lawvere theories:
-
-\todo{How to format and label this?}  {\b Bounded Lawvere theories
-  correspond precisely to those (infinitary) Lawvere theories whose
-  term monad can be expressed as a W-type}
-
 \end{document}

### 2025-07-09 paper.tex
Write additional things

@@
   direct style. This provides all the benefits of typed algebraic
-  effects and handlers without requiring an effect system. We
-  demonstrate effect reflection by implementing it as a library in
-  OxCaml\cite{lorenzen2024oxidizing,} and working through some classic
+  effects and handlers without requiring an effect system or other
+  specific additions to the type system. We demonstrate effect
+  reflection by implementing it as a library in
+  OxCaml\cite{lorenzen2024oxidizing} and working through some classic
   examples of programming with effects.
@@
 
-Note that, in OCaml, nothing prevents attempting to perform an algebraic
-effect operation that is not supported by the current ambient monad,
-resulting in an \lstinline[style=oxcaml]{Unhandled} exception being
-raised. Other languages, e.g. Koka\cite{??}, use an effect system to
-statically ensure operations are only performed when they are available.
+Note that, in OCaml, the effects are \emph{untracked}: nothing prevents
+attempting to perform an algebraic effect operation that is not
+supported by the current ambient monad, resulting in an
+\lstinline[style=oxcaml]{Unhandled} exception being raised.
+
+Other languages track effects to ensure that operations are only
+performed when they are supported. However, these require extending the
+type system with explicit mechanisms for tracking effects, as in
+Koka\cite{leijen2017type}, or for tracking a general notion of
+capability, as in Effekt\cite{brachthauser2022effects}.
+
+The ``Lightweight effect polymorphism'' approach of Brachthauser et
+al. elegantly manages to track effects without requiring explicit
+effect-tracking machinery in the type system. However, it is restricted
+to second-class functions. We would like to extend this approach to
+first-class functions without having to fall back to explicit
+effect-tracking machinery.
 
@@
 Treating the ambient monad as a notion of location allows a language to
-provide an isomorphism:
-\begin{equation*}
-  \mathop{M}(\glob X) \qquad \cong \qquad Y \mathbin{\rightarrow} \glob X
-\end{equation*}
-where $M$ is any algebraic monad, $Y$ is a specific type related to that
-monad, and $\glob$ is a comonadic global modality. We call this
-isomorphism \emph{effect reflection}.
-
-\todo{Restrict the above claim to signatures}
+provide, for any algebraic signature $\Sigma$, an isomorphism:
+\begin{equation*}
+  \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
+\end{equation*}
+where $\ind{\Sigma}$ is an inductive type representing the free
+algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
+to inject terms of $\Sigma$ into the current ambient
+monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
+$\Sigma$. We call this isomorphism \emph{effect reflection}.
 
 The ability to use the ordinary arrow type
-$Y \mathbin{\rightarrow} \glob X$ in place of the monad
-$\mathop{M}(\glob X)$ allows us to conveniently write effectful
-code in direct style, just as algebraic effects do.
+$\yoneda{\Sigma} \mathbin{\rightarrow} \glob X$ in place of the monad
+$\ind{\Sigma}(\glob X)$ allows us to conveniently write effectful code,
+using multiple effects, in direct style, just as algebraic effects do.
 
@@
 
-Effect reflection is closely related to \emph{lexical effect handlers} in
-systems such as Effeckt\cite{??}. Lexical effect handlers have a value that
-represents the effect handler, but these values are second class and
-require their own special typing rules.
-
-Effect reflection is also closely related to \emph{named effect
-  handlers}\cite{??}. Named effect handlers use higher-rank polymorphism
-to emulate locality tracking in much the same way as the
-\lstinline[style=haskell]{St} monad in Haskell. Using higher-rank
-polymorphism prevents global type inference, requiring complex inference
-that extends beyond classic Hindly-Milner, and demanding a higher
-annotation burden on the programmer. \todo{Mention the explicit monad
-that named effect handlers presumably need.}
-
-Effect reflection is a variation on monadic reflection\cite{??}.
-\todo{Make this relationship precise.}
+\todo{Add some mention of connection to lexical and named handlers}
 
@@
 \end{itemize}
-
-These presheaf semantics are closely related to presheaf semantics for
-higher-order abstract syntax\cite{??}.
 
@@
 Using \lstinline[style=oxcaml]{perform} we can implement reflection of
-terms as:
+global terms as:
 \begin{lstlisting}[style=oxcaml]
@@
   | Op(op, k) -> reflect (k (perform op h)) h
-
-let rec reflect_local (t @ local) h =
-  match t with
-  | Return x -> x
-  | Op(op, k) -> reflect_local (k (perform op h)) h
-\end{lstlisting}
-with types:
+\end{lstlisting}
+with type:
 \begin{lstlisting}[style=oxcaml]
 val reflect : 'a Term(O).t -> (Handler(O).t @ local -> 'a)
-
-val reflect_local :
-   'a Term(O).t @ local -> (Handler(O).t @ local -> 'a) @ local
-\end{lstlisting}
+\end{lstlisting}
+Reflection of local terms can be defined similarly.
 
@@
 \lstinline[style=oxcaml]{List.iter} to be reused with different ambient
-monads achieves the ``lightweight effect polymorphism'' described by ??
-et al.\cite{??} without requiring a separate notion of second-class
-function.
+monads achieves the ``lightweight effect polymorphism'' described by
+Brachth{\"a}user et al.\cite{brachthauser2020effects} without requiring
+a separate notion of second-class function.
 
@@
 
-\subsection{Effect sums}
+\subsection{Effect sums and effect forwarding}
 
@@
 \lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
-\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}.
-
-Terms in the free algebras of the sum of two effects
+\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}. Terms in
+the free algebras of the sum of two effects
 (i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
 recursive datatype that is built from the terms of one of the effects
-and the operations of the other\cite{??}. We can definet this type as:
+and the operations of the other\cite{hyland2006combining}. We can define
+this type and associated isomorphism as:
 \begin{lstlisting}[style=oxcaml]
@@
 end
-\end{lstlisting}
-and expose the isomorphism between
-\lstinline[style=oxcaml]{Handler(L).t @ local -> Handler(R).t @ local -> 'a}
-and \lstinline[style=oxcaml]{Handler(L).t @ local -> 'a Half_term(L)(R).t}:
-\begin{lstlisting}[style=oxcaml]
+
 module Half_reflection (L : Op) (R : Op) : sig
-  module L_eff = Reflection(L)
-  module R_eff = Reflection(R)
-  module Sum_eff = Reflection(Sum(L)(R))
-  module Proj = Project(L)(R)
-
-  let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
-    let rec loop lh = function
-      | Term(Sum(L)(R)).Return ->
-          Half_term(L)(R).Return v
-      | Term(Sum(L)(R)).Op(Left op, k) ->
-          loop lh (k (L_eff.perform op lh))
-      | Term(Sum(L)(R)).Op(Right op, k) ->
-          Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
-    in
-    loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
-
-  let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
-    match f lh with
-    | Return x -> x
-    | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
+  val reify :
+    (Handler(L).t @ local -> Handler(R).t @ local -> 'a)
+    -> Handler(L).t @ local -> 'a Half_term(L)(R).t
+
+  val reflect :
+    (Handler(L).t @ local -> 'a Half_term(L)(R).t)
+    -> Handler(L).t @ local -> Handler(R).t @ local -> 'a
 end
 \end{lstlisting}
+
+Note that the key difference between
+\lstinline[style=oxcaml]{Half_term(L)(R).t} and
+\lstinline[style=oxcaml]{Term(R).t} is that continuing the computation
+after an operation requires passing in a local
+\lstinline[style=oxcaml]{Handler(L).t} again.
+
+%\begin{lstlisting}[style=oxcaml]
+%module Half_reflection (L : Op) (R : Op) : sig
+%  module L_eff = Reflection(L)
+%  module R_eff = Reflection(R)
+%  module Sum_eff = Reflection(Sum(L)(R))
+%  module Proj = Project(L)(R)
+%
+%  let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
+%    let rec loop lh = function
+%      | Term(Sum(L)(R)).Return ->
+%          Half_term(L)(R).Return v
+%      | Term(Sum(L)(R)).Op(Left op, k) ->
+%          loop lh (k (L_eff.perform op lh))
+%      | Term(Sum(L)(R)).Op(Right op, k) ->
+%          Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
+%    in
+%    loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
+%
+%  let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
+%    match f lh with
+%    | Return x -> x
+%    | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
+%end
+%\end{lstlisting}
 
@@
   | Finished
-  | More of int * (Handler(Await).t @ local -> ints)
+  | More of int * (Handler(Await).t @ local -> aints)
 
@@
     | Half_term(Await)(Gen).Return () -> Finished
-    | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, k ())
+    | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
   in
-  loop (Await_gen_eff.reify f)
+  loop (Await_gen_eff.reify f ah)
 \end{lstlisting}
@@
 reduction relations $\reducestov$ and $\reducestoc$ for values and
-computations respectively shown in Fig.\ref{fig:reduction} The
+computations respectively shown in Fig.\ref{fig:reduction}. The
 definitions of normal forms for values and computations are given in
@@
 \item There is no rule to convert
-  $\; \letv{x_2}{\letv{x_1}{c_1}}{c_2}{c_3} \;$ to
-  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \;$ or vice versa.
-\item There is no rule to convert $\; \letv{x}{c}{\return x} \;$ to $c$.
+  $\; \letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} \,$ to
+  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \,$ or vice versa.
+\item There is no rule to convert $\; \letv{x}{c}{\return x} \,$ to $c$.
 \end{itemize}
@@
 Simply Typed Lambda Calculus (STLC) into this language as values. We
-define an embedding $\embed{\_}$ on types, contexts and terms of STLC:
+define an embedding $\embed{\_}$ on types, contexts and terms of STLC in
+Fig.\ref{fig:stlc}. Note that $\reducestov$ matches the behaviour of
+$\beta$ reduction of STLC too, with each STLC reduction step
+corresponding to up to 3 steps of $\reducestov$.
+\begin{figure}
 \begin{equation*}
@@
 \end{equation*}
-
-Note that $\reducestov$ matches the behaviour of $\beta$ reduction of
-STLC too, with each STLC reduction step corresponding to up to 3 steps
-of $\reducestov$.
+\caption{Embedding of Simply-Typed Lambda Calculus into values}
+\label{fig:stlc}
+\end{figure}
 
@@
 
-\subsection{Effect sums and inductive types}
+\subsection{Effect reification and inductive types}
 
@@
 \end{lemma}
-where $\mu X.F(X)$ is the intial algebra of a suitable endofunctor $F$,
-and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$ is
-the free monad.
+where $\mu X.G(X)$ is the intial algebra of some polynomial endofunctor
+$G$, and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$
+is the free monad.
 
 Effect reification works by exposing $\mu X.T_L(S + F_\Sigma(X))$ to the
-user via an inductive type. In particular, given $\Sigma$ and a
-presheave $P$, we can define the following initial algebra:
+user via an inductive type. Categories of presheaves have initial
+algebras of polynomial endofunctors, i.e. W-types, so we can construct
+the required initial algebra as a presheave. Given an algebraic
+signature $\Sigma$ and a presheave $P$, we define the following family
+of initial algebras:
 \begin{align*}
@@
 and observe the following variation on Lemma \ref{lemma:sums}.
+\begin{lemma}
+\label{lemma:sums2}
+  For any Lawvere theory $L$ and algebraic signature $\Sigma$
   \begin{equation*}
@@
   \end{equation*}
+\end{lemma}
 
@@
 
-\subsection{Effect reification}
-
 Exponentials in categories of presheaves are defined such that:
@@
 \begin{align*}
-\arrow{\yoneda{L}}{\mon{(\glob P)}}(L')
-  &\cong \homset{\pshlawcop}{\yoneda{L'} \times \yoneda{L}}{\mon{(\glob P)}} \\
-  &\cong \homset{\pshlawcop}{\yoneda{L' + L}}{\mon{(\glob P)}} \\
-  &\cong \mon{(\glob P)}(L' + L) \\
-  &\cong T_{(L' + L)}(P(L'))
-\end{align*}
-
-In the case where $L$ is a signature $\Sigma$, we can further apply our
-variation of Lemma \ref{lemma:sums}, and observing that all the steps in
-this proof were functorial, produce this isomorphism:
+\arrow{\yoneda{L}}{\mon{(\glob P)}}(L^{\prime})
+  &\cong \homset{\pshlawcop}{\yoneda{L^{\prime}} \times \yoneda{L}}{\mon{(\glob P)}} \\
+  &\cong \homset{\pshlawcop}{\yoneda{L^{\prime} + L}}{\mon{(\glob P)}} \\
+  &\cong \mon{(\glob P)}(L^{\prime} + L) \\
+  &\cong T_{(L^{\prime} + L)}(P(L^{\prime}))
+\end{align*}
+
+In the case where $L$ is a signature $\Sigma$, we can further apply
+Lemma \ref{lemma:sums2} and, observing that all the steps in this proof
+were natural in $L^{\prime}$, produce this isomorphism:
 \begin{equation*}
@@
 
-$\reifyc{\Sigma}{}$ directly gives us our semantics for reification.
+$\reifyc{\Sigma}{}$ gives us our semantics for reification.
 \begin{align*}
@@
 
-\subsection{Implementation with untyped effect handlers}
+\subsection{Implementation with untracked effect handlers}
 
 We can implement the interface from Section \ref{sec:interface} directly
-in terms of OCaml's untyped effect
+in terms of OCaml's untracked effect
 handlers\cite{Sivaramakrishnan2021retrofitting}. The trick is to
@@
 constructor around as part of the \lstinline[style=oxcaml]{handler}, and
-then use it to perform operations during reflection.
-
+then use it to perform operations during reflection. The full implementation
+is given in Fig.\ref{fig:implementation}.
+\begin{figure}
 \begin{lstlisting}[style=oxcaml]
 module Handler (O : Op) = struct
+
   type t = { perform : 'r. 'r O.t -> 'r }
+
 end
@@
 
-  let reflect t =
-    fun {perform} ->
-      let rec loop = function
-        | Return v -> v
-        | Op(o, f) -> loop (f (perform o))
-      in
-      loop t
-
   let reify_local f = exclave
@@
 
-  let reflect_local t = exclave
-    fun {perform} ->
-      let rec loop = function
-        | Return v -> v
-        | Op(o, f) -> loop (f (perform o))
-      in
-      loop t
+  let perform t =
+   fun h -> h.perform t
 
 end
-\end{lstlisting}
-Sums are then implemented by just lifting the injections on operations:
-\begin{lstlisting}[style=oxcaml]
-module Sum (L : Op) (R : Op) = struct
-
-  module Op = struct
-    type 'a t =
-      | Left : 'a L.t -> 'a t
-      | Right : 'a R.t -> 'a t
+
+module Project (L : Op) (R : Op) = struct
+
+  let outl {perform} =
+    {perform = fun op -> perform (Sum(L)(R).Left op)}
+
+  let outr {perform} =
+    {perform = fun op -> perform (Sum(L)(R).Right op)}
+
 end
-
-  let outl {perform} =
-    {perform = fun op -> perform (Left op)}
-
-  let outr {perform} =
-    {perform = fun op -> perform (Right op)}
-
-end
-\end{lstlisting}
+\end{lstlisting}
+\caption{Implementation}
+\label{fig:implementation}
+\end{figure}
 
@@
 
-In the above implementation, whenever an operation is reflected we
-perform a linear search up the stack of handlers until we find the
-corresponding one. This search is not actually necessary...
+In the implementation in Fig.\ref{fig:implementation}, whenever an
+operation is reflected we perform a linear search up the stack of
+handlers until we find the corresponding one. This search is not
+actually necessary: there is a one-to-one relationship between the
+handlers and the handler values used to reflect operations. By modifying
+the OxCaml runtime to allow values on the locals stack that directly
+point to the fibers used to implement effect handlers we can reflect
+operations by jumping directly to the relevant handler. This is roughly
+the implementation technique described by Ma et al.\cite{ma2024lexical}.
 
@@
 
+Effect reflection is closely related to \emph{lexical effect handlers}
+in systems such as Effekt\cite{brachthauser2020effects} or Biernacki et
+al. \cite{biernacki2019binders}. Lexical effect handlers have a value
+that represents the effect handler, similar to $\yoneda{\Sigma}$ values,
+but these values are second class and require their own special typing
+rules. More recent work on Effekt\cite{brachthauser2022effects}
+addresses the second-class nature of these values, but requires further
+extending the type system by tracking the set of capabilities used by
+each value.
+
+Effect reflection is also closely related to \emph{named effect
+  handlers}\cite{xie2022first}. Named effect handlers use higher-rank
+polymorphism to emulate locality tracking in much the same way as the
+\lstinline[style=haskell]{St} monad in Haskell. This still requires
+using an effect system to track which named effects are used by a
+computation, in order to ensure that those effects do not escape the
+scope of their handler. Using higher-rank polymorphism also prevents
+global type inference, requiring complex inference that extends beyond
+classic Hindly-Milner, and demanding a higher annotation burden on the
+programmer.
+
 \subsection{Monadic reflection}
 
-\subsection{Semantics of algebraic effects}
+Effect reflection is a variation on monadic
+reflection\cite{filinski1999representing}. By restricting ourselves to
+algebraic monads with no equations we remove the need to manually define
+\emph{glue} that describes the interaction between the monad being
+reified and the current ambient monad. By using locality and
+$\yoneda{\Sigma}$ values we avoid the need to have an effect system.
+
+\subsection{Semantics of algebraic effect handlers}
+
+Hyland et al.\cite{hyland2006combining} investigate combining algebraic
+effects, including the key property of sums of algebraic theories (our
+Lemma \ref{lemma:sums}) on which handlers are based.
+
+The original work on algebraic effect
+handlers\cite{plotkin2009handlers,plotkin2013handling} gives a semantics
+in terms of free algebras but the semantics work with a fixed algebraic
+theory: handlers do not actually change the ambient monad.
 
@@
 
+Our use of a comonadic global modality to track values which do not
+depend on the ambient monad is essentially the same as the purity
+comonad described by Choudhury and
+Krishnaswami\cite{choudhury2020recovering}. Their semantic description
+of their comonad is very different -- based on a semantics of
+capabilities -- but the typing rules, and the practical result, are the
+same.
+
 \subsection{Locality and presheaves}
 
-\section{Future work}
-\label{sec:future-work}
+Our presheaf semantics are closely related to presheaf semantics for
+higher-order abstract syntax\cite{hu2022category}. In both cases
+presheaves are used to index the semantics of types and terms by some
+notion of context, with the Yoneda embedding used to access this context
+and the global modality used to control it.
 

### 2025-07-09 paper.tex
Maybe a first draft?

@@
 
-\paragraph{Algebraic effects} Algebraic effects and handlers are an
-increasingly popular approach to programming with effects. Handlers
-allow the programmer to add additional effects to the \emph{ambient
-  monad} of their language -- the monad describing which effects an
-ordinary function can perform -- for some region of their program.
+\subsection{Algebraic effects}
+Algebraic effects and handlers are an increasingly popular approach to
+programming with effects. Handlers allow the programmer to add
+additional effects to the \emph{ambient monad} of their language -- the
+monad describing which effects an ordinary function can perform -- for
+some region of their program.
 
@@
 
-\paragraph{Locality and effect reflection}
+\subsection{Locality and effect reflection}
 Locality captures the general idea of a value depending on some notion
@@
 
-\paragraph{Effect reflection in OxCaml}
+\subsection{Effect reflection in OxCaml}
 OxCaml\cite{lorenzen2024oxidizing} provides type system support for
@@
 
-\todo{Add some mention of connection to lexical and named handlers}
-
-\paragraph{Denotational semantics} The design of effect reflection falls
+\subsection{Denotational semantics} The design of effect reflection falls
 out naturally from a simple denotational semantics of algebraic
@@
 
-\item The semantics of our handler types $Y$ are given by the Yoneda
-  embedding, and our effect reflection isomorphism becomes a simple
-  consequence of the Yoneda lemma.
+\item The semantics of our handler types $\yoneda{\Sigma}$ are given by
+  the Yoneda embedding, and our effect reflection isomorphism becomes a
+  simple consequence of the Yoneda lemma and some properties of the sums
+  of algebraic theories.
 \end{itemize}
 
-\paragraph{Structure of the paper} We proceed as follows:
+\subsection{Structure of the paper} We proceed as follows:
 \begin{itemize}
-\item In Section \ref{sec:pracice}, we give an overview of effect reflection in
-  practice in OxCaml
-\item In Section \ref{sec:semantics}, we describe the constructions for a
-  novel denotational semantics of algebraic effects
+\item In Section \ref{sec:practice}, we give an overview of effect reflection in
+  practice in OxCaml.
 \item In Section \ref{sec:language}, we define a simply typed calculus featuring
-  effect reflection along with its denotational semantics
+  effect reflection.
+\item In Section \ref{sec:semantics}, we describe our denotational
+  semantics of effect reflection.
 \item In Section \ref{sec:implementation}, we give an implementation of effect
@@
   inherits from OCaml.
-\item In Section \ref{sec:related-work}, we discuss related work, and in
-  Section \ref{sec:future-work}, we discuss future work.
+\item In Section \ref{sec:related-work}, we discuss related work.
 \end{itemize}
@@
 \section{Effect reflection in practice}
-\label{sec:pracice}
+\label{sec:practice}
 
@@
 algebra of the corresponding effect -- otherwise known as the freer
-monad\cite{??}. \lstinline[style=oxcaml]{Handler(O).t} is the type of
-mappings from the operations of \lstinline[style=oxcaml]{O} into the
-operations of the current ambient monad. Alternatively,
+monad\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
+the type of mappings from the operations of \lstinline[style=oxcaml]{O}
+into the operations of the current ambient monad. Alternatively,
 \lstinline[style=oxcaml]{Handler(O).t} can be described as representing
@@
 This shows how to implement state using effect reflection, but note that
-the same interface can also be implemented direcly using mutable state
+the same interface can also be implemented directly using mutable state
 and locality:
@@
 
-This transformation from effect relection to local mutable state works
+This transformation from effect reflection to local mutable state works
 for any \emph{tail-resumptive} effect handler.
@@
 
-To acheive this more general form of composition, we require an
+To achieve this more general form of composition, we require an
 additional capability in our effect reflection API. We need to be able
@@
 The literature on lexical effect handlers often ignores this issue. For
-example, \cite{??} proposed a system which only supported directly
-nesting effect handlers and had no mechanism for writing a composable
-handler for generators that could be moved between different underlying
-asynchronous I/O handlers.
+example, Xie et al. \cite{xie2020effect} proposed a system which only
+supported directly nesting effect handlers and had no mechanism for
+writing a composable handler for generators that could be moved between
+different underlying asynchronous I/O handlers.
 
@@
 The need for a new copy of the interface for each type constructor arity
-is an artifact of OCaml's approach to higher-kinded types. Languages
+is an artefact of OCaml's approach to higher-kinded types. Languages
 with direct support for higher-kinded types should be able to support
@@
 algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
-  monad} of $\Sigma$ \cite{??}.
+  monad} of $\Sigma$ \cite{kiselyov2015freer}.
 
@@
 An operation type $\tau_p \rightarrowtail \tau_a$ indicates an operation
-parameterised by $\tau_p$ and with arity $\tau_a$.
+parameterized by $\tau_p$ and with arity $\tau_a$.
 
@@
 These rules are excluded because they are not $\beta$ rules. Instead
-they are better thought of as $\nu$ rules \cite{??}: additional
-equalities between neutral terms that are immediate consequences of
-structural induction. Rather than be included as part of ordinary
-reduction they can be done as an additional standardization pass on
-neutral terms after reduction and $\eta$-expansion have completed. This
-staged approach to computing equality of terms works because $\nu$ rules
-cannot produce additional $\beta$ redexes.
+they are better thought of as $\nu$ rules \cite{allais2013new}:
+additional equalities between neutral terms that are immediate
+consequences of structural induction. Rather than be included as part of
+ordinary reduction they can be done as an additional standardization
+pass on neutral terms after reduction and $\eta$-expansion have
+completed. This staged approach to computing equality of terms works
+because $\nu$ rules cannot produce additional $\beta$ redexes.
 
@@
   monad}: one that can be expressed as the free algebras of an algebraic
-theory. Note that this defintion of algebraic monad is not precise:
+theory. Note that this definition of algebraic monad is not precise:
 different choices for the definition of ``algebraic theory'' give you
@@
 where $op \types P \rightarrowtail A$ indicates an operation $op$
-parameterised by $P$ and with arity $A$.
+parameterized by $P$ and with arity $A$.
 
@@
 essentially one of size: if you try to write down the collection of
-algebirac operations for the continuation monad you get something that
+algebraic operations for the continuation monad you get something that
 is too large to be a set. This is closely related to the strict
@@
   A \emph{Lawvere theory} is a category $L$ with finite products in
-  which every object is isomorphic to some finite cartesion product
+  which every object is isomorphic to some finite cartesian product
   $x^n = x \times \ldots \times x$ of a distinguished object $x$.
@@
 elements which do not depend on the ambient monad. This is very similar
-to the purity comonad in ??\cite{??}. In particular,
+to the purity comonad of Choudhury and
+Krishnaswami\cite{choudhury2020recovering}. In particular,
 \begin{equation*}
@@
 $T_{L + \Sigma}(S)$ the free algebras of the sum of an algebraic theory
-$L$ and an algebraic signature $\Sigma$, spliting out and handling
+$L$ and an algebraic signature $\Sigma$, splitting out and handling
 $\Sigma$, leaving a computation in only the algebraic theory
@@
 \end{lemma}
-where $\mu X.G(X)$ is the intial algebra of some polynomial endofunctor
+where $\mu X.G(X)$ is the initial algebra of some polynomial endofunctor
 $G$, and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$
@@
 algebras of polynomial endofunctors, i.e. W-types, so we can construct
-the required initial algebra as a presheave. Given an algebraic
-signature $\Sigma$ and a presheave $P$, we define the following family
+the required initial algebra as a presheaf. Given an algebraic
+signature $\Sigma$ and a presheaf $P$, we define the following family
 of initial algebras:
@@
 global type inference, requiring complex inference that extends beyond
-classic Hindly-Milner, and demanding a higher annotation burden on the
+classic Hindley-Milner, and demanding a higher annotation burden on the
 programmer.

### 2025-07-09 paper.tex
Some edits

@@
 operations can only be performed when they are available in the ambient
-monad. This provides the same guarantees as using an effect system.
+monad. This provides the same guarantees as using an effect system
+without any type system machinery beyond tracking locality.
 
@@
 
-The semantics is defined in terms of presheaves of the opposite of the
-category of \emph{bounded Lawvere theories}. Bounded Lawvere theories
-are a categorical representation of algebraic theories: we represent the
-ambient algebraic monad by the corresponding algebraic theory. The key
-constructs of effect reflection are then defined as follows:
+The semantics is defined in terms of presheaves of \emph{bounded Lawvere
+  theories}. Bounded Lawvere theories are a categorical representation
+of algebraic theories: we represent the ambient algebraic monad by the
+corresponding algebraic theory. The key constructs of effect reflection
+are then defined as follows:
 
@@
 %\begin{lstlisting}[style=oxcaml]
-%module Half_reflection (L : Op) (R : Op) : sig
+%  module Half_reflection (L : Op) (R : Op) : sig
 %    module L_eff = Reflection(L)

### 2025-07-10 paper.tex
More edits

@@
-\documentclass[acmsmall, screen, nonacm]{acmart}
+\documentclass[acmsmall, screen, review, anonymous]{acmart}
 
@@
 \begin{abstract}
-  Algebraic effects and handlers allow the programmer to extend their
+  Programming languages increasingly support algebraic effects and
+  handlers for writing effectful code in a principled way by extending a
   language's \emph{ambient monad} \textbf{--} the monad describing which
   effects ordinary functions can perform \textbf{--} with additional
-  effectful operations.
-
-  Locality captures the general idea of a value depending on some notion
-  of ``location''. We can represent it in a type system using a
-  comonadic \emph{global modality} $\glob$ that tracks which values do
-  not depend on their location. A particularly useful notion of
-  ``location'' is the ambient monad.
-
-  We introduce an alternative presentation of algebraic effects and
-  handlers, \emph{effect reflection}, that provides, for any algebraic
-  signature $\Sigma$, an isomorphism:
-  \begin{equation*}
-    \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
-  \end{equation*}
-  where $\ind{\Sigma}$ is an inductive type representing the free
-  algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
-  to inject terms of $\Sigma$ into the current ambient
-  monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
-  $\Sigma$. This presentation falls out naturally from a simple
-  denotational semantics for algebraic effects in terms of presheaves
-  over bounded Lawvere theories.
-
-  Effect reflection lets you work with any algebraic monad using an
-  ordinary function type, allowing effectful code to be written in
-  direct style. This provides all the benefits of typed algebraic
-  effects and handlers without requiring an effect system or other
-  specific additions to the type system. We demonstrate effect
-  reflection by implementing it as a library in
-  OxCaml\cite{lorenzen2024oxidizing} and working through some classic
-  examples of programming with effects.
+  effectful operations. However, integrating effects into type systems
+  typically requires specialized effect systems that add complexity to
+  the language.
+
+  We introduce \emph{effect reflection}, an alternative presentation of
+  algebraic effects and handlers that provides, for any algebraic
+  signature $\Sigma$ and type $\tau$, an isomorphism:
+  \begin{equation*}
+    \ind{\Sigma}(\glob \tau)
+    \qquad \cong \qquad
+    \yoneda{\Sigma} \rightarrow \glob \tau
+  \end{equation*}
+  where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
+  $\glob \tau$ \textbf{--} sometimes called the \emph{freer monad} of
+  $\Sigma$, $\yoneda{\Sigma}$ is a type representing how to inject terms
+  of $\Sigma$ into the current ambient monad, and $\glob$ is a comonadic
+  \emph{global modality} used to track which values do not depend on the
+  ambient monad. This isomorphism emerges naturally from a denotational
+  semantics for algebraic effects in terms of presheaves over bounded
+  Lawvere theories.
+
+  Effect reflection provides all the expressiveness of typed algebraic
+  effects without requiring language-level support for effect systems. We
+  demonstrate this by implementing effect reflection as a library in
+  OxCaml~\cite{lorenzen2024oxidizing} and showing how classic effectful
+  programs can be written in natural direct style.
 \end{abstract}
@@
 \subsection{Algebraic effects}
-Algebraic effects and handlers are an increasingly popular approach to
-programming with effects. Handlers allow the programmer to add
-additional effects to the \emph{ambient monad} of their language -- the
-monad describing which effects an ordinary function can perform -- for
-some region of their program.
+Algebraic effects and handlers have emerged as a powerful approach to
+structured programming with effects. Handlers allow the programmer to
+temporarily extend the \emph{ambient monad} of their language -- the
+monad describing which effects an ordinary function can perform -- with
+additional effectful operations.
 
@@
 
-Note that, in OCaml, the effects are \emph{untracked}: nothing prevents
-attempting to perform an algebraic effect operation that is not
-supported by the current ambient monad, resulting in an
+However, effect handlers face a fundamental tension: \emph{untracked}
+effects are unsafe, while \emph{tracked} effects require complex type
+system extensions.
+
+In OCaml, the effects are untracked: nothing prevents attempting to
+perform an algebraic effect operation that is not supported by the
+current ambient monad, resulting in an
 \lstinline[style=oxcaml]{Unhandled} exception being raised.
 
-Other languages track effects to ensure that operations are only
-performed when they are supported. However, these require extending the
-type system with explicit mechanisms for tracking effects, as in
-Koka\cite{leijen2017type}, or for tracking a general notion of
-capability, as in Effekt\cite{brachthauser2022effects}.
-
-The ``Lightweight effect polymorphism'' approach of Brachthauser et
-al. elegantly manages to track effects without requiring explicit
-effect-tracking machinery in the type system. However, it is restricted
-to second-class functions. We would like to extend this approach to
-first-class functions without having to fall back to explicit
-effect-tracking machinery.
+In languages like Koka~\cite{leijen2017type} the type system is extended
+to explicitly track the effects that a function may perform. This
+prevents the unsafety on an unhandled effect but adds complexity to the
+language.
+
+The ``effects as capabilities'' approach of the Effekt
+language~\cite{brachthauser2020effects} points the way by elegantly
+tracking effects without requiring explicit effect-tracking machinery in
+the type system. However, it is restricted to second-class
+functions\footnote{That restriction is lifted in more recent work on Effekt~\cite{brachthauser2022effects}, but only by falling back to explicit
+effect-tracking machinery.}. We aim to extend this approach to first
+class functions.
 
@@
 \emph{global modality} $\glob$ that tracks which values do not depend on
-their location. A particularly useful notion of ``location'' is the
-ambient monad.
-
-Treating the ambient monad as a notion of location allows a language to
-provide, for any algebraic signature $\Sigma$, an isomorphism:
-\begin{equation*}
-  \ind{\Sigma}(\glob \tau) \qquad \cong \qquad \yoneda{\Sigma} \rightarrow \glob \tau
-\end{equation*}
-where $\ind{\Sigma}$ is an inductive type representing the free
-algebra of $\Sigma$, and $\yoneda{\Sigma}$ is a type representing how
-to inject terms of $\Sigma$ into the current ambient
-monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad} of
-$\Sigma$. We call this isomorphism \emph{effect reflection}.
-
-The ability to use the ordinary arrow type
-$\yoneda{\Sigma} \mathbin{\rightarrow} \glob X$ in place of the monad
-$\ind{\Sigma}(\glob X)$ allows us to conveniently write effectful code,
-using multiple effects, in direct style, just as algebraic effects do.
+their location. Our key insight is that the ambient monad can be viewed
+as a form of location and controlled using $\glob$.
+
+Based on this insight, we present \emph{effect reflection}, providing,
+for any algebraic signature $\Sigma$ and type $\tau$, an isomorphism:
+\begin{equation*}
+  \ind{\Sigma}(\glob \tau)
+  \qquad \cong \qquad
+  \yoneda{\Sigma} \rightarrow \glob \tau
+\end{equation*}
+where $\ind{\Sigma}(\glob \tau)$ is an inductive type representing the
+free algebra of $\Sigma$ at $\glob \tau$ \textbf{--} sometimes called
+the \emph{freer monad} of $\Sigma$, and $\yoneda{\Sigma}$ is a type
+representing how to inject terms of $\Sigma$ into the current ambient
+monad.
+
+This isomorphism reveals that effectful computations can be represented
+as ordinary functions, enabling direct-style programming with full
+effect safety.
 
 \subsection{Effect reflection in OxCaml}
-OxCaml\cite{lorenzen2024oxidizing} provides type system support for
+OxCaml~\cite{lorenzen2024oxidizing} provides type system support for
 tracking locality and runtime support for algebraic effects. For
@@
 \lstinline[style=oxcaml]{[]s}. To get parameters or results that are not
-implicitly global, we write
+implicitly global we write \lstinline[style=oxcaml]{@ local} on them,
 e.g. \lstinline[style=oxcaml]{t @ local -> s @ local}.
 
-Using this support, with can provide effect reflection as a simple API
-that we can use for programming with effects.  For example, we can use
-it to write an effect handler equivalent to the OCaml one above:
+Using this support, we can provide effect reflection as a simple API.
+We represent algebraic signatures as modules with a single type
+constructor, providing $\ind{\Sigma}$ as
+\lstinline[style=oxcaml]{'a Term(Sigma).t} and $\yoneda{Sigma}$ as
+\lstinline[style=oxcaml]{Handler(Sigma).t}.
+
+For example, we can use it to write an effect handler equivalent to the
+OCaml one above:
 \begin{lstlisting}[style=oxcaml]
@@
 
-Note that the locality of the handler value ensures that these
-operations can only be performed when they are available in the ambient
-monad. This provides the same guarantees as using an effect system
-without any type system machinery beyond tracking locality.
-
-\subsection{Denotational semantics} The design of effect reflection falls
-out naturally from a simple denotational semantics of algebraic
-effects. The underlying idea is to index the semantics by the ambient
-monad. Rather than the semantics of a type being a set it is instead a
-function from ambient monads to set: if you tell it what the ambient
-monad is it can tell you what its set of values is.
-
-The semantics is defined in terms of presheaves of \emph{bounded Lawvere
-  theories}. Bounded Lawvere theories are a categorical representation
-of algebraic theories: we represent the ambient algebraic monad by the
-corresponding algebraic theory. The key constructs of effect reflection
-are then defined as follows:
-
+The locality of the \lstinline[style=oxcaml]{Handler(State).t} value
+passed to \lstinline[style=oxcaml]{f} ensures that these operations can
+only be performed when available in the ambient monad. The type system
+prevents this handler value from escaping the scope of
+\lstinline[style=oxcaml]{f}, which in turn prevents
+\lstinline[style=oxcaml]{get} or \lstinline[style=oxcaml]{set} from
+being called outside that scope.
+
+\subsection{Denotational semantics}
+
+This design emerges naturally from a denotational semantics where types
+are interpreted as presheaves over \emph{bounded Lawvere theories} -- a
+categorical framework that makes the connection between locality and
+effects precise.
+
+The key idea is to index the semantics by the ambient monad: rather than
+interpreting types as sets, we interpret them as functions from ambient
+monads to sets. We represent ambient monads using their corresponding
+algebraic theories in the framework of bounded Lawvere theories.
+
+The key constructs of effect reflection are then defined as follows:
 \begin{itemize}
 \item The ambient monad can be defined as a monad $\mon$ on such
-  presheaves, which maps each algebraic theory to its free algebras. We use
-  $\mon$ to define the semantics of computations.
-
-\item The global modality $\glob$ on such presheaves is defined such
-  that $\glob(P)(L) = P(\initial)$. In other words, for a given presheaf
-  $P$ it gives the constant presheaf that always ignores the ambient
-  monad and instead returns $P$ at $\initial$, where $\initial$ is the
-  algebraic theory with no operations or equations. This allows us to
-  restrict a type to its terms that do not depend on the ambient monad.
+  presheaves, which maps each algebraic theory to its free algebras.
+
+\item The global modality $\glob P$ is defined as the constant presheaf
+  that always ignores the ambient monad and instead returns
+  $P(\initial)$, where $\initial$ is the algebraic theory with no
+  operations or equations.
 
 \item The semantics of our handler types $\yoneda{\Sigma}$ are given by
-  the Yoneda embedding, and our effect reflection isomorphism becomes a
-  simple consequence of the Yoneda lemma and some properties of the sums
-  of algebraic theories.
+  the Yoneda embedding
+
+\item The effect reflection isomorphism becomes a simple consequence of
+  the Yoneda lemma and some properties of the sums of algebraic
+  theories.
 \end{itemize}
@@
 
-Values in OxCaml are associated with both a type -- which describe how
+Values in OxCaml are associated with both a type -- which describes how
 values can be introduced and eliminated -- and a \emph{mode} -- which
-track properties related to operations other than introduction and
+tracks properties related to operations other than introduction and
 elimination. Modes are a product of a number of different axes, tracking
@@
 
-Locality tracks the ability of a value to leave \emph{regions}.  Regions
-describe some period of time during a program's execution. Every
-function call defines a corresponding region that, by default, lasts for
-the entire extent of the call. This means that, by default,
-\lstinline[style=oxcaml]{local} values cannot be returned from a
-function.
+Locality tracks whether a value can 'escape' the current
+\emph{region}. By default, the current region ends when a function
+returns, so local values cannot be returned from functions.
 \begin{lstlisting}[style=oxcaml]
@@
 
-To support functions that return local values, the region of a function
-can be ended early using the \lstinline[style=oxcaml]{exclave} construct.
+The \lstinline[style=oxcaml]{exclave} construct allows functions to
+return local values.
 \begin{lstlisting}[style=oxcaml]
@@
 \end{lstlisting}
+These types directly implement the types from our core isomorphism:
+\lstinline[style=oxcaml]{'a Term(Sigma).t} corresponds to
+$\ind{\Sigma}(\glob \tau)$, while
+\lstinline[style=oxcaml]{Handler(Sigma).t} corresponds to
+$\yoneda{\Sigma}$
+
 For an effect signature \lstinline[style=oxcaml]{O},
@@
 algebra of the corresponding effect -- otherwise known as the freer
-monad\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
+monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
 the type of mappings from the operations of \lstinline[style=oxcaml]{O}
@@
 monads achieves the ``lightweight effect polymorphism'' described by
-Brachth{\"a}user et al.\cite{brachthauser2020effects} without requiring
-a separate notion of second-class function.
+Brachth{\"a}user et al.~\cite{brachthauser2020effects} without requiring
+second-class functions -- locality provides the necessary restrictions
+while maintaining first-class status.
 
@@
 continuations. Note that OxCaml's \lstinline[style=oxcaml]{once}
-mode\cite{lorenzen2024oxidizing} allows us to enforce this restriction
+mode~\cite{lorenzen2024oxidizing} allows us to enforce this restriction
 statically.
@@
 If we tried to do something similar using our
-\lstinline[style=oxcaml]{handle_gen} handler, then we get and error:
+\lstinline[style=oxcaml]{handle_gen} handler, then we get an error:
 \begin{lstlisting}[style=oxcaml]
@@
 recursive datatype that is built from the terms of one of the effects
-and the operations of the other\cite{hyland2006combining}. We can define
+and the operations of the other~\cite{hyland2006combining}. We can define
 this type and associated isomorphism as:
@@
 The literature on lexical effect handlers often ignores this issue. For
-example, Xie et al. \cite{xie2020effect} proposed a system which only
+example, Xie et al.~\cite{xie2020effect} proposed a system which only
 supported directly nesting effect handlers and had no mechanism for
@@
 
+\subsection{Summary}
+
+This section has demonstrated how effect reflection provides a practical
+and expressive approach to algebraic effects through a simple library
+API. The key benefits include: safety guarantees through locality
+without requiring specialized effect systems, seamless composition of
+effects through nesting for common cases, support for more complex
+composition patterns through effect forwarding when needed, and the
+ability to achieve lightweight effect polymorphism with first-class
+functions. The examples show that effect reflection can handle the full
+spectrum of algebraic effects—from simple state management to complex
+asynchronous I/O and generators—while maintaining the direct-style
+programming that makes algebraic effects appealing. Most importantly,
+the approach scales from basic usage to sophisticated effect composition
+without requiring programmers to learn complex type system machinery
+beyond locality tracking.
+
 \section{A language with effect reflection}
@@
 
-The design of effect reflection falls out naturally from a semantics in
-terms of presheaves. To illustrate this semantics we first build a
-simple language with effect reflection.
-
-\subsection{The standard parts}
+The design of effect reflection falls out naturally from a denotational
+semantics in terms of presheaves. To make this connection precise and
+demonstrate the key properties of effect reflection, we present a
+simply-typed calculus that directly incorporates the constructs needed
+for effect reflection.
+
+\subsection{The base language}
 
@@
 and typing judgements for values and computations. The syntax and typing
-rules for these standard parts of the language are shown in
-Fig.\ref{fig:typing-standard}, and there isn't much interesting to say
-about them.
+rules for the base language are shown in Fig.\ref{fig:typing-standard}.
 \begin{figure}
@@
 $\glob$. In our setting being global means not depending on the ambient
-monad. For example, the function type $1 \rightarrow 1$ represents
-computations in the ambient monad -- e.g. if the ambient monad includes
-state then such a function may mutate that state -- whereas the global
-function type $\glob (1 \rightarrow 1)$ has only a single value: the
-function that does nothing and returns $()$.
+monad. For example, $1 \rightarrow 1$ represents computations that may
+use operations from the ambient monad, while $\glob (1 \rightarrow 1)$
+contains only pure computations that ignore the ambient monad.
 
@@
 algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
-  monad} of $\Sigma$ \cite{kiselyov2015freer}.
-
-As with other algebraic effects systems in the literature, we only
-support handlers for algebraic theories without any equations, even
-though our semantics is defined in terms of arbitrary theories. We
-define algebraic signatures $\Sigma$ as finite sets of operations:
+  monad} of $\Sigma$~\cite{kiselyov2015freer}.
+
+As is standard for algebraic effects systems, we only support handlers
+for algebraic theories without any equations, even though our semantics
+is defined in terms of arbitrary theories. We define algebraic
+signatures $\Sigma$ as finite sets of operations:
 \begin{equation*}
@@
 $\Sigma$, providing a value of type $\yoneda{\Sigma}$ for constructing
-those operations via reflection. Their syntax and typing rules are shown
-in Fig.\ref{fig:typing-reflection}.
+those operations via reflection.  These operations implement the core
+isomorphism of effect reflection: reification can convert functions
+$\yoneda{\Sigma} \rightarrow \glob \tau$ into terms $\ind{\Sigma}(\glob
+\tau)$, while reflection provides the inverse direction. Their syntax
+and typing rules are shown in Fig.\ref{fig:typing-reflection}.
 
@@
 These rules are excluded because they are not $\beta$ rules. Instead
-they are better thought of as $\nu$ rules \cite{allais2013new}:
+they are better thought of as $\nu$ rules~\cite{allais2013new}:
 additional equalities between neutral terms that are immediate
@@
 Using the global modality and $\grun{c}$ construct we can embed the
-Simply Typed Lambda Calculus (STLC) into this language as values. We
-define an embedding $\embed{\_}$ on types, contexts and terms of STLC in
+Simply Typed Lambda Calculus (STLC) into this language as values. This
+demonstrates that the language is a proper extension of STLC. We define
+an embedding $\embed{\_}$ on types, contexts and terms of STLC in
 Fig.\ref{fig:stlc}. Note that $\reducestov$ matches the behaviour of
@@
 The design of effect reflection falls out naturally from a simple
-denotational semantics built on presheaves. The key idea is to index the
-meaning of types by an ambient algebraic monad of effects: rather than
-have the semantics of each type be a set of elements, the semantics of
-each type is a mapping from the ambient monad to a set of elements.
+denotational semantics built on presheaves. This semantic foundation not
+only validates our approach but also makes the connection between
+locality and effects precise. The key idea is to index the meaning of
+types by an ambient algebraic monad of effects: rather than have the
+semantics of each type be a set of elements, the semantics of each type
+is a mapping from the ambient monad to a set of elements.
 
@@
   \end{lstlisting}
-  and they form a monad:
-  \begin{lstlisting}[style=oxcaml]
-  let return x = Return x
-  let rec bind x f =
-    match x with
-    | Return x -> f x
-    | Get((), k) -> Get((), (fun s -> bind (k s) f))
-    | Set(s, k) -> Set(s, (fun () -> bind (k ()) f))
-  \end{lstlisting}  
 \end{example}
 
-Traditionally, algebraic signatures are restricted to the case of a
-finite set of operations each with a finite arity. Most theoretical work
-in the algebraic effects literature extends this to the case of
-countable sets of operations with countable arities, which allows us to
-construct the signature for natural number state:
-
-\begin{example}
-  The algebraic signature for natural number state is:
-  \begin{equation*}
-    \{ \mathrm{get} \types 1 \rightarrowtail \nat;
-       \mathrm{set} \types \nat \rightarrowtail 1 \}
-  \end{equation*}
-\end{example}
-
-However, this extension is not sufficient to construct the signature for
-state containing a function $\nat \rightarrow \nat$. For that we must
-extend algebraic signatures further to allow for arbitrary sets of
-operations with arbitrary sets as arities. Such algebraic theories are
-called \emph{bounded infinitary algebraic theories} because for each
-such theory there is some regular cardinal $\kappa$ that is a bound on
-the arities of the theory. An algebraic theory with some bound $\kappa$
-corresponds to a monad with rank $\kappa$. Thus we take monads with a
-rank as our precise definition of algebraic monad.
+Traditional algebraic signatures are restricted to the case of a finite
+set of operations each with a finite arity, but programming languages
+require more generality. $\nat$ state needs countable arities, while
+$\nat \rightarrow \nat$ state requires uncountable arities.
+For these we must extend algebraic signatures to allow for arbitrary
+sets of operations with arbitrary sets as arities. Such algebraic
+theories are called \emph{bounded infinitary algebraic theories} because
+for each such theory there is some regular cardinal $\kappa$ that is a
+bound on the arities of the theory. An algebraic theory with some bound
+$\kappa$ corresponds to a monad with rank $\kappa$. Thus we take monads
+with a rank as our precise definition of algebraic monad.
 
@@
 
-Lawvere theories correspond to the traditional for of algebraic
+Lawvere theories correspond to the traditional form of algebraic
 theories: those with a finite set of operations with finite arities, but
@@
 For a category $\cat$, $\psh = \setc^{\catop}$ is the category of
-presheaves over $\cat$. It's objects are contravariant functors from
+presheaves over $\cat$. Its objects are contravariant functors from
 $\cat$ to $\setc$ and it's morphisms are natural transformations between
@@
 thought of as some form of context over which everything is
-parameterized. In our case we wish to be parameterized by a context of
-\emph{algebraic theories that can be mapped into the ambient
-  monad}. Note that the algebraic theories in such a context are
-\emph{contravariant}: they describe theories that can be handled by the
-ambient monad. As such, we use
+parameterized. We use presheaves over $\lawc$ because types should be
+parameterized by the ambient algebraic theory--but contravariantly,
+since we want to be parameterized by theories that can be handled by the
+ambient monad.
+
+As such, we use
 \begin{equation*}
@@
 to the purity comonad of Choudhury and
-Krishnaswami\cite{choudhury2020recovering}. In particular,
+Krishnaswami~\cite{choudhury2020recovering}. In particular,
 \begin{equation*}
@@
 
-The key to this process is the following lemma
-\cite{hyland2006combining}.
+The key to this process is the following
+lemma~\cite{hyland2006combining}.
 \begin{lemma}
@@
 $G$, and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$
-is the free monad.
+is the free monad. This lemma is crucial because it shows that
+computations using both ambient effects $L$ and signature $\Sigma$ can
+be reorganized as a recursive structure that separates the two effect
+sources -- exactly what effect handlers need.
 
@@
 in terms of OCaml's untracked effect
-handlers\cite{Sivaramakrishnan2021retrofitting}. The trick is to
+handlers~\cite{Sivaramakrishnan2021retrofitting}. The trick is to
 generate a fresh effect constructor for each reification, pass that
@@
 is given in Fig.\ref{fig:implementation}.
+
 \begin{figure}
@@
 actually necessary: there is a one-to-one relationship between the
-handlers and the handler values used to reflect operations. By modifying
-the OxCaml runtime to allow values on the locals stack that directly
-point to the fibers used to implement effect handlers we can reflect
-operations by jumping directly to the relevant handler. This is roughly
-the implementation technique described by Ma et al.\cite{ma2024lexical}.
-
+handlers and the handler values used to reflect operations. The linear
+search can be eliminated by extending the OxCaml runtime to support
+direct references to handlers. By allowing local values to point
+directly to the fiber structures that implement effect handlers,
+operations can jump immediately to the correct handler without
+traversing the stack. This optimization maintains the same interface
+while providing constant-time operation dispatch. It is roughly the
+implementation technique described by Ma et al.~\cite{ma2024lexical}.
+
+\pagebreak
 \section{Related work}
@@
 
+Effect reflection draws from several lines of research in programming
+language theory. We discuss the most relevant connections, highlighting
+how our approach relates to and differs from existing work on effect
+handlers, monadic reflection, and semantic foundations.
+
 \subsection{Lexical and named effect handlers}
@@
 Effect reflection is closely related to \emph{lexical effect handlers}
-in systems such as Effekt\cite{brachthauser2020effects} or Biernacki et
-al. \cite{biernacki2019binders}. Lexical effect handlers have a value
+in systems such as Effekt~\cite{brachthauser2020effects} or Biernacki et
+al.~\cite{biernacki2019binders}. Lexical effect handlers have a value
 that represents the effect handler, similar to $\yoneda{\Sigma}$ values,
 but these values are second class and require their own special typing
-rules. More recent work on Effekt\cite{brachthauser2022effects}
+rules. More recent work on Effekt~\cite{brachthauser2022effects}
 addresses the second-class nature of these values, but requires further
@@
 Effect reflection is also closely related to \emph{named effect
-  handlers}\cite{xie2022first}. Named effect handlers use higher-rank
+  handlers}~\cite{xie2022first}. Named effect handlers use higher-rank
 polymorphism to emulate locality tracking in much the same way as the
@@
 Effect reflection is a variation on monadic
-reflection\cite{filinski1999representing}. By restricting ourselves to
+reflection~\cite{filinski1999representing}. By restricting ourselves to
 algebraic monads with no equations we remove the need to manually define
-\emph{glue} that describes the interaction between the monad being
-reified and the current ambient monad. By using locality and
+\emph{glue} functions that describe the interaction between the monad
+being reified and the current ambient monad. By using locality and
 $\yoneda{\Sigma}$ values we avoid the need to have an effect system.
@@
 
-Hyland et al.\cite{hyland2006combining} investigate combining algebraic
+Hyland et al.~\cite{hyland2006combining} investigate combining algebraic
 effects, including the key property of sums of algebraic theories (our
@@
 The original work on algebraic effect
-handlers\cite{plotkin2009handlers,plotkin2013handling} gives a semantics
+handlers~\cite{plotkin2009handlers,plotkin2013handling} gives a semantics
 in terms of free algebras but the semantics work with a fixed algebraic
-theory: handlers do not actually change the ambient monad.
+theory: handlers do not actually change the ambient monad. Our presheaf
+semantics naturally accommodates changing ambient monads.
 
@@
 comonad described by Choudhury and
-Krishnaswami\cite{choudhury2020recovering}. Their semantic description
+Krishnaswami~\cite{choudhury2020recovering}. Their semantic description
 of their comonad is very different -- based on a semantics of
@@
 Our presheaf semantics are closely related to presheaf semantics for
-higher-order abstract syntax\cite{hu2022category}. In both cases
+higher-order abstract syntax~\cite{hu2022category}. In both cases
 presheaves are used to index the semantics of types and terms by some
@@
 
+\pagebreak
 \bibliography{references}

### 2025-07-10 paper.tex
Fix up code examples

@@
   module State_eff = Reflection(State)
+  open Term(State)
 
   let handle_state (f : Handler(State).t @ local -> 'a) (i : int) : 'a =
-    let rec loop s : 'a Term(State).t @ local -> 'a  = function
+    let rec loop (s : int) : 'a Term(State).t @ local -> 'a  = function
       | Return x -> x
@@
          (*@ \textcolor{red}{\textasciicircum} @*)
-  (*@ \textcolor{red}{This value is local and would escape its region.} @*)
+  (*@ \textcolor{red}{This value escapes its region.} @*)
 \end{lstlisting}
@@
   module State_eff = Reflection(State)
-
-  let handle_state f i =
-    let rec loop s = function
-      | Term(State).Return x -> x
-      | Term(State).Op(Get, k) -> loop s (k s)
-      | Term(State).Op(Set s', k) -> loop s' (k ())
+  open Term(State)
+
+  let handle_state i f =
+    let rec loop (s : int) = function
+      | Return x -> x
+      | Op(Get, k) -> loop s (k s)
+      | Op(Set s', k) -> loop s' (k ())
     in
@@
 
-  val handle_state : (handler @ local -> 'a) @ local -> int -> 'a
+  val handle_state : int -> (handler @ local -> 'a) @ local -> 'a
   val get : handler @ local -> int
@@
 
-  let handle_state f i = f (ref i)
+  let handle_state i f = f (ref i)
   let get r = !r
@@
   let total l =
-    handle_state (fun h ->
+    handle_state 0 (fun h ->
       List.iter (fun x -> set h (get h + x)) l;
@@
   module Gen_eff = Reflection(Gen)
+  open Term(Gen)
 
@@
     let rec loop = function
-      | Term(Gen).Return () -> Finished
-      | Term(Gen).Op(Gen i, k) -> More(i, k)
+      | Return () -> Finished
+      | Op(Gen i, k) -> More(i, fun () -> loop (k ()))
     in
     loop (Gen_eff.reify f)
+
 \end{lstlisting}
@@
   module Await_eff = Reflection(Await)
+  open Term(Await)
 
@@
     let rec loop = function
-      | Term(Await).Return x -> Deferred.return x
-      | Term(Await).Op(Await d, k) ->
+      | Return x -> Deferred.return x
+      | Op(Await d, k) ->
           Deferred.bind d (fun x -> loop (k x))
@@
     handle_state i (fun sh ->
-      List.iter files (fun fl ->
+      List.iter (fun fl ->
         let s = await ch (File.read fl) in
         let i = int_of_string s in
-        set sh (i + get sh));
+        set sh (i + get sh)) files;
       get sh))
@@
   handle_await (fun ch ->
-    handle_gen i (fun gh ->
-      List.iter files (fun fl ->
+    handle_gen (fun gh ->
+      List.iter (fun fl ->
         let s = await ch (File.read fl) in
         let i = int_of_string s in
-        gen gh i)))
+        gen gh i) files))
   (*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
@@
   module Project(L : Op) (R : Op) : sig
-    val outl : Handler(Sum(L)(R)).t -> Handler(L).t
-
-    val outr : Handler(Sum(L)(R)).t -> Handler(R).t
+    val outl : Handler(Sum(L)(R)).t @ local -> Handler(L).t @ local
+
+    val outr : Handler(Sum(L)(R)).t @ local -> Handler(R).t @ local
   end
@@
 
-%\begin{lstlisting}[style=oxcaml]
-%  module Half_reflection (L : Op) (R : Op) : sig
-%    module L_eff = Reflection(L)
-%    module R_eff = Reflection(R)
-%    module Sum_eff = Reflection(Sum(L)(R))
-%    module Proj = Project(L)(R)
-%
-%    let reify (f : Handler(L).t @ local -> Handler(R).t @ local -> 'a) lh =
-%      let rec loop lh = function
-%        | Term(Sum(L)(R)).Return ->
-%            Half_term(L)(R).Return v
-%        | Term(Sum(L)(R)).Op(Left op, k) ->
-%            loop lh (k (L_eff.perform op lh))
-%        | Term(Sum(L)(R)).Op(Right op, k) ->
-%            Half_term(L)(R).Op(op, (fun res lh -> loop lh (k res)))
-%      in
-%      loop (Sum_eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))) lh
-%
-%    let rec reflect (f : Handler(L).t @ local -> 'a Half_term(L)(R).t) lh rh =
-%      match f lh with
-%      | Return x -> x
-%      | Op(op, k) -> reflect (k (R_eff.perform op rh)) lh rh
-%end
-%\end{lstlisting}
-
 Now we can use this machinery to conveniently build a handler of
@@
   module Await_gen_eff = Half_reflection(Await)(Gen)
+  open Half_term(Await)(Gen)
 
@@
       let rec loop = function
-      | Half_term(Await)(Gen).Return () -> Finished
-      | Half_term(Await)(Gen).Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
+      | Return () -> Finished
+      | Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
     in
@@
   val handle_gen_in_await :
-    (Handler(Await).t @ local -> Handler(Gen).t @ local -> 'a)
+    (Handler(Await).t @ local -> Handler(Gen).t @ local -> unit)
     -> Handler(Await).t @ local
@@
     type ('a, 'e) t =
-      | Return : 'a @@ global -> 'a t
-      | Op : ('r, 'e) O.t @@ global * ('r -> 'a t) -> 'a t
+      | Return : 'a @@ global -> ('a, 'e) t
+      | Op : ('r, 'e) O.t @@ global * ('r -> ('a, 'e) t) -> ('a, 'e) t
   end
@@
   module State = struct
-    type ('a, 'b) t =
-      | Get : 'b t
-      | Set : 'b -> unit t
+    type ('a, 's) t =
+      | Get : ('s, 's) t
+      | Set : 's -> (unit, 's) t
   end
@@
 
-  val handle : ('s handler @ local -> 'a) @ local -> 's -> 'a
+  val handle_state : 's -> ('s handler @ local -> 'a) @ local -> 'a
   val get : 's handler @ local -> 's
@@
 in terms of OCaml's untracked effect
-handlers~\cite{Sivaramakrishnan2021retrofitting}. The trick is to
-generate a fresh effect constructor for each reification, pass that
-constructor around as part of the \lstinline[style=oxcaml]{handler}, and
-then use it to perform operations during reflection. The full implementation
-is given in Fig.\ref{fig:implementation}.
+handlers~\cite{Sivaramakrishnan2021retrofitting}\footnote{For
+  convenience, we use the new handler syntax from OCaml 5.3. However,
+  the most recently released version of OxCaml is based on OCaml 5.2 so
+  the actual implementation uses
+  \lstinline[style=oxcaml]{Effect.Deep.match_with} instead}. The trick
+is to generate a fresh effect constructor for each reification, pass
+that constructor around as part of the
+\lstinline[style=oxcaml]{handler}, and then use it to perform operations
+during reflection. The full implementation is given in
+Fig.\ref{fig:implementation}.
 
@@
 module Reflection (O : Op) = struct
+  open Term(O)
+  open Handler(O)
 
   let reify f =
-    let module Handler =
-      struct type 'a Effect.t += H : 'a O.t -> 'a Effect.t end
+    let module Eff =
+      struct type 'a Effect.t += C : 'a O.t -> 'a Effect.t end
     in
-    let handler : Handler(O).t =
-      { perform = fun o -> Effect.perform (Handler.H o) }
+    let handler =
+      { perform = fun o -> Effect.perform (Eff.C o) }
     in
@@
     | v -> Return v
-    | effect Handler.H o, k -> Op(o, fun x -> Effect.continue k x)
+    | effect Eff.C o, k -> Op(o, fun x -> Effect.continue k x)
 
   let reify_local f = exclave
-    let module Handler =
-      struct type 'a Effect.t += H : 'a O.t -> 'a Effect.t end
+    let module Eff =
+      struct type 'a Effect.t += C : 'a O.t -> 'a Effect.t end
     in
-    let handler = { perform = fun o -> Effect.perform (Handler.H o) } in
+    let handler = { perform = fun o -> Effect.perform (Eff.C o) } in
     match f handler with
     | v -> Return v
-    | effect Handler.H o, k -> Op(o, fun x -> Effect.continue k x)
+    | effect Eff.C o, k -> Op(o, fun x -> Effect.continue k x)
 
@@
 module Project (L : Op) (R : Op) = struct
-
-  let outl {perform} =
-    {perform = fun op -> perform (Sum(L)(R).Left op)}
-
-  let outr {perform} =
-    {perform = fun op -> perform (Sum(L)(R).Right op)}
+  open Sum(L)(R)
+
+  let outl ({perform} : Handler(Sum(L)(R)).t) : Handler(L).t =
+    exclave {perform = fun op -> perform (Left op)}
+
+  let outr ({perform} : Handler(Sum(L)(R)).t) : Handler(R).t =
+    exclave {perform = fun op -> perform (Right op)}
 
@@
 
-\pagebreak
 \section{Related work}

### 2025-07-10 paper.tex
Remove summary subsection

@@
 
-\subsection{Summary}
-
-This section has demonstrated how effect reflection provides a practical
-and expressive approach to algebraic effects through a simple library
-API. The key benefits include: safety guarantees through locality
-without requiring specialized effect systems, seamless composition of
-effects through nesting for common cases, support for more complex
-composition patterns through effect forwarding when needed, and the
-ability to achieve lightweight effect polymorphism with first-class
-functions. The examples show that effect reflection can handle the full
-spectrum of algebraic effects—from simple state management to complex
-asynchronous I/O and generators—while maintaining the direct-style
-programming that makes algebraic effects appealing. Most importantly,
-the approach scales from basic usage to sophisticated effect composition
-without requiring programmers to learn complex type system machinery
-beyond locality tracking.
-
 \section{A language with effect reflection}

### 2025-07-10 paper.tex
Final tweaks

@@
 \newcommand{\cdiv}[1]{#1/\Box}
+\newcommand{\bdiv}[1]{\cdiv{\mathop{!_{#1}}}}
 \newcommand{\return}[1]{\mathop{\mathrm{return}} #1}
@@
   where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
-  $\glob \tau$ \textbf{--} sometimes called the \emph{freer monad} of
-  $\Sigma$, $\yoneda{\Sigma}$ is a type representing how to inject terms
-  of $\Sigma$ into the current ambient monad, and $\glob$ is a comonadic
-  \emph{global modality} used to track which values do not depend on the
-  ambient monad. This isomorphism emerges naturally from a denotational
-  semantics for algebraic effects in terms of presheaves over bounded
-  Lawvere theories.
+  $\glob \tau$, $\yoneda{\Sigma}$ is a type representing how to inject
+  terms of $\Sigma$ into the current ambient monad, and $\glob$ is a
+  comonadic \emph{global modality} used to track which values do not
+  depend on the ambient monad. $\ind{\Sigma}$ is sometimes called the
+  \emph{freer monad} of $\Sigma$. This isomorphism emerges naturally
+  from a denotational semantics for algebraic effects in terms of
+  presheaves over bounded Lawvere theories.
 
@@
 Locality captures the general idea of a value depending on some notion
-of ``location''. We can represent it in a type system using a comonadic
-\emph{global modality} $\glob$ that tracks which values do not depend on
-their location. Our key insight is that the ambient monad can be viewed
-as a form of location and controlled using $\glob$.
-
-Based on this insight, we present \emph{effect reflection}, providing,
-for any algebraic signature $\Sigma$ and type $\tau$, an isomorphism:
+of ``location''. Our key insight is that the ambient monad can be viewed
+as a form of location, allowing us to reuse existing type system
+features for tracking locality.
+
+We can represent locality in a type system using a comonadic
+\emph{global modality} $\glob$ that tracks which values do \emph{not}
+depend on their location. Values of an ordinary type $\tau$ are assumed
+to potentially depend on the current location, whereas values of the
+global type $\glob \tau$ are known not to.
+
+In our setting being global means not depending on the ambient
+monad. For example, $A \rightarrow B$ represents computations that may
+use operations from the ambient monad, while $\glob (A \rightarrow B)$
+contains only pure computations that ignore the ambient monad.
+
+Based on this relationship between locality and effects, we present
+\emph{effect reflection}, providing, for any algebraic signature
+$\Sigma$ and type $\tau$, an isomorphism:
 \begin{equation*}
@@
 where $\ind{\Sigma}(\glob \tau)$ is an inductive type representing the
-free algebra of $\Sigma$ at $\glob \tau$ \textbf{--} sometimes called
-the \emph{freer monad} of $\Sigma$, and $\yoneda{\Sigma}$ is a type
-representing how to inject terms of $\Sigma$ into the current ambient
-monad.
+free algebra of $\Sigma$ at $\glob \tau$, and $\yoneda{\Sigma}$ is a
+type representing how to inject terms of $\Sigma$ into the current
+ambient monad. $\ind{\Sigma}$ is sometimes called the \emph{freer monad}
+of $\Sigma$~\cite{kiselyov2015freer}.
 
@@
 constructor, providing $\ind{\Sigma}$ as
-\lstinline[style=oxcaml]{'a Term(Sigma).t} and $\yoneda{Sigma}$ as
+\lstinline[style=oxcaml]{'a Term(Sigma).t} and $\yoneda{\Sigma}$ as
 \lstinline[style=oxcaml]{Handler(Sigma).t}.
@@
   module State_eff = Reflection(State)
-  open Term(State)
 
@@
 interpreting types as sets, we interpret them as functions from ambient
-monads to sets. We represent ambient monads using their corresponding
-algebraic theories in the framework of bounded Lawvere theories.
+monads to sets. We represent ambient monads by their corresponding
+algebraic theories using bounded Lawvere theories.
 
@@
 
-Locality tracks whether a value can 'escape' the current
+Locality tracks whether a value can escape the current
 \emph{region}. By default, the current region ends when a function
@@
 \end{lstlisting}
-
 The \lstinline[style=oxcaml]{exclave} construct allows functions to
@@
 
-By default, modes are deep: a local \lstinline[style=oxcaml]{string list}
-is a local list of local strings. However, the depth that a mode applies
-can be controlled via \emph{modalities}. Record fields can be annotated
-with \lstinline[style=oxcaml]{global} modality to indicate that the
-contents of the field are global even when the surrounding record is
-local. For example:
+By default, modes are deep: a local
+\lstinline[style=oxcaml]{string list} is a local list of local
+strings. However, the depth that a mode applies can be controlled via
+\emph{modalities}. Record fields can be annotated with the
+\lstinline[style=oxcaml]{global} modality to indicate that the contents
+of the field are global even when the surrounding record is local. For
+example:
 \begin{lstlisting}[style=oxcaml]
@@
 algebra of the corresponding effect -- otherwise known as the freer
-monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t} is
-the type of mappings from the operations of \lstinline[style=oxcaml]{O}
-into the operations of the current ambient monad. Alternatively,
-\lstinline[style=oxcaml]{Handler(O).t} can be described as representing
-algebraic effect handlers that are handling the current computation --
-hence the name.
+monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t}
+is the type of mappings from the operations of
+\lstinline[style=oxcaml]{O} into the operations of the current ambient
+monad. Alternatively, \lstinline[style=oxcaml]{Handler(O).t} can be
+described as representing algebraic effect handlers that are handling
+the current computation.
 
@@
 \begin{lstlisting}[style=oxcaml]
-  handle_await (fun ch ->
+  handle_await (fun ah ->
     handle_state i (fun sh ->
       List.iter (fun fl ->
-        let s = await ch (File.read fl) in
+        let s = await ah (File.read fl) in
         let i = int_of_string s in
@@
 uses \lstinline[style=oxcaml]{reify_local} and so is able to operate on a
-local computation that closes over the local handler value
-\lstinline[style=oxcaml]{ch}.
+local computation that closes over the local await handler value
+\lstinline[style=oxcaml]{ah}.
 
@@
 \begin{lstlisting}[style=oxcaml]
-  handle_await (fun ch ->
+  handle_await (fun ah ->
     handle_gen (fun gh ->
       List.iter (fun fl ->
-        let s = await ch (File.read fl) in
+        let s = await ah (File.read fl) in
         let i = int_of_string s in
         gen gh i) files))
-  (*@ \textcolor{red}{Error: let s = await ch (File.read fl) in} @*)
+  (*@ \textcolor{red}{Error: let s = await ah (File.read fl) in} @*)
                        (*@ \textcolor{red}{\textasciicircum\textasciicircum} @*)
@@
 of the stream requires performing additional
-\lstinline[style=oxcaml]{Await} operations. Whenever we force part of the
-stream we'll need to provide a handler for those
-\lstinline[style=oxcaml]{Await} operations.
+\lstinline[style=oxcaml]{Await} operations, so whenever we force part of
+the stream we'll need to provide a handler for those
+\lstinline[style=oxcaml]{Await} operations again.
 
@@
 
-This is because this composition requiring forwarding is much less
-common than nesting and it fundamentally requires each use of the
-\lstinline[style=oxcaml]{Await} operation to perform work proportional to
-the number of handlers it is forwarded through -- often referred to as a
-``linear search''. There has been a tendency to treat this linear search
-as some form of deficiency of algebraic effect handlers, but it is
-simply a classic trade-off between the extra expressivity allowed by an
-indirection and the runtime cost of traversing that indirection. With
+This is because composition requiring forwarding is much less common
+than nesting and it fundamentally requires each use of the
+\lstinline[style=oxcaml]{Await} operation to perform work proportional
+to the number of handlers it is forwarded through -- often referred to
+as a ``linear search''. There has been a tendency to treat this linear
+search as some form of deficiency of algebraic effect handlers, but it
+is simply a classic trade-off between the extra expressivity allowed by
+an indirection and the runtime cost of traversing that indirection. With
 our effect reflection API we are able to use nesting for the cases that
@@
 \end{mathpar}
-\caption{Standard syntax and typing rules}
+\caption{Basic syntax and typing rules}
 \label{fig:typing-standard}
@@
 reflected, an inductive type $\ind{\Sigma}$ to represent the free
-algebra of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
+algebras of $\Sigma$. $\ind{\Sigma}$ is sometimes called the \emph{freer
   monad} of $\Sigma$~\cite{kiselyov2015freer}.
@@
 \begin{align*}
-\{ \; &\mathrm{left}(op_{1,1}) \types \tau_{p_{1,1}} \rightarrowtail \tau_{a_{1,1}} \seq
-        \ldots \seq
-        \mathrm{left}(op_{1, n}) \types \tau_{p_{1, n}} \rightarrowtail \tau_{a_{1, n}} \\
-      &\mathrm{right}(op_{2, 1}) \types \tau_{p_{2, 1}} \rightarrowtail \tau_{a_{2,1}} \seq
-        \ldots \seq
+\{ \; &\mathrm{left}(op_{1,1}) \types \tau_{p_{1,1}} \rightarrowtail \tau_{a_{1,1}}
+        \seq \ldots \seq
+        \mathrm{left}(op_{1, n}) \types \tau_{p_{1, n}} \rightarrowtail \tau_{a_{1, n}}
+        \seq \\
+      &\mathrm{right}(op_{2, 1}) \types \tau_{p_{2, 1}} \rightarrowtail \tau_{a_{2,1}}
+        \seq \ldots \seq
         \mathrm{right}(op_{2, m}) \types \tau_{p_{2, m}} \rightarrowtail \tau_{a_{2, m}}
@@
 $\ind{\Sigma}(\tau)$ has a constructor $\indintro{\Sigma}{op}{v_p}{v_k}$
-where $v_p \types \tau_p$ and
-$v_k \types \tau_a \rightarrow \ind{\Sigma}(\tau)$. $\ind{\Sigma}(\tau)$
+where $v_p \types \glob \tau_p$ and
+$v_k \types \glob \tau_a \rightarrow \ind{\Sigma}(\tau)$. $\ind{\Sigma}(\tau)$
 also has a constructor $\ret{\Sigma}{v}$ where $v \types \tau$.
@@
   \begin{align*}
-    &\indintro{\Sigma_{bs}}{get}{()}{\abs{x_a}{1 + 1}{\\
+    &\indintro{\Sigma_{bs}}{get}{()}{\\
+      &\quad\abs{x_a}{1 + 1}{\\
         &\qquad \letv{x_n}{\app{\mathit{negate}}{x_a}}{\\
@@
 ambient monad. Effect reflection appears in the language as two
-operations: reflection and reification. Reflection
-$\reflectraw{\Sigma}{op}$ uses a value of type $\yoneda{\Sigma}$ to map
+constructs: reflection and reification. Reflection
+$\performraw{v}{op}$ uses a value of type $\yoneda{\Sigma}$ to map
 the operation $op$ from $\Sigma$ into the ambient monad.  Reification
@@
 $\Sigma$, providing a value of type $\yoneda{\Sigma}$ for constructing
-those operations via reflection.  These operations implement the core
+those operations via reflection.  These constructs implement the core
 isomorphism of effect reflection: reification can convert functions
@@
 \begin{example}
-  The algebraic signature for boolean state is:
+  The algebraic signature for boolean state $\Sigma_{bs}$ is:
   \begin{equation*}
@@
 \begin{example}
-  The algebraic theory for boolean state adds the following equations to
-  the signature of boolean state:
+  The algebraic theory for boolean state $L_{bs}$ adds the following
+  equations to the signature of boolean state $\Sigma_{bs}$:
   \begin{align*}
@@
 \begin{example}
-  We can model the theory of boolean state using the carrier set
-  $2 \rightarrow 2$ and interpreting the operations with:
+  We can model the theory of boolean state $L_{bs}$ using the carrier
+  set $2 \rightarrow 2$ and interpreting the operations with:
   \begin{align*}
@@
 \begin{example}
-  The algebraic signature for boolean state can be written as:
+  The algebraic signature for boolean state $\Sigma_{bs}$ can be written
+  as:
   \begin{equation*}
@@
 
-Given a Lawvere theory $L$ and a set $A$ we can construct the free
-model, or \emph{free algebra}, $T_L(A)$. The values $T_L(A)$ are terms
-built using the elements of $A$ and the operations of $L$ quotiented by
-the equations in $L$. $T_L$ is a monad and $T$ is one half of an
-equivalence between algebraic theories and algebraic monads.
+Given an algebraic theory $L$ and a set $A$ we can construct
+the free model, or \emph{free algebra}, $T_L(A)$. The values $T_L(A)$
+are terms built using the elements of $A$ and the operations of $L$
+quotiented by the equations in $L$. $T_L$ is a monad and $T$ is one half
+of an equivalence between algebraic theories and algebraic monads.
 
 \begin{example}
-  The free algebras of the theory of boolean state are equivalent
-  to the type:
+  The free algebras of the signature of boolean state $T_{\Sigma_{bs}}$
+  are equivalent to the type:
   \begin{lstlisting}[style=oxcaml]
@@
   \end{lstlisting}
+  The free algebras of the theory of boolean state $T_{L_{bs}}$ are
+  equivalent to the above type quotiented by the equations of $L_{bs}$.
+\end{example}
+
+For an algebraic signature $\Sigma$ the free algebra $T_{\Sigma}$ is the
+free monad of an endofunctor $F_{\Sigma}$. The values of $F_{\Sigma}(A)$
+are of the form $op(a_1, \ldots, a_k)$ where $op$ is an operation of
+$\Sigma$ with arity $k$ and each $a_i$ is an element of $A$.
+\begin{example}
+  $F_{\Sigma_{bs}}$ is equivalent to the type:
+  \begin{lstlisting}[style=oxcaml]
+  type 'a t =
+    | Get of unit * (bool -> 'a)
+    | Set of bool * (unit -> 'a)
+  \end{lstlisting}
 \end{example}
@@
 require more generality. $\nat$ state needs countable arities, while
-$\nat \rightarrow \nat$ state requires uncountable arities.
-For these we must extend algebraic signatures to allow for arbitrary
-sets of operations with arbitrary sets as arities. Such algebraic
-theories are called \emph{bounded infinitary algebraic theories} because
-for each such theory there is some regular cardinal $\kappa$ that is a
-bound on the arities of the theory. An algebraic theory with some bound
-$\kappa$ corresponds to a monad with rank $\kappa$. Thus we take monads
-with a rank as our precise definition of algebraic monad.
+$\nat \rightarrow \nat$ state requires uncountable arities.  For these
+we must extend algebraic signatures to allow for arbitrary sets of
+operations with arbitrary sets as arities. Such algebraic theories are
+called \emph{bounded infinitary algebraic theories}. They are
+``bounded'' because for each such theory there is some regular cardinal
+$\kappa$ that is a bound on the arities of the theory. An algebraic
+theory with some bound $\kappa$ corresponds to a monad with rank
+$\kappa$. Thus we take monads with a rank as our precise definition of
+algebraic monad.
 
@@
   which every object is isomorphic to some finite cartesian product
-  $x^n = x \times \ldots \times x$ of a distinguished object $x$.
+  $X^n = X \times \ldots \times X$ of a distinguished object $X$.
 \end{definition}
 The objects of a Lawvere theory correspond to the different possible
-arities and the morphisms $x^n \rightarrow x$ correspond to terms
+arities and the morphisms $X^n \rightarrow X$ correspond to terms
 of $L$ with $n$ free variables quotiented by the equations in $L$. A
@@
   products in which every object is isomorphic to some small cartesian
-  product $\Pi_S x$ of a distinguished object $x$.
+  product $\Pi_S X$ of a distinguished object $X$.
 \end{definition}
@@
 $\lawc$ has an initial object $\initial$, which is the algebraic theory
-with no operations. It also has coproducts $L + M$ which is the theory
-made by unioning the operations and equations of $L$ and $M$.
+with no operations. It also has coproducts $L + L^{\prime}$ which is the
+theory made by unioning the operations and equations of $L$ and
+$L^{\prime}$.
 
@@
 types.
-\begin{align*}
-&\sem{1} = 1 \\
-&\sem{\tau_1 \times \tau_2} = \sem{\tau_1} \times \sem{\tau_2} \\
-&\sem{\tau_1 + \tau_2} = \sem{\tau_1} + \sem{\tau_2} \\
-\end{align*}
+\begin{equation*}
+\sem{1} = 1 \qquad
+\sem{\tau_1 \times \tau_2} = \sem{\tau_1} \times \sem{\tau_2} \qquad
+\sem{\tau_1 + \tau_2} = \sem{\tau_1} + \sem{\tau_2} \qquad
+\end{equation*}
 The semantics of the corresponding introduction and elimination forms
@@
 We also use products for the semantics of contexts:
-\begin{align*}
-&\sem{\cempty} = 1 \\
-&\sem{\lbind{\Gamma}{x}{\tau}} = \sem{\Gamma} \times \sem{\tau} \\
-\end{align*}
+\begin{equation*}
+\sem{\cempty} = 1 \qquad
+\sem{\lbind{\Gamma}{x}{\tau}} = \sem{\Gamma} \times \sem{\tau}
+\end{equation*}
 
@@
 free algebras are the ambient monad. In particular, we would like the
-interpretation of types of computations to be applications of the
-ambient monad.
-
-Given a Lawvere theory $L$, the underlying set of its free algebras form
-a monad $T_L$ on $\setc$. In addition to the unit of the monad, we can
-construct elements of $T_L$ using the operations of $L$. For an
+interpretation of computation types to be applications of the ambient
+monad.
+
+Given a Lawvere theory $L$, the underlying sets of its free algebras
+form a monad $T_L$ on $\setc$. In addition to the unit of the monad, we
+can construct elements of $T_L$ using the operations of $L$. For an
 operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$ and
@@
 free algebras monad to give us our semantics for function types.
-\begin{align*}
-  &\sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})} \\
-  &\sem{\Gamma \turnv \abs{x}{\tau_1}{c} \types \tau_1 \rightarrow \tau_2} =
-    \currymap(\sem{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2})
-     \\
-  &\sem{\Gamma \turnc \app{v_1}{v_2} \types \tau_2} =
-    \evmap
-    \circ \langle \sem{\Gamma \turnv v_1 \types \tau_1 \rightarrow \tau_2},
-                  \sem{\Gamma \turnv v_2 \types \tau_1} \rangle
-\end{align*}
-where $\evmap$ is the evaluation map and $\currymap$ the currying of
-exponentials.
+\begin{equation*}
+  \sem{\tau_1 \rightarrow \tau_2} = \arrow{\sem{\tau_1}}{\mon(\sem{\tau_2})}
+\end{equation*}
 
@@
   \glob \sem{\cdiv{\Gamma} \turnv v \types \tau}
-  \circ \eta \circ \cdiv{\Gamma} \\
+  \circ \eta \circ \bdiv{\Gamma} \\
 &\sem{\Gamma \turnv \gunbox{x}{v_1}{v_2} \types \tau_2} =
@@
   \circ \glob \sem{\cdiv{\Gamma} \turnc c \types \tau}
-  \circ \eta \circ \cdiv{\Gamma}\\
+  \circ \eta \circ \bdiv{\Gamma}\\
 \end{align*}
 where $\epsilon$ and $\eta$ are the counit and comultiplication of
-$\glob$, $\cdiv{\Gamma} \types \Gamma \rightarrow \cdiv{\Gamma}$ is the
+$\glob$, $\bdiv{\Gamma} \types \Gamma \rightarrow \cdiv{\Gamma}$ is the
 morphism that projects the global bindings from $\Gamma$, and $\rho_P$
@@
 For an operation $op \types P \rightarrowtail A$ of a Lawvere theory $L$
-and some set $S$, we can build a morphism:
-\begin{align*}
-&op_{\mon{S}} \types
+and some presheaf $P$, we can build a morphism:
+\begin{align*}
+&op_{\mon{P}} \types
   \homset{\pshlawcop}
-         {\yoneda{L} \times \glob P \times \arrow{\glob A}{\mon{S}}}
-         {\mon{S}} \\
-&{(op_{\mon{S}})}_{L^\prime}
+         {\yoneda{L} \times \glob P \times \arrow{\glob A}{(\mon{P})}}
+         {\mon{P}} \\
+&{(op_{\mon{P}})}_{L^\prime}
   = \lambda(y, p, k). (y(op))_{T_{L^\prime}}(p, \lambda a. k_{L^\prime}(id_{L^\prime}, a))
@@
 We use $\yoneda{L}$ as the semantics for our handler types, and
-$op_{\mon{S}}$ as the semantics for reflection.
+$op_{\mon{P}}$ as the semantics for reflection.
 \begin{align*}
@@
 In terms of our semantics, this means that computations of type
-$\glob \tau$ with a single variable of type $\yoneda{\Sigma}$ are
+$\glob \tau$ with a single free variable of type $\yoneda{\Sigma}$ are
 isomorphic to values of the free algebra of $\sem{\Sigma}$ at
@@
 where $\mu X.G(X)$ is the initial algebra of some polynomial endofunctor
-$G$, and $F_\Sigma$ is the endofunctor on $\setc$ for which $T_{\Sigma}$
-is the free monad. This lemma is crucial because it shows that
-computations using both ambient effects $L$ and signature $\Sigma$ can
-be reorganized as a recursive structure that separates the two effect
-sources -- exactly what effect handlers need.
+$G$. This lemma is crucial because it shows that computations using both
+ambient effects $L$ and the signature $\Sigma$ can be reorganized as a
+recursive structure that separates the two effect sources -- exactly
+what effect handlers need.
 
@@
 handlers and the handler values used to reflect operations. The linear
-search can be eliminated by extending the OxCaml runtime to support
+search could be eliminated by extending the OxCaml runtime to support
 direct references to handlers. By allowing local values to point
@@
 \label{sec:related-work}
-
-Effect reflection draws from several lines of research in programming
-language theory. We discuss the most relevant connections, highlighting
-how our approach relates to and differs from existing work on effect
-handlers, monadic reflection, and semantic foundations.
 
@@
 notion of context, with the Yoneda embedding used to access this context
-and the global modality used to control it.
-
-\pagebreak
+and the global modality used to control dependence on it.
+
 \bibliography{references}

### 2025-09-15 paper.tex
Remove parameterized effects subsection

@@
 
-\subsection{Parameterized effects and handlers}
-
-The state handlers in this paper so far have all been restricted to
-\lstinline[style=oxcaml]{int} state. This is a limitation of the effect
-reflection API that we have been using. We would like to support
-\emph{parameterized effects} -- allowing a single handler to be used
-with a whole family of effects. For example, a state handler
-parameterized by the type of the state.
-
-We can support parameterized effects by adding a module type for them,
-along with parameterized versions of \lstinline[style=oxcaml]{Term} and
-\lstinline[style=oxcaml]{Handler}:
-\begin{lstlisting}[style=oxcaml]
-  module type Op1 = sig type ('a, 'e) t end
-
-  module Term1 (O : Op1) : sig
-    type ('a, 'e) t =
-      | Return : 'a @@ global -> ('a, 'e) t
-      | Op : ('r, 'e) O.t @@ global * ('r -> ('a, 'e) t) -> ('a, 'e) t
-  end
-
-  module Handler1(O : Op1) : sig
-    type 'e t
-  end
-\end{lstlisting}
-with a corresponding parameterized version of
-\lstinline[style=oxcaml]{Reflection}. We can then define a parameterized
-state effect:
-\begin{lstlisting}[style=oxcaml]
-  module State = struct
-    type ('a, 's) t =
-      | Get : ('s, 's) t
-      | Set : 's -> (unit, 's) t
-  end
-\end{lstlisting}
-and write a handler that provides a parameterized interface for the
-state effect:
-\begin{lstlisting}[style=oxcaml]
-  type 's handler
-
-  val handle_state : 's -> ('s handler @ local -> 'a) @ local -> 'a
-  val get : 's handler @ local -> 's
-  val set : 's handler @ local -> 's -> unit
-\end{lstlisting}
-
-The need for a new copy of the interface for each type constructor arity
-is an artefact of OCaml's approach to higher-kinded types. Languages
-with direct support for higher-kinded types should be able to support
-parameterized effects of arbitrary arity with a single interface.
-
 \section{A language with effect reflection}

### 2025-09-15 paper.tex
Shorten the sums subsection by removing Half_term

@@
 
-\lstinline[style=oxcaml]{Sum(L)(R)} is the coproduct of the effects
-\lstinline[style=oxcaml]{L} and \lstinline[style=oxcaml]{R}. Terms in
-the free algebras of the sum of two effects
-(i.e. \lstinline[style=oxcaml]{Term(Sum(L)(R)).t}) are isomorphic to a
-recursive datatype that is built from the terms of one of the effects
-and the operations of the other~\cite{hyland2006combining}. We can define
-this type and associated isomorphism as:
-\begin{lstlisting}[style=oxcaml]
-  module Half_term (L : Op) (R : Op) : sig
-    type 'a t =
-      | Return : 'a @@ global -> 'a t
-      | Op : 'r R.t @@ global * ('r -> Handler(L).t @ local -> 'a t) -> 'a t
-  end
-
-  module Half_reflection (L : Op) (R : Op) : sig
-    val reify :
-      (Handler(L).t @ local -> Handler(R).t @ local -> 'a)
-      -> Handler(L).t @ local -> 'a Half_term(L)(R).t
-
-    val reflect :
-      (Handler(L).t @ local -> 'a Half_term(L)(R).t)
-      -> Handler(L).t @ local -> Handler(R).t @ local -> 'a
-  end
-\end{lstlisting}
-
-Note that the key difference between
-\lstinline[style=oxcaml]{Half_term(L)(R).t} and
-\lstinline[style=oxcaml]{Term(R).t} is that continuing the computation
-after an operation requires passing in a local
-\lstinline[style=oxcaml]{Handler(L).t} again.
-
-Now we can use this machinery to conveniently build a handler of
-\lstinline[style=oxcaml]{Gen} that also forwards operations from
-\lstinline[style=oxcaml]{Await}:
-\begin{lstlisting}[style=oxcaml]
-  module Await_gen_eff = Half_reflection(Await)(Gen)
-  open Half_term(Await)(Gen)
+We can now write our desired handler for the
+\lstinline[style=oxcaml]{Gen} effect by instead writing a handler for
+\lstinline[style=oxcaml]{Sum(Await)(Gen)} that forwards all the
+\lstinline[style=oxcaml]{Await} operations on to another handler.
+
+\begin{lstlisting}[style=oxcaml]
+  module Eff = Reflection(Sum(Await)(Gen))
+  module Proj = Project(Await)(Gen)
 
@@
   let handle_gen_in_await f ah =
-      let rec loop = function
+    let rec loop ah = function
       | Return () -> Finished
-      | Op(Gen i, k) -> More(i, fun ah -> loop (k () ah))
+      | Op(Left op, k) -> loop ah (k (perform ah op))
+      | Op(Right(Gen i), k) -> More(i, fun ah -> loop ah (k ()))
     in
-    loop (Await_gen_eff.reify f ah)
-\end{lstlisting}
-where \lstinline[style=oxcaml]{handle_gen_in_await} has the type:
-\begin{lstlisting}[style=oxcaml]
-  val handle_gen_in_await :
-    (Handler(Await).t @ local -> Handler(Gen).t @ local -> unit)
-    -> Handler(Await).t @ local
-    -> aints
-\end{lstlisting}
-
-The literature on lexical effect handlers often ignores this issue. For
-example, Xie et al.~\cite{xie2020effect} proposed a system which only
-supported directly nesting effect handlers and had no mechanism for
-writing a composable handler for generators that could be moved between
-different underlying asynchronous I/O handlers.
-
-This is because composition requiring forwarding is much less common
-than nesting and it fundamentally requires each use of the
-\lstinline[style=oxcaml]{Await} operation to perform work proportional
-to the number of handlers it is forwarded through -- often referred to
-as a ``linear search''. There has been a tendency to treat this linear
-search as some form of deficiency of algebraic effect handlers, but it
-is simply a classic trade-off between the extra expressivity allowed by
-an indirection and the runtime cost of traversing that indirection. With
-our effect reflection API we are able to use nesting for the cases that
-it supports -- avoiding the cost of the indirection -- and fall back to
-forwarding for the more general cases.
+    loop ah (Eff.reify (fun h -> f (Proj.outl h) (Proj.outr h))
+\end{lstlisting}
+
+The need for effect forwarding is a general short-coming with lexical
+handlers that applies to all such systems. Examples like this
+fundamentally require that performing an \lstinline[style=oxcaml]{Await}
+operation to does work proportional to the number of handlers it is
+forwarded through. Lexical handlers do a constant amount of work to
+perform operations and so cannot handle such examples
+directly. Non-lexical effect handlers *always* do work proportional to
+the number of nested handlers, allowing them to support such examples
+directly but paying a higher cost for the cases that do not need
+forwarding.
 

### 2025-09-17 paper.tex
Tighten up some text in Section 2

@@
 
-We can provide effect reflection as a simple library API in OxCaml. In
-this section we describe this API and show how to use it to write a
-number of typical examples of algebraic effects.
+We provide effect reflection as a simple library API in OxCaml. In this
+section we describe this API and show how to use it to write typical
+examples of algebraic effects.
 
@@
 The effect reflection API relies on OxCaml's support for tracking
-locality in types, so let's start with an overview of that support.
+locality in types, which we describe here. For a full treatment see
+Lorenzen et al.~\cite{lorenzen2024oxidizing}.
 
@@
 within a function.  It ends the function's region early and then
-executes the expression
-\lstinline[style=oxcaml]{e}. \lstinline[style=oxcaml]{e} is essentially
-executed within the region of the caller of the function and any local
-values created within it can be safely returned to that caller.
-
-The place where modes appear within the type algebra is on arrow
-types. A function being \lstinline[style=oxcaml]{global} doesn't tell you
-that the values it accepts or returns should be
-\lstinline[style=oxcaml]{global} too, so instead the arrow type is
-labelled with two modes: one for the argument and one for the return
-value.
+executes the expression \lstinline[style=oxcaml]{e}. Since
+\lstinline[style=oxcaml]{e} is executed within the region of the caller
+of the function, any local values created within it can be safely
+returned to that caller.
+
+Arrow types are labelled with two modes: one for the argument and one
+for the return value.
 \begin{lstlisting}[style=oxcaml]
@@
 sense for a value which has the ability to leave a region to contain a
-value which doesn't, as when the first value leaves a region it will
-implicitly cause the second one to leave that region as well. This means
-the \lstinline[style=oxcaml]{global} mode is always deep.
+value which doesn't have that ability: when the first value leaves a
+region it will implicitly cause the second one to leave that region as
+well. This means the \lstinline[style=oxcaml]{global} mode is always
+deep.
 
@@
 effect. In OxCaml, such higher-kinded interfaces must be built using
-functors, so we start with a simple module type
+functors. We start with a simple module type
 \lstinline[style=oxcaml]{Op} to act as the parameter type for these
@@
 monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t}
-is the type of mappings from the operations of
-\lstinline[style=oxcaml]{O} into the operations of the current ambient
-monad. Alternatively, \lstinline[style=oxcaml]{Handler(O).t} can be
-described as representing algebraic effect handlers that are handling
-the current computation.
+represents the algebraic effect handlers that are handling the current
+computation.
 
@@
 \end{equation*}
-Due to the absence of mode polymorphism in OxCaml, we expose this
-isomorphism as three functions: \lstinline[style=oxcaml]{reify} and
-\lstinline[style=oxcaml]{reify_local} which implement the left-to-right
-direction at the \lstinline[style=oxcaml]{global} and
-\lstinline[style=oxcaml]{local} modes respectively; and
-\lstinline[style=oxcaml]{perform}, which is equivalent to the
-right-to-left direction but makes for a more convenient primitive in
-practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
-individual operations as opposed to full terms.
+We expose this isomorphism as three functions:
 \begin{lstlisting}[style=oxcaml]
@@
 \end{lstlisting}
-Using \lstinline[style=oxcaml]{perform} we can implement reflection of
-global terms as:
+\lstinline[style=oxcaml]{reify} and
+\lstinline[style=oxcaml]{reify_local} implement the left-to-right
+direction. We need two separate functions because OxCaml lacks any form
+of mode polymorphism.
+
+\lstinline[style=oxcaml]{perform} is equivalent to the right-to-left
+direction but makes for a more convenient primitive in
+practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
+individual operations as opposed to full terms. Using
+\lstinline[style=oxcaml]{perform} we can implement reflection of global
+terms as:
 \begin{lstlisting}[style=oxcaml]

### 2025-09-17 paper.tex
Add skeleton for more complete appendices

@@
 \newcommand{\embed}[1]{\mathopen{\lbag}#1\mathclose{\rbag}}
+\newcommand{\ctxeq}{\cong_{\mathrm{ctx}}}
 
@@
 definitions of normal forms for values and computations are given in
-Fig.\ref{fig:normal}. Subject reduction and strong normalization of
-these rules can be proved using standard techniques.
+Fig.\ref{fig:normal}.
 
@@
 
+We prove subject reduction for this semantics in
+Appendix~\ref{apdx:subject-reduction} by straightforward induction on
+the reduction relations.
+\begin{lemma}
+\label{lemma:subject-reduction}
+  \begin{equation*}
+    \Gamma \turnv v_1 \types \tau
+    \wedge v_1 \reducestov v_2 \implies \Gamma \turnv v_2 \types \tau
+  \end{equation*}
+\end{lemma}
+
+These reduction relations are strongly normalizing. We prove this via
+Tait computability\cite{??} in
+Appendix~\ref{apdx:strong-normalization}. This is made more involved by
+the presence of infinitary inductive types, which requires us to
+\todo{something something Knaster-Tarski}.
+\begin{lemma}
+\label{lemma:strong-normalization}
+\todo{Find simplest formal description of strong normalization}
+\end{lemma}
+
 Note that the reduction relations do not include $\eta$ expansion. They
@@
 
+\subsection{Properties of the denotational semantics}
+
+The denotational semantics is sound:
+\begin{lemma}
+\label{lemma:soundness}
+  \begin{equation*}
+    \Gamma \turnv v_1 \types \tau
+    \wedge v_1 \reducestov v_2
+    \implies \sem{\Gamma \turnv v_1 \types \tau}
+           = \sem{\Gamma \turnv v_2 \types \tau}
+  \end{equation*}
+\end{lemma}
+and adequate for contextual equality $\ctxeq$:
+\begin{lemma}
+\label{lemma:adequacy}
+  \begin{equation*}
+    \sem{\turnv v_1 \types \tau} = \sem{\turnv v_2 \types \tau}
+    \implies v_1 \ctxeq v_2
+  \end{equation*}
+\end{lemma}
+
+We present the full semantics, precisely define contextual equality, and
+prove these properties in Appendix~\ref{apdx:denotational}.
+
 \section{Implementation}
@@
 
+\appendix
+\section{Operational semantics}
+\subsection{Subject reduction}
+\label{apdx:subject-reduction}
+\subsection{Strong normalization}
+\label{apdx:strong-normalization}
+\section{Denotational semantics}
+\label{apdx:denotational}
+\subsection{Soundness}
+\subsection{Contextual equivalence}
+\subsection{Compositionality}
+\subsection{Adequacy}
 \end{document}

### 2025-10-06 paper.tex
More changes

@@
   \end{equation*}
+
   where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
@@
   \emph{freer monad} of $\Sigma$. This isomorphism emerges naturally
-  from a denotational semantics for algebraic effects in terms of
-  presheaves over bounded Lawvere theories.
+  from the \emph{ambient monad model}, a novel semantic model of
+  algebraic effect handlers in terms of presheaves over bounded Lawvere
+  theories.
 
@@
       | Return x -> x
-      | Op(Get, k) -> loop s (k s)
-      | Op(Set s', k) -> loop s' (k ())
+      | Op(Get, k) -> loop s (State_eff.continue_local k s)
+      | Op(Set s', k) -> loop s' (State_eff.continue_local k ())
     in
@@
 
-\subsection{Denotational semantics}
-
-This design emerges naturally from a denotational semantics where types
-are interpreted as presheaves over \emph{bounded Lawvere theories} -- a
-categorical framework that makes the connection between locality and
-effects precise.
+\subsection{The ambient monad model}
+
+This design emerges naturally from a semantic model of algenbraic
+effects where types are interpreted as presheaves over \emph{bounded
+  Lawvere theories} -- a categorical framework that makes the connection
+between locality and effects precise.
 
@@
 \begin{itemize}
-\item The ambient monad can be defined as a monad $\mon$ on such
+\item Computations can be represented as a monad $\mon$ on such
   presheaves, which maps each algebraic theory to its free algebras.
@@
 \item The effect reflection isomorphism becomes a simple consequence of
-  the Yoneda lemma and some properties of the sums of algebraic
-  theories.
+  the Yoneda lemma and a key property of the sums of algebraic theories.
 \end{itemize}
@@
 \begin{lstlisting}[style=oxcaml]
+  module Continuation(O : Op) : sig
+    type ('a, 'b) t
+  end
+
   module Term (O : Op) : sig
@@
       | Return : 'a @@ global -> 'a t
-      | Op : 'r O.t @@ global * ('r -> 'a t) -> 'a t
+      | Op : 'r O.t @@ global * ('r, 'a) Continuation(O).t -> 'a t
   end
@@
 \end{lstlisting}
-These types directly implement the types from our core isomorphism:
+These types directly implement the types from our core isomorphism.
 \lstinline[style=oxcaml]{'a Term(Sigma).t} corresponds to
-$\ind{\Sigma}(\glob \tau)$, while
-\lstinline[style=oxcaml]{Handler(Sigma).t} corresponds to
-$\yoneda{\Sigma}$
-
-For an effect signature \lstinline[style=oxcaml]{O},
-\lstinline[style=oxcaml]{'a Term(O).t} is the type of terms in the free
-algebra of the corresponding effect -- otherwise known as the freer
-monad~\cite{kiselyov2015freer}. \lstinline[style=oxcaml]{Handler(O).t}
-represents the algebraic effect handlers that are handling the current
+$\ind{\Sigma}(\glob \tau)$ and represents terms in the free algebra of
+\lstinline[style=oxcaml]{Sigma}. \lstinline[style=oxcaml]{Handler(Sigma).t}
+corresponds to $\yoneda{\Sigma}$ and represents algebraic effect
+handlers handling \lstinline[style=oxcaml]{Sigma} in the current
 computation.
+
+\lstinline[style=oxcaml]{('a, 'b) Continuation(Sigma).t} is equivalent
+to \lstinline[style=oxcaml]{'a -> 'b Term(Sigma).t} when global and
+equivalent to \lstinline[style=oxcaml]{'a -> 'b Term(Sigma).t @ local}
+when local. This ensures that the global modality distributes over
+\lstinline[style=oxcaml]{Term(Sigma).t}, which is an important property
+of $\ind{\Sigma}$. The need for it is an artifact of OxCaml's approach
+to locality: the global modality does not distribute over the function
+arrow in OxCaml. In the language we describe in
+Section~\ref{sec:language} it is replaced with an ordinary function
+type.
 
@@
 
+    val continue : ('a, 'b) Continuation(O).t -> 'a -> 'b Term(O).t
+
     val reify_local :
@@
 
+    val continue_local :
+      ('a, 'b) Continuation(O).t @ local -> 'a -> 'b Term(O).t @ local
+
     val perform : 'a O.t -> (Handler(O).t @ local -> 'a)
@@
 \lstinline[style=oxcaml]{reify_local} implement the left-to-right
-direction. We need two separate functions because OxCaml lacks any form
-of mode polymorphism.
+direction of the isomorphism. We need two separate functions because
+OxCaml lacks any form of mode polymorphism.
 
 \lstinline[style=oxcaml]{perform} is equivalent to the right-to-left
-direction but makes for a more convenient primitive in
-practice. \lstinline[style=oxcaml]{perform} amounts to reflection of
-individual operations as opposed to full terms. Using
-\lstinline[style=oxcaml]{perform} we can implement reflection of global
-terms as:
+direction of the isomorphism but is a more convenient primitive in
+practice. It amounts to reflection of individual operations as opposed
+to full terms. We can implement reflection of global full terms as:
 \begin{lstlisting}[style=oxcaml]
@@
     | Return x -> x
-    | Op(op, k) -> reflect (k (perform op h)) h
+    | Op(op, k) -> reflect (continue_local k (perform op h)) h
 \end{lstlisting}
@@
       | Return x -> x
-      | Op(Get, k) -> loop s (k s)
-      | Op(Set s', k) -> loop s' (k ())
+      | Op(Get, k) -> loop s (State_eff.continue_local k s)
+      | Op(Set s', k) -> loop s' (State_eff.continue_localk ())
     in
@@
 \end{lstlisting}
-Note that we are able to use \lstinline[style=oxcaml]{reify_local} to allow
-the function passed to \lstinline[style=oxcaml]{handle_state} to be local.
+Note that we are able to use \lstinline[style=oxcaml]{reify_local} to
+allow the function passed to \lstinline[style=oxcaml]{handle_state} to
+be local.
 
 This shows how to implement state using effect reflection, but note that
-the same interface can also be implemented directly using mutable state
-and locality:
+the same interface can also be implemented directly in OxCaml using
+mutable state and locality:
 \begin{lstlisting}[style=oxcaml]
@@
 effect reflection. Crucially it is just as easy to reason about, which
-in turn means it is just as easy to reason about as the state monad. In
-some sense, the improved reasoning ability of the state monad over
-ordinary mutable state is precisely captured by the notion of locality.
+in turn means it is just as easy to reason about as the state
+monad. This shows that the improved reasoning ability of the state monad
+over ordinary mutable state is precisely captured by the notion of
+locality.
 
 This transformation from effect reflection to local mutable state works
-for any \emph{tail-resumptive} effect handler.
+for any \emph{tail-resumptive}\cite{??} effect handler.
 
@@
       | Return () -> Finished
-      | Op(Gen i, k) -> More(i, fun () -> loop (k ()))
+      | Op(Gen i, k) -> More(i, fun () -> loop (Gen_eff.continue k ()))
     in
     loop (Gen_eff.reify f)
-
 \end{lstlisting}
@@
       | Op(Await d, k) ->
-          Deferred.bind d (fun x -> loop (k x))
+          Deferred.bind d (fun x -> loop (Await_eff.continue k x))
     in
@@
 
-A key feature of algebraic effects is how easy they are to compose. We
-can use state whilst using asynchronous I/O by simply nesting their
+A key feature of algebraic effects is how easily they compose. We can
+use state whilst using asynchronous I/O by simply nesting their
 handlers:
@@
 
-We could change the definition of \lstinline[style=oxcaml]{handle_gen} to
-use \lstinline[style=oxcaml]{reify_local} instead of
-\lstinline[style=oxcaml]{reify}, but then we would lose the property that
-the generator we produce can be passed around freely. Instead we need to
-build a handler that can forward the \lstinline[style=oxcaml]{Await}
-operation to different effect handlers at different points in the
-program's execution.
+We could change the definition of \lstinline[style=oxcaml]{handle_gen}
+to use \lstinline[style=oxcaml]{reify_local} instead of
+\lstinline[style=oxcaml]{reify}, but then we would lose the property
+that the generator we produce can be passed around freely. Instead we
+need to build a handler that can \emph{forward} the
+\lstinline[style=oxcaml]{Await} operation to different effect handlers
+at different points in the program's execution.
 
@@
 
-Composing handlers, like \lstinline[style=oxcaml]{handle_gen}, that use
+Composing handlers like \lstinline[style=oxcaml]{handle_gen}, that use
 \lstinline[style=oxcaml]{reify} rather than
-\lstinline[style=oxcaml]{reify_local} is a fundamentally more involved
+\lstinline[style=oxcaml]{reify_local}, is a fundamentally more involved
 operation. Consider the case of generators that are able to use
@@
       | Return () -> Finished
-      | Op(Left op, k) -> loop ah (k (perform ah op))
-      | Op(Right(Gen i), k) -> More(i, fun ah -> loop ah (k ()))
+      | Op(Left op, k) -> loop ah (Eff.continue k (perform ah op))
+      | Op(Right(Gen i), k) ->
+          More(i, fun ah -> loop ah (Eff.continue k ()))
     in
@@
 
-The need for effect forwarding is a general short-coming with lexical
+The need for effect forwarding is a general short-coming of lexical
 handlers that applies to all such systems. Examples like this
 fundamentally require that performing an \lstinline[style=oxcaml]{Await}
-operation to does work proportional to the number of handlers it is
+operation does work proportional to the number of handlers it is
 forwarded through. Lexical handlers do a constant amount of work to
@@
 \grun{\return{v}} &\reducestov v \\
-\letv{x}{\return{v}}{c} &\reducestov \sbst{c}{x}{v} \\
-\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducestov \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
+\letv{x}{\return{v}}{c} &\reducestoc \sbst{c}{x}{v} \\
+\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducestoc \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
 \app{(\abs{x}{\tau}{c})}{v} &\reducestoc \sbst{c}{x}{v} \\
@@
 
-\section{Presheaves for algebraic effects}
+\section{The ambient monad model}
 \label{sec:semantics}
 
-The design of effect reflection falls out naturally from a simple
-denotational semantics built on presheaves. This semantic foundation not
-only validates our approach but also makes the connection between
-locality and effects precise. The key idea is to index the meaning of
-types by an ambient algebraic monad of effects: rather than have the
-semantics of each type be a set of elements, the semantics of each type
-is a mapping from the ambient monad to a set of elements.
+The design of effect reflection falls out naturally from a novel
+semantic model built on presheaves. This semantic foundation not only
+validates our approach but also makes the connection between locality
+and effects precise. The key idea is to index the meaning of types by an
+ambient algebraic monad of effects: rather than have the semantics of
+each type be a set of elements, the semantics of each type is a mapping
+from the ambient monad to a set of elements. We give a denotational
+semantics to the language from Section~\ref{sec:language} using this
+model.
 
@@
 \end{equation*}
-as the basis of our semantics. The $\op$ in the definition of presheaves
-and the $\op$ due to our dependence being contravariant cancel out, but
-it is still best to think in terms of presheaves as that is a more
-precise description of what we are doing.
-
-Our semantics will use objects from $\pshlawcop$ to represent types and
-contexts, and morphisms in $\pshlawcop$ to represent values. This allows
-their meanings to be parameterized by the ambient monad. By using
-presheaves rather than ordinary functions we ensure terms and types at a
-particular ambient monad can always be mapped naturally to a larger
-ambient monad.
+as the basis of the ambient monad model. The $\op$ in the definition of
+presheaves and the $\op$ due to our dependence being contravariant
+cancel out, but it is still best to think in terms of presheaves as that
+is a more precise description of what we are doing.
+
+The ambient monad model uses objects from $\pshlawcop$ to represent
+types and contexts, and morphisms in $\pshlawcop$ to represent
+values. This allows their meanings to be parameterized by the ambient
+monad. By using presheaves rather than ordinary functions we ensure
+terms and types at a particular ambient monad can always be mapped
+naturally to a larger ambient monad.
 
@@
 \end{equation*}
-We call $\mon$ the \emph{free algebras monad}, and we will use it as the
-computation monad in our denotational semantics.
+We call $\mon$ the \emph{free algebras monad}, and it is the
+representation of computations in the ambient monad model.
 \begin{align*}
@@
 
-Within our semantics, this amounts to restricting a type to only the
-elements which do not depend on the ambient monad. This is very similar
-to the purity comonad of Choudhury and
+Within the abient monad model, this amounts to restricting a type to
+only the elements which do not depend on the ambient monad. This is very
+similar to the purity comonad of Choudhury and
 Krishnaswami~\cite{choudhury2020recovering}. In particular,
@@
 
-We use these initial algebras as the denotational semantics of
-$\ind{\Sigma}$:
+We use these initial algebras as the semantics of $\ind{\Sigma}$:
 \begin{mathpar}
@@
 \label{apdx:subject-reduction}
+
+We prove subject reduction (Lemma~\ref{lemma:subject-reduction}):
+\begin{equation*}
+  \Gamma \turnv v_1 \types \tau
+  \wedge v_1 \reducestov v_2 \implies \Gamma \turnv v_2 \types \tau
+\end{equation*}
+by mutual induction over the two reduction relations. The reduction
+relations are the least reflexive, transitive congruences closed under
+the rules in Fig.~\ref{fig:reduction}.
+
+The reflexivity case is immediate and the transitive case follows
+directly by induction. As all the typing rules are entirely
+compositional, the congruence cases also all follow direectly by
+induction. This leaves only the rules from Fig.~\ref{fig:reduction}.
+
+Note that the typing rules are syntax directed with the exception of the
+two var rules, which are distinguished by how the variable is bound in
+the context. Since none of the left-hand sides of our reduction rules is
+a variable, there is always only a single typing rule that we need to
+consider.
+
+The cases for the products rules:
+\begin{align*}
+\fst{(v_1, v_2)} &\reducestov v_1 \\
+\snd{(v_1, v_2)} &\reducestov v_2 \\
+\end{align*}
+follow immediately by inversion on the typing rule.
+
+\begin{equation*}
+\gunbox{x}{\gbox{v_1}}{v_2} \reducestov \sbst{v_2}{x}{v_1}
+\end{equation*}
+\todo{Requires substitution rule of a term under $\cdiv{\Gamma}$ for a
+  boxed binding}.
+
+The case for the $\grun$ rule
+\begin{equation*}
+\grun{\return{v}} \reducestov v
+\end{equation*}
+\todo{Requires weakening from $\cdiv{\Gamma}$ to $\Gamma$}.
+
+The case for the let-return rule:
+\begin{equation*}
+\letv{x}{\return{v}}{c} \reducestoc \sbst{c}{x}{v}
+\end{equation*}
+\todo{Requires ordinary substitution lemma}.
+
+The case for the let-perform rule:
+\begin{equation*}
+\letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} \reducestoc \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}}
+\end{equation*}
+follows immediately by inversion on the typing rule.
+
+The case for the standard beta rule:
+\begin{equation*}
+\app{(\abs{x}{\tau}{c})}{v} \reducestoc \sbst{c}{x}{v}
+\end{equation*}
+\todo{Requires ordinary substitution lemma}.
+
+The cases for the case rules:
+\begin{align*}
+\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_1}{x_1}{v}
+\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_2}{x_2}{v}
+\end{align*}
+\todo{Requires ordinary substitution lemma}.
+
+The case for the $\retname$ rule:
+\begin{equation*}
+\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} \reducestoc \sbst{c}{x}{v}
+\end{equation*}
+\todo{Requires ordinary substitution lemma}.
+
+The case for the $\indintro{\Sigma}{op}{\_}{\_}$ rule:
+\begin{align*}
+&\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
+&\qquad \reducestoc \sbst{\sbst{c}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}}
+\end{align*}
+\todo{Requires ordinary substitution lemma}.
+
+The case for the return rule:
+\begin{equation*}
+\reify{\Sigma}{x}{\return{v}} \reducestoc \return{\ret{\Sigma}{v}}
+\end{equation*}
+\todo{Requires strengthening by local bindings of handlers for global values,
+  which doesn't actually hold because you can have global terms that ignore their
+  inputs}.
+
+\begin{equation*}
+\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} \reducestoc \return{\indintro{\Sigma}{op}{v}{\abs{x_2}{\reify{\Sigma}{x_1}{c}}}}
+\end{equation*}
+
+\begin{equation*}
+\reify{\Sigma}{x_1}{\perform{x_2}{op}{v}{x_3}{c}} \reducestoc \perform{x_2}{op}{v}{x_3}{\reify{\Sigma}{x_1}{c}} \qquad (x_1 \neq x_2)
+\end{equation*}
+
+\begin{equation*}
+\perform{(\outl{v_1})}{op}{v_2}{x}{c} \reducestoc \perform{v_1}{\mathrm{left}(op)}{v_2}{x}{c}
+\end{equation*}
+
+\begin{equation*}
+\perform{(\outr{v_1})}{op}{v_2}{x}{c} \reducestoc \perform{v_1}{\mathrm{right}(op)}{v_2}{x}{c}
+\end{equation*}
+
 \subsection{Strong normalization}

### 2025-12-16 paper.tex
Make some adjustments to the language

@@
 \newcommand{\ind}[1]{W_{#1}}
-\newcommand{\indintro}[4]{#2_{#1}(#3,\, #4)}
+\newcommand{\indintro}[4]{{#2}_{#1}(#3,\, #4)}
 \newcommand{\retname}{\mathrm{ret}}
 \newcommand{\ret}[2]{\mathop{\retname_{#1}}(#2)}
-\newcommand{\indelim}[3]{\mathop{\mathrm{rec}_{#1}} #2 \,\{#3\}}
+\newcommand{\indelimname}{\mathrm{rec}}
+\newcommand{\indelim}[3]{\mathop{\indelimname_{#1}} #2 \,\{#3\}}
+\newcommand{\indelimboxname}{\indelimname^{\glob}}
+\newcommand{\indelimbox}[3]{\mathop{\indelimboxname_{#1}} #2 \,\{#3\}}
+\newcommand{\indelimmname}{\indelimname^{m}}
+\newcommand{\indelimm}[3]{\mathop{\indelimmname_{#1}} #2 \,\{#3\}}
 \newcommand{\indretcase}[2]{\retname(#1) \Rightarrow #2}
 \newcommand{\indcase}[4]{#1(#2, #3) \Rightarrow #4}
+\newcommand{\letret}[4]{\mathop{\mathrm{let}} (\ret{#1}{#2}) = #3 \mathop{\mathrm{in}} #4}
 \newcommand{\seq}{\,;\,}
@@
 \newcommand{\inr}[1]{\mathop{\mathrm{in}_R} #1}
-\newcommand{\case}[5]{\mathop{\mathrm{case}} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
+\newcommand{\casename}{\mathrm{case}}
+\newcommand{\case}[5]{\mathop{\casename} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
+\newcommand{\casemname}{\casename^{m}}
+\newcommand{\casem}[5]{\mathop{\casemname} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
+\newcommand{\caseboxname}{\casename^{\glob}}
+\newcommand{\casebox}[5]{\mathop{\caseboxname} #1 \,\{\,\inl{#2} \Rightarrow #3 \seq \inr{#4} \Rightarrow #5 \,\}}
 \newcommand{\abs}[3]{\mathop{\lambda}(#1 \types #2).\,#3}
@@
 \newcommand{\types}{\mathrel{:}}
+\newcommand{\mtypes}{\mathrel{:_{m}}}
 \newcommand{\gtypes}{\mathrel{:_{\Box}}}
@@
 \newcommand{\lbind}[3]{\ccons{#1}{#2\types#3}}
+\newcommand{\mbind}[3]{\ccons{#1}{#2\mtypes#3}}
 \newcommand{\gbind}[3]{\ccons{#1}{#2\gtypes#3}}
 \newcommand{\cdiv}[1]{#1/\Box}
+\newcommand{\cmdiv}[1]{#1/m}
+\newcommand{\cldiv}[1]{#1/\text{\textvisiblespace}}}
 \newcommand{\bdiv}[1]{\cdiv{\mathop{!_{#1}}}}
@@
 \newcommand{\gbox}[1]{\mathop{\mathrm{box}} #1}
-\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox #1) = #2 \mathop{\mathrm{in}} #3}
+\newcommand{\gunbox}[3]{\mathop{\mathrm{let}} (\gbox{#1}) = #2 \mathop{\mathrm{in}} #3}
 \newcommand{\grun}[1]{\mathop{\mathrm{run}}(#1)}
@@
   \begin{equation*}
-    \ind{\Sigma}(\glob \tau)
+    \ind{\Sigma}(\tau)
     \qquad \cong \qquad
@@
 
-  where $\ind{\Sigma}(\glob \tau)$ is the free algebra of $\Sigma$ at
-  $\glob \tau$, $\yoneda{\Sigma}$ is a type representing how to inject
+  where $\ind{\Sigma}(\tau)$ is the free algebra of $\Sigma$ at
+  $\tau$, $\yoneda{\Sigma}$ is a type representing how to inject
   terms of $\Sigma$ into the current ambient monad, and $\glob$ is a
@@
 \begin{equation*}
-  \ind{\Sigma}(\glob \tau)
+  \ind{\Sigma}(\tau)
   \qquad \cong \qquad
@@
 \end{equation*}
-where $\ind{\Sigma}(\glob \tau)$ is an inductive type representing the
-free algebra of $\Sigma$ at $\glob \tau$, and $\yoneda{\Sigma}$ is a
+where $\ind{\Sigma}(\tau)$ is an inductive type representing the
+free algebra of $\Sigma$ at $\tau$, and $\yoneda{\Sigma}$ is a
 type representing how to inject terms of $\Sigma$ into the current
@@
 \lstinline[style=oxcaml]{'a Term(Sigma).t} corresponds to
-$\ind{\Sigma}(\glob \tau)$ and represents terms in the free algebra of
+$\ind{\Sigma}(\tau)$ and represents terms in the free algebra of
 \lstinline[style=oxcaml]{Sigma}. \lstinline[style=oxcaml]{Handler(Sigma).t}
@@
 perform operations and so cannot handle such examples
-directly. Non-lexical effect handlers *always* do work proportional to
-the number of nested handlers, allowing them to support such examples
-directly but paying a higher cost for the cases that do not need
-forwarding.
+directly. Non-lexical effect handlers \textbf{always} do work
+proportional to the number of nested handlers, allowing them to support
+such examples directly but paying a higher cost for the cases that do
+not need forwarding.
 
@@
                               \mid \tau \rightarrow \tau \mid \ldots \\
-  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \lbind{\Gamma}{x}{\tau} \mid \ldots \\
+  \mathrm{contexts}, \Gamma ::=\;& \cempty \mid \mbind{\Gamma}{x}{\tau} \mid \ldots \\
   \mathrm{values}, v :=\;& x \mid (v, v) \mid \fst v \mid \snd v
-                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c} \mid \ldots \\
+                           \mid () \mid \inl{v} \mid \inr{v} \mid \abs{x}{\tau}{c}
+                           \mid \case{v}{x}{v}{x}{v} \ldots \\
   \mathrm{computations}, c :=\;& \return{v} \mid \letv{x}{c}{c}
-                                 \mid \app{v}{v} \mid \\
-                            & \case{v}{x}{c}{x}{c} \mid \ldots
+                                 \mid \app{v}{v} \mid \ldots
 \end{align*}
@@
 \and
+\inferrule{\Gamma \turnv v_1 \types \tau_1 + \tau_2 \\
+           \lbind{\Gamma}{x_1}{\tau_1} \turnv v_2 \types \tau_3 \\
+           \lbind{\Gamma}{x_2}{\tau_2} \turnv v_3 \types \tau_3}
+{\Gamma \turnv \case{v_1}{x_1}{v_2}{x_2}{v_3} \types \tau_3}
+\and
 \inferrule{\lbind{\Gamma}{x}{\tau_1} \turnc c \types \tau_2}
@@
 {\Gamma \turnc v_1 v_2 \types \tau_2}
-\and
-\inferrule{\Gamma \turnv v \types \tau_1 + \tau_2 \\
-           \lbind{\Gamma}{x_1}{\tau_1} \turnc c_1 \types \tau_3 \\
-           \lbind{\Gamma}{x_2}{\tau_2} \turnc c_2 \types \tau_3}
-{\Gamma \turnc \case{v}{x_1}{c_1}{x_2}{c_2} \types \tau_3}
 \end{mathpar}
@@
 \end{figure}
+
+Note that $\casename$ -- the eliminator for sum types -- is a value
+rather than a computation. We can define a computation version as:
+\begin{align*}
+    &\case{v}{x_1}{c_1}{x_2}{c_2} := \\
+    &\qquad (\case{v}{x_1}{\abs{\_}{1}{c_1}}{x_2}{\abs{\_}{1}{c_2}}) \; ()
+\end{align*}
 
@@
 
+The global comonad distributes over sum types:
+\begin{equation*}
+  \glob (\tau_1 + \tau_2)
+  \qquad \cong \qquad
+  (\glob \tau_1) + (\glob \tau_2)
+\end{equation*}
+which we realize via $\caseboxname$, a global form of the
+$\casename$ construct, which allows you to match on a global sum and get
+out global components.
+
+We also add a modalities category, $m$, to the syntax, which is either
+blank space or $\glob$, to make it easier to describe rules that are
+parametric in the presence or absence of the global modality. As part of
+this, the $\cdiv{\Gamma}$ operation is extended to work on modalities
+$\cmdiv{\Gamma}$ with $\cldiv{\Gamma} = \Gamma$.
+
 The syntax and typing rules related to locality are in
@@
   \mathrm{contexts}, \Gamma ::=\;& \ldots \mid \gbind{\Gamma}{x}{\tau} \\
-  \mathrm{values}, v :=\;& \ldots \mid \gbox{v} \mid \gunbox{x}{v}{v}
-                           \mid \grun{c} \mid \ldots
+  \mathrm{values}, v :=\;& \ldots \mid \gbox{v} \mid \gunbox{x}{v}{v} \mid
+                           \casebox{v}{x}{v}{x}{v} \mid \grun{c} \mid \ldots \\
+  \mathrm{modalities}, m ::=\;& \text{\textvisiblespace} \mid \glob
 \end{align*}
@@
 \and
+\inferrule{\cdiv{\Gamma} \turnv v_1 \types \tau_1 + \tau_2 \\
+           \gbind{\Gamma}{x_1}{\tau_1} \turnv v_2 \types \tau_3 \\
+           \gbind{\Gamma}{x_2}{\tau_2} \turnv v_3 \types \tau_3}
+{\Gamma \turnv \casebox{v_1}{x_1}{v_2}{x_2}{v_3} \types \tau_3}
+\and
 \inferrule{\cdiv{\Gamma} \turnc c \types \tau}
@@
 $v_k \types \glob \tau_a \rightarrow \ind{\Sigma}(\tau)$. $\ind{\Sigma}(\tau)$
-also has a constructor $\ret{\Sigma}{v}$ where $v \types \tau$.
+also has a constructor $\ret{\Sigma}{v}$ where $v \types \glob \tau$.
 
@@
 \begin{align*}
-\indelim{\Sigma}{v}
-  {&\indretcase{x}{c_r} \seq \\
-   &\indcase{op_1}{x_{p_1}}{x_{k_1}}{c_1} \seq \\
+\indelim{\Sigma}{v_s}
+  {&\indretcase{x}{v_r} \seq \\
+   &\indcase{op_1}{x_{p_1}}{x_{k_1}}{v_1} \seq \\
    &\ldots \seq \\
-   &\indcase{op_n}{x_{p_n}}{x_{k_n}}{c_n}}
+   &\indcase{op_n}{x_{p_n}}{x_{k_n}}{v_n}}
 \end{align*}
@@
 Fig.\ref{fig:typing-inductive}.
+
+As with $\casename$, $\indelimname$ is defined as a value and we can
+define a computation version in terms of it.
+
+The global comonad distributes over inductive types, which we realize
+via $\indelimboxname$, a global form of the $\indelimname$ construct,
+which allows you to match on a global inductive type and get out global
+components.
+
+For convenience, we also define the following abbreviation:
+\begin{align*}
+    \letret{\Sigma}{x}{c_1}{c_2} := \quad &\letv{x_s}{c_1}{\\
+        &\indelim{\Sigma}{x_s}{ \\
+        &\qquad \indretcase{x}{c_2} \seq \\
+        &\qquad \indcase{op_1}{x_{p_1}}{x_{k_1}}{\return{\indintro{\Sigma}{op_1}{x_{p_1}}{x_{k_1}}}} \seq \\
+        &\qquad \ldots \seq \\
+        &\qquad \indcase{op_n}{x_{p_n}}{x_{k_n}}{\return{\indintro{\Sigma}{op_n}{x_{p_n}}{x_{k_n}}}}}}
+\end{align*}
 
@@
   \mathrm{values}, v :=\;& \ldots \mid \indintro{\Sigma}{op}{v}{v}
-                           \mid \ret{\Sigma}{v} \mid \ldots \\
-  \mathrm{computations}, c :=\;& \ldots \mid \indelim{\Sigma}{v}
-                                 {\indretcase{x}{c} \seq
-                                 \indcase{op}{x}{x}{c} \seq \ldots
-                                 \seq \indcase{op}{x}{x}{c}} \mid \ldots
+                           \mid \ret{\Sigma}{v} \mid \\
+                         & \indelimm{\Sigma}{v}
+                           {\indretcase{x}{v} \seq
+                           \indcase{op}{x}{x}{v} \seq \ldots
+                           \seq \indcase{op}{x}{x}{v}} \ldots
 \end{align*}
@@
 \inferrule{op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
-           \Gamma \turnv v_1 \types \glob \tau_1 \\
+           \cdiv{\Gamma} \turnv v_1 \types \tau_1 \\
            \Gamma \turnv v_2 \types \glob \tau_2 \rightarrow \ind{\Sigma}(\tau)}
@@
 \and
-\inferrule{\Gamma \turnv v \types \tau}
+\inferrule{\cdiv{\Gamma} \turnv v \types \tau}
 {\Gamma \turnv \ret{\Sigma}{v} \types \ind{\Sigma}(\tau)}
 \and
-\inferrule{\Gamma \turnv v \types \ind{\Sigma}(\tau_1) \\
+\inferrule{\cmdiv{\Gamma} \turnv v_s \types \ind{\Sigma}(\tau_1) \\
            \Sigma = \{ op_1 \types \tau_{p_1} \rightarrowtail \tau_{a_1};
              \ldots; op_n\types \tau_{p_n} \rightarrowtail \tau_{a_n} \} \\
-           \lbind{\Gamma}{x_r}{\tau_1} \turnc c_r \types \tau_2 \\
-           \lbind{\lbind{\Gamma}{x_{p_1}}{\glob \tau_{p_1}}}{x_{k_1}}{\glob \tau_{a_1} \rightarrow \tau_2} \turnc c_1 \types \tau_2 \\
+           \gbind{\Gamma}{x_r}{\tau_1} \turnv v_r \types \tau_2 \\
+           \mbind{\gbind{\Gamma}{x_{p_1}}{\tau_{p_1}}}{x_{k_1}}{\glob \tau_{a_1} \rightarrow \tau_2} \turnv v_1 \types \tau_2 \\
            \ldots \\
-           \lbind{\lbind{\Gamma}{x_{p_n}}{\glob \tau_{p_n}}}{x_{k_n}}{\glob \tau_{a_n} \rightarrow \tau_2} \turnc c_n \types \tau_2}
-{\Gamma \turnc \indelim{\Sigma}{v}{\indretcase{x_r}{c_r} \seq
-                                   \indcase{op_1}{x_{p_1}}{x_{k_1}}{c_1} \seq \ldots \seq
-                                   \indcase{op_n}{x_{p_n}}{x_{k_n}}{c_n}} \types \tau_2}
+           \mbind{\gbind{\Gamma}{x_{p_n}}{\tau_{p_n}}}{x_{k_n}}{\glob \tau_{a_n} \rightarrow \tau_2} \turnv v_n \types \tau_2}
+{\Gamma \turnv \indelimm{\Sigma}{v_s}{\indretcase{x_r}{v_r} \seq
+                                   \indcase{op_1}{x_{p_1}}{x_{k_1}}{v_1} \seq \ldots \seq
+                                   \indcase{op_n}{x_{p_n}}{x_{k_n}}{v_n}} \types \tau_2}
 \end{mathpar}
@@
 
+\todo{The above is missing boxing operations}
+
 \subsection{Effect reflection}
@@
 ambient monad. Effect reflection appears in the language as two
-constructs: reflection and reification. Reflection
-$\performraw{v}{op}$ uses a value of type $\yoneda{\Sigma}$ to map
-the operation $op$ from $\Sigma$ into the ambient monad.  Reification
-$\reifyraw{\Sigma}$ extends the ambient monad with the operations from
-$\Sigma$, providing a value of type $\yoneda{\Sigma}$ for constructing
-those operations via reflection.  These constructs implement the core
-isomorphism of effect reflection: reification can convert functions
-$\yoneda{\Sigma} \rightarrow \glob \tau$ into terms $\ind{\Sigma}(\glob
-\tau)$, while reflection provides the inverse direction. Their syntax
-and typing rules are shown in Fig.\ref{fig:typing-reflection}.
+constructs: reflection and reification. Reflection $\performraw{v}{op}$
+uses a value of type $\yoneda{\Sigma}$ to map the operation $op$ from
+$\Sigma$ into the ambient monad.  Reification $\reifyraw{\Sigma}$
+extends the ambient monad with the operations from $\Sigma$, providing a
+value of type $\yoneda{\Sigma}$ for constructing those operations via
+reflection.  These constructs implement the core isomorphism of effect
+reflection: reification can convert functions
+$\yoneda{\Sigma} \rightarrow \glob \tau$ into terms
+$\ind{\Sigma}(\tau)$, while reflection provides the inverse
+direction. Their syntax and typing rules are shown in
+Fig.\ref{fig:typing-reflection}.
 
@@
            op \types \tau_1 \rightarrowtail \tau_2 \in \Sigma \\
-           \Gamma \turnv v_2 \types \glob \tau_1 \\
+           \cdiv{\Gamma} \turnv v_2 \types \tau_1 \\
            \lbind{\Gamma}{x}{\glob \tau_2} \turnc c \types \tau_3}
@@
 \inferrule{\lbind{\Gamma}{x}{\yoneda{\Sigma}} \turnc c \types \glob \tau}
-{\Gamma \turnc \; \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\glob \tau)}
+{\Gamma \turnc \; \reify{\Sigma}{x}{c} \types \ind{\Sigma}(\tau)}
 \end{mathpar}
@@
 \inferrule{\Gamma \turnv v \types \yoneda{\Sigma} \\
-           \Gamma \turnv c \types \ind{\Sigma}(\glob \tau)}
+           \Gamma \turnv c \types \ind{\Sigma}(\tau)}
 {\Gamma \turnc \; \reflect{\Sigma}{v}{c} \types \glob \tau}
@@
 \snd{(v_1, v_2)} &\reducestov v_2 \\
+\casem{(\inl{v_1})}{x_1}{v_2}{x_2}{v_3} &\reducestov \sbst{v_2}{x_1}{v_1} \\
+\casem{(\inr{v_1})}{x_1}{v_2}{x_2}{v_3} &\reducestov \sbst{v_3}{x_2}{v_1} \\
 \gunbox{x}{\gbox{v_1}}{v_2} &\reducestov \sbst{v_2}{x}{v_1} \\
 \grun{\return{v}} &\reducestov v \\
+\grun{\letv{x_1}{c_1}{c_2}} &\reducestov \grun{\sbst{c_2}{x_1}{\grun{c_1}}} \\
+\indelimm{\Sigma}{(\ret{\Sigma}{v_s})}{\indretcase{x}{v_r} \seq \ldots} &\reducestov \sbst{v_r}{x}{v_s} \\
+\indelimm{\Sigma}{(\indintro{\Sigma}{op}{v_1}{v_2})}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2} \reducestov \\
+  \sbst{\sbst{v_3}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{&\letv{x_w}{\app{v_2}{x_a}}{\return{\indelimm{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2}}}}} \\
 \letv{x}{\return{v}}{c} &\reducestoc \sbst{c}{x}{v} \\
+\letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} &\reducestoc \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}}\\
 \letv{x_1}{\perform{v_1}{op}{v_2}{x_2}{c_1}}{c_2} &\reducestoc \perform{v_1}{op}{v_2}{x_2}{\letv{x_1}{c_1}{c_2}} \\
 \app{(\abs{x}{\tau}{c})}{v} &\reducestoc \sbst{c}{x}{v} \\
-\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_1}{x_1}{v} \\
-\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_2}{x_2}{v} \\
-\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} &\reducestoc \sbst{c}{x}{v} \\
-\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
-  \reducestoc \sbst{\sbst{c}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{&\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}} \\
 \reify{\Sigma}{x}{\return{v}} &\reducestoc \return{\ret{\Sigma}{v}} \\
-\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} &\reducestoc \return{\indintro{\Sigma}{op}{v}{\abs{x_2}{\reify{\Sigma}{x_1}{c}}}} \\
+\reify{\Sigma}{x_1}{\letv{x_2}{c_1}{c_2}} &\reducestoc \letret{\Sigma}{x_2}{\reify{\Sigma}{x_1}{c_1}}{\reify{\Sigma}{x_1}{c_2}} \\
+\reify{\Sigma}{x_1}{\perform{x_1}{op}{v}{x_2}{c}} &\reducestoc \return{\indintro{\Sigma}{op}{v}{\abs{x_2}{\tau}{\reify{\Sigma}{x_1}{c}}}} \\
 \reify{\Sigma}{x_1}{\perform{x_2}{op}{v}{x_3}{c}} &\reducestoc \perform{x_2}{op}{v}{x_3}{\reify{\Sigma}{x_1}{c}} \qquad (x_1 \neq x_2) \\
@@
 \begin{align*}
-  \mathrm{neutral\ values}, \eta :=\;& x \mid \fst \eta \mid \snd \eta
-                                     \mid \gunbox{x}{\eta}{\omega} \mid \grun{\mu} \\
-  \mathrm{neutral\ computations}, \mu :=\;& \letv{x}{\mu}{\gamma}
-                                           \mid \app{\eta}{\omega} \mid \\
-                                   & \case{\eta}{x}{\gamma}{x}{\gamma} \mid \\
-                            & \indelim{\Sigma}{\eta}
-                              {\indretcase{x}{\gamma} \seq
-                               \indcase{op}{x}{x}{\gamma} \seq \ldots \seq
-                               \indcase{op}{x}{x}{\gamma}} \\
-                                   & \reify{\Sigma}{x}{\mu} \\
+  \mathrm{neutral\ values}, \eta :=\;& x \mid \fst \eta \mid \snd \eta \mid \\
+                                   &\case{\eta}{x}{\omega}{x}{\omega} \mid \\
+                                   &\casebox{\eta}{x}{\omega}{x}{\omega} \mid \\
+                                   &\gunbox{x}{\eta}{\omega} \mid \grun{\mu} \mid \\
+                                   &\indelim{\Sigma}{\eta}
+                                     {\indretcase{x}{\omega} \seq
+                                     \indcase{op}{x}{x}{\omega} \seq \ldots \seq
+                                     \indcase{op}{x}{x}{\omega}} \\
+  \mathrm{neutral\ computations}, \mu :=\;& \app{\eta}{\omega}
+                                            \mid \reify{\Sigma}{x}{\mu} \\
   \mathrm{normal\ values}, \omega :=\;& \eta \mid (\omega, \omega)
@@
   \mathrm{normal\ computations}, \gamma :=\;& \mu \mid \return{\omega}
+                                              \mid \letv{x}{\mu}{\gamma}
                                                \mid \perform{\eta}{op}{\omega}{x}{\gamma}
@@
 \end{lemma}
-
-Note that the reduction relations do not include $\eta$ expansion. They
-also do not include some conversions you might expect for computations:
-\begin{itemize}
-\item There is no rule to convert
-  $\; \letv{x_2}{\letv{x_1}{c_1}{c_2}}{c_3} \,$ to
-  $\; \letv{x_1}{c_1}{\letv{x_2}{c_2}{c_3}} \,$ or vice versa.
-\item There is no rule to convert $\; \letv{x}{c}{\return x} \,$ to $c$.
-\end{itemize}
-These rules are excluded because they are not $\beta$ rules. Instead
-they are better thought of as $\nu$ rules~\cite{allais2013new}:
-additional equalities between neutral terms that are immediate
-consequences of structural induction. Rather than be included as part of
-ordinary reduction they can be done as an additional standardization
-pass on neutral terms after reduction and $\eta$-expansion have
-completed. This staged approach to computing equality of terms works
-because $\nu$ rules cannot produce additional $\beta$ redexes.
 
@@
 
+\todo{Prove that this embedding respects reduction of STLC}
+
 \section{The ambient monad model}
@@
 &\ind{\Sigma} \types [\pshlawcop, \pshlawcop] \\
-&\ind{\Sigma}(P) = \mu X. (P + F_\Sigma \circ \mon{X})
+&\ind{\Sigma}(P) = \mu X. ((\glob P) + F_\Sigma \circ \mon{X})
 \end{align*}
@@
   \begin{equation*}
-  T_{L + \Sigma}(P(L)) \cong \mon{(\ind{\Sigma}(P))}(L)
+  T_{L + \Sigma}(P(\initial)) \cong \mon{(\ind{\Sigma}(P))}(L)
   \end{equation*}
@@
   &\cong \mon{(\glob P)}(L^{\prime} + L) \\
-  &\cong T_{(L^{\prime} + L)}(P(L^{\prime}))
+  &\cong T_{(L^{\prime} + L)}(P(\initial))
 \end{align*}
@@
 \begin{equation*}
-  \arrow{\yoneda{\Sigma}}{\mon{(\glob P)}} \cong \mon{(\ind{\Sigma}(\glob P))}
+  \arrow{\yoneda{\Sigma}}{\mon{(\glob P)}} \cong \mon{(\ind{\Sigma}(P))}
 \end{equation*}
@@
 \begin{align*}
-\case{(\inl{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_1}{x_1}{v}
-\case{(\inr{v})}{x_1}{c_1}{x_2}{c_2} &\reducestoc \sbst{c_2}{x_2}{v}
+\case{(\inl{v_1})}{x_1}{v_2}{x_2}{v_3} &\reducestoc \sbst{v_2}{x_1}{v_1}
+\case{(\inr{v_1})}{x_1}{v_2}{x_2}{v_3} &\reducestoc \sbst{v_3}{x_2}{v_1}
 \end{align*}
@@
 \begin{equation*}
-\indelim{\Sigma}{(\ret{\Sigma}{v})}{\indretcase{x}{c} \seq \ldots} \reducestoc \sbst{c}{x}{v}
+\indelim{\Sigma}{(\ret{\Sigma}{v_1})}{\indretcase{x}{v_2} \seq \ldots} \reducestoc \sbst{v_2}{x}{v_1}
 \end{equation*}
@@
 \begin{align*}
-&\indelim{\Sigma}{\indintro{\Sigma}{op}{v_1}{v_2}}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2} \\
-&\qquad \reducestoc \sbst{\sbst{c}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{c} \seq H_2}}}}
+&\indelim{\Sigma}{(\indintro{\Sigma}{op}{v_1}{v_2})}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2} \\
+&\qquad \reducestoc \sbst{\sbst{v_3}{x_1}{v_1}}{x_2}{\abs{x_a}{\tau_a}{\letv{x_w}{\app{v_2}{x_a}}{\indelim{\Sigma}{x_w}{H_1 \seq \indcase{op}{x_1}{x_2}{v_3} \seq H_2}}}}
 \end{align*}
